---
title: "How a ggplot is Built"
subtitle: "A layer-by-layer walkthrough with the Challenger O-ring example"
format:
  html:
    toc: true
    code-fold: false
    code-summary: "Show code"
execute:
  echo: true
  warning: false
  message: false
---

# Purpose

This document is designed for teaching `ggplot2` by **building a plot one stage at a time**.

The goal is not just to produce a final figure, but to show students how a `ggplot` is assembled from distinct components:

-   **data**
-   **aesthetic mappings**
-   **geometric layers**
-   **statistical layers**
-   **annotations**
-   **labels**
-   **theme elements**

We will use the Challenger O-ring example because it is a classic case in data visualization: the relationship between temperature and damage becomes much easier to see once the data are plotted.

# Load packages and data

We will use the `orings` data from the `DAAG` package and add a damage index used in many retellings of this example.

```{r}
library(DAAG)
library(ggplot2)

data(orings)

Damage <- c(11, 4, 4, 2, 0, 0, 0, 0, 0, 0, 4, 0, 4, 0, 0, 0, 0, 4, 0, 0, 0, 0, 0)
oring <- cbind(orings, Damage)

oring
```

# Step 0: Look at the data

Before plotting, it helps to inspect the variables.

```{r}
head(oring)
```

The main variables we will use are:

-   `Temperature`: launch temperature in degrees Fahrenheit
-   `Damage`: O-ring damage index

# Step 1: Create the plotting canvas

A `ggplot` usually begins with the dataset and the variable mappings.

```{r}
p0 <- ggplot(data = oring, aes(x = Temperature, y = Damage))
p0
```

This produces an empty plotting area.

That is useful to show students because `ggplot()` by itself does **not** draw points or lines. It only sets up the plot structure.

# Step 2: Add points

Now we add a geometric layer.

```{r}
p1 <- p0 + geom_point()
p1
```

This is the first complete graph.

What changed?

-   `p0` defined the data and axes
-   `geom_point()` told `ggplot2` to represent each observation as a point

# Step 3: Adjust point appearance

We can set visual properties directly in the geom.

```{r}
p2 <- p0 + geom_point(size = 2)
p2
```

Because `size = 2` is outside `aes()`, it is a fixed styling choice rather than a mapping to a variable.

# Step 4: Show mapping versus setting

This is one of the most important ideas for beginners.

## Fixed styling

```{r}
p_fixed <- ggplot(oring, aes(x = Temperature, y = Damage)) +
  geom_point(color = "steelblue", size = 3)

p_fixed
```

Here every point gets the same color.

## Mapping to a variable

```{r}
p_mapped <- ggplot(oring, aes(x = Temperature, y = Damage, color = factor(Blowby))) +
  geom_point(size = 3)

p_mapped
```

Here color depends on the data.

A good rule:

-   inside `aes()` means **map to data**
-   outside `aes()` means **set a fixed value**

# Step 5: Build the core scatter plot explicitly

For the main walkthrough, we will return to the simpler plot and keep building.

```{r}
p3 <- ggplot(oring, aes(x = Temperature, y = Damage)) +
  geom_point(size = 2)

p3
```

# Step 6: Add a smoother

Now add a statistical layer.

```{r}
p4 <- p3 + stat_smooth(method = "loess", se = FALSE)
p4
```

This adds a smooth curve summarizing the overall relationship between temperature and damage.

This is a useful teaching moment:

-   `geom_point()` shows the raw observations
-   `stat_smooth()` adds a derived summary

# Step 7: Use jitter to reduce overlap

Some observations share similar values, so points can overlap.

```{r}
p5 <- ggplot(oring, aes(x = Temperature, y = jitter(Damage))) +
  geom_point(size = 2) +
  stat_smooth(method = "loess", se = FALSE)

p5
```

Jitter makes overlapping points easier to see, though it also slightly perturbs the exact positions.

# Step 8: Highlight the forecast range

The forecast launch temperature for Challenger was about 26 to 29 degrees Fahrenheit.

We can highlight that region with a rectangle.

```{r}
p6 <- ggplot(oring, aes(x = Temperature, y = Damage)) +
  geom_point(size = 2) +
  stat_smooth(method = "loess", se = FALSE) +
  annotate(
    "rect",
    xmin = 26, xmax = 29,
    ymin = 0, ymax = 12,
    alpha = 0.2,
    fill = "red"
  )

p6
```

This is an annotation rather than a data layer.

# Step 9: Add explanatory text

Annotations can also include text.

```{r}
p7 <- p6 + annotate(
  "text",
  x = 31,
  y = 3,
  label = "Forecast\n26 to 29°F",
  hjust = 0
)

p7
```

This makes the figure more explanatory and less dependent on narration.

# Step 10: Add labels

Now add titles and axis labels.

```{r}
p8 <- p7 + labs(
  title = "O-ring damage versus temperature",
  subtitle = "Lower temperatures were associated with more damage",
  x = "Temperature (°F)",
  y = "Damage index",
  caption = "Data: DAAG::orings with added damage index"
)

p8
```

This is a good stage to discuss how labels shape interpretation.

# Step 11: Adjust the x-axis range

To focus attention on the forecast range and the observed temperatures, we can control the axis limits.

```{r}
p9 <- p8 + xlim(25, 85)
p9
```

Now the visual framing better supports the point of the graph.

# Step 12: Apply a theme

Themes control non-data ink: the background, grid lines, text styling, and related elements.

```{r}
p10 <- p9 + theme_minimal(base_size = 13)
p10
```

At this stage we have a polished plot assembled step by step.

# Step 13: Show the whole build in one block

After walking through each stage explicitly, it is useful to show the full code in one place.

```{r}
ggplot(oring, aes(x = Temperature, y = Damage)) +
  geom_point(size = 2) +
  stat_smooth(method = "loess", se = FALSE) +
  annotate(
    "rect",
    xmin = 26, xmax = 29,
    ymin = 0, ymax = 12,
    alpha = 0.2,
    fill = "red"
  ) +
  annotate(
    "text",
    x = 31,
    y = 3,
    label = "Forecast\n26 to 29°F",
    hjust = 0
  ) +
  labs(
    title = "O-ring damage versus temperature",
    subtitle = "Lower temperatures were associated with more damage",
    x = "Temperature (°F)",
    y = "Damage index",
    caption = "Data: DAAG::orings with added damage index"
  ) +
  xlim(25, 85) +
  theme_minimal(base_size = 13)
```

# Summary: what are the components?

The final plot contains:

-   **Data**: `oring`
-   **Mappings**: `aes(x = Temperature, y = Damage)`
-   **Geom layer**: `geom_point()`
-   **Statistical layer**: `stat_smooth()`
-   **Annotations**: `annotate("rect", ...)` and `annotate("text", ...)`
-   **Labels**: `labs(...)`
-   **Scale control**: `xlim(25, 85)`
-   **Theme**: `theme_minimal()`

# A useful mental model

One helpful way to describe `ggplot2` is:

> Start with data and mappings, then add layers that display, summarize, and explain the data.

That suggests a sequence of questions students can ask:

1.  What data am I using?
2.  Which variables belong on the axes?
3.  What geom should represent the observations?
4.  Do I want a statistical summary?
5.  Do I need annotations?
6.  What labels will help the viewer?
7.  What theme or formatting best supports the message?

# Optional student exercise

Modify the final plot in one way:

-   change the theme
-   change point size
-   add a color mapping
-   remove the smoother
-   rewrite the title and subtitle
-   move the annotation text

Then explain whether the change improves the figure and why.
