---
title: "Data Inspection & Analysis Using LLMs"
subtitle: "BIO/CHEM 806 — Live-Coding Companion  |  Session 2"
author: "Gretchen LeBuhn"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: "Session contents"
    code-fold: false
    theme: flatly
    highlight-style: github
    css: |
      .guardian { border-left: 5px solid #167B62; background: #FAFFFE;
                  padding: 1em 1.2em; margin: 1.2em 0; border-radius: 4px; }
      .guardian-title { color: #167B62; font-weight: bold; font-size: 1.05em;
                        margin-bottom: 0.5em; }
      .prompt-block { background: #F5F5F5; font-family: monospace; font-size: 0.9em;
                      padding: 0.8em 1em; border-radius: 4px; white-space: pre-wrap; }
execute:
  echo: true
  eval: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
# Packages used in this session — run this chunk first
# install.packages(c("tidyverse", "MASS", "broom", "performance"))
library(tidyverse)
library(MASS)       # glm.nb() for negative binomial
library(broom)      # tidy model output
library(performance) # check_overdispersion()
```

::: {.callout-note icon="false"}
## 📌 How to use this document

This document has two purposes:

1.  **During the session:** follow along in your browser (ChatGPT) for
    Parts 1–4, then switch to RStudio for Part 5.
2.  **After the session:** paste the R code exported from ChatGPT into
    the chunks in Part 5 and run it to verify reproducibility.

Set `eval: true` on any chunk you want to execute. All chunks are
`eval: false` by default so nothing runs unexpectedly during the demo.
:::

------------------------------------------------------------------------

# Part 1 — How LLMs Handle Tabular Data {#framing}

> No code this section — conceptual framing. Follow the discussion in
> chat.

## The three modes of LLM data work

| Mode | What it does | How to verify |
|-----------------|--------------------------|------------------------------|
| **Interpretation** | Reads data as text, reasons in natural language | Compare to what you know about the data |
| **Code execution** | Writes & runs Python/R internally; result is real computation | Inspect the code; re-run in R |
| **Narrative generation** | Converts results to plain language | Check numbers against model output |

## Token limits — know your model

```{r}
#| label: token-limits-table
#| eval: false

# Not R code — reference table for discussion
# 
# Model             | Tokens    | Approx. CSV capacity
# -------------------------------------------------------
# GPT-3.5           | ~4,000    | ~200 rows × 10 columns
# GPT-4 (8k)        | ~8,000    | ~400 rows × 10 columns
# GPT-4-turbo       | ~128,000  | ~6,000–10,000 rows × 10–20 cols
#
# Rule: each cell ≈ 5–10 tokens
# Risk: exceeding the limit causes silent truncation — the model doesn't warn you
```

------------------------------------------------------------------------

# Part 2 — Dataset Inspection in ChatGPT {#inspection}

> Work in the browser for this section. Paste the prompts into ChatGPT
> after uploading `Annual_survey_NAPL_only.csv`.

**Dataset URL:**
`https://sbclter.msi.ucsb.edu/external/Reef/Data/trial_data/Annual_survey_NAPL_only.csv`

::: {.callout-tip icon="false"}
## 🎯 Instructor cue

Upload via paperclip — not a URL link. Start a **new** chat each time
you upload a new dataset.
:::

## 2.1 First look

::: prompt-block
Tell me about this dataset. What does it appear to contain, what types
of variables are present, and what might this data be used to study?
:::

**What to look for in the response:**

- Does it correctly identify the ecological context (reef fish survey,
  site codes, species codes)?
- Does it distinguish COUNT (abundance) from SIZE (body size)?
- What does it get wrong or oversimplify?

------------------------------------------------------------------------

## 2.2 Missing values

::: prompt-block
Which columns contain missing values, and how many are in each column?
Present this as a table. Then tell me whether the missingness pattern
suggests data not collected versus species not observed.
:::

**Biology check:** In fish survey data, a missing COUNT likely means the
species wasn't observed (a structural zero), while a missing SIZE means
the animal was counted but not measured. These have different
implications for analysis.

------------------------------------------------------------------------

## 2.3 Summary statistics

::: prompt-block
Provide summary statistics — mean, median, standard deviation, min, max
— for the COUNT and SIZE columns. Also list unique values for SITE and
YEAR.
:::

**Watch for:** COUNT is almost certainly right-skewed (mean \>\>
median). This informs model family selection in Part 4.

------------------------------------------------------------------------

## 2.4 Data transformation

::: prompt-block
Convert this dataset to wide format using SP_CODE as column headers and
COUNT as values. If a species appears more than once for the same site
and year combination, sum the counts. Show me the first 10 rows and tell
me the dimensions of the resulting table.
:::

**Verify:** Number of rows in wide format should equal number of unique
site × year combinations.

------------------------------------------------------------------------

# Part 3 — Data Guardian Prompts {#guardian}

> Study these prompts. During Parts 4–5 we will deploy them in sequence.

The following 10 prompts are designed to catch the most common silent
failures in LLM-assisted data analysis. Use them proactively — before
each operation, not after you suspect something went wrong.

------------------------------------------------------------------------

::::: guardian
::: guardian-title
Guardian Prompt 1 — Data Ingestion Verification
:::

::: prompt-block
After loading the data, confirm: (1) the number of rows and columns, (2)
the column names and data types, (3) the first 5 rows. Do not proceed
with any analysis until you have shown me this information.
:::

**Why:** The very first thing that can go wrong is that the model didn't
read what you think it read. Run this immediately after every upload.

**Watch for:** Row count doesn't match your file → stop and investigate
before continuing.
:::::

------------------------------------------------------------------------

::::: guardian
::: guardian-title
Guardian Prompt 2 — Row Accountability After Every Transformation
:::

::: prompt-block
After each data transformation step, print the row count and the count
of any NA or zero values in the key response variable. If any
transformation removes rows, stop and report which rows were removed and
why before continuing.
:::

**Why:** Filtering, joining, dropping NAs, and pivoting can all remove
rows silently. This creates an audit trail you can check.

**Watch for:** LLMs tend to pipeline multiple steps together. Push back
— ask for counts after *each individual step*, not just at the end.
:::::

------------------------------------------------------------------------

::::: guardian
::: guardian-title
Guardian Prompt 3 — Duplicate Detection Before Aggregation
:::

::: prompt-block
Before performing any aggregation or join, check whether there are
duplicate rows based on the key columns \[list them\]. Report the number
of duplicate groups and show me 3 example duplicate records. Do not
aggregate until I confirm the duplicates are expected.
:::

**Why:** Undetected duplicates inflate counts and means. A species
appearing twice in the same site-year inflates trend estimates without
any warning.

**Watch for:** Even confirming *no* duplicates is valuable — absence is
as important as presence.
:::::

------------------------------------------------------------------------

::::: guardian
::: guardian-title
Guardian Prompt 4 — Join Audit
:::

::: prompt-block
After performing the join, report: (1) the row count of each input
table, (2) the row count of the result, (3) any keys present in one
table but not the other. If the result has more rows than the larger
input, explain why.
:::

**Why:** Inner joins silently drop unmatched rows. Left joins can
inflate rows if the right table has duplicates. The model produces a
result either way.

**Watch for:** Result rows \> larger input rows → many-to-many
duplication. Investigate immediately.
:::::

------------------------------------------------------------------------

::::: guardian
::: guardian-title
Guardian Prompt 5 — Zero vs. NA — Force Explicit Distinction
:::

::: prompt-block
Before any analysis involving COUNT or abundance: distinguish between
cells where the value is 0 (species observed but none counted) and cells
where the value is NA (data not collected or species not surveyed).
Report the count of each. Do not treat zeros and NAs as equivalent in
any analysis.
:::

**Why:** In ecological count data, a zero and an NA have fundamentally
different biological meanings. Models that conflate them — or drop both
— produce wrong answers.

**Watch for:** Especially important for zero-inflated models and
occupancy analyses.
:::::

------------------------------------------------------------------------

::::: guardian
::: guardian-title
Guardian Prompt 6 — Outlier Flagging Before Modeling
:::

::: prompt-block
Before fitting any model, identify values in \[response variable\] that
are more than 3 standard deviations from the mean. List them with their
row identifiers and ask me whether to include, exclude, or investigate
them before proceeding.
:::

**Why:** Outliers can dominate regression results, especially with count
data. The model will fit whatever you ask for regardless of whether one
record is driving the trend.

**Watch for:** Do NOT automatically exclude outliers — in ecology,
extreme values are often real and ecologically meaningful. Surface them
for your judgment.
:::::

------------------------------------------------------------------------

::::: guardian
::: guardian-title
Guardian Prompt 7 — Scope Confirmation
:::

::: prompt-block
Before running any analysis, restate in plain language: (1) which rows
are included (what filters have been applied), (2) which columns are
used, and (3) what the analysis unit is (e.g., one row = one
species-site-year observation). Confirm this matches my research
question.
:::

**Why:** After multiple transformation steps, it's easy to lose track of
what subset the analysis is actually running on. This forces a
restatement before any model is fitted.

**Watch for:** If the model's restatement doesn't match your intent,
correct it *before* it runs anything. Far easier than diagnosing a wrong
answer afterward.
:::::

------------------------------------------------------------------------

::::: guardian
::: guardian-title
Guardian Prompt 8 — Model Assumption Check Before Interpretation
:::

::: prompt-block
After fitting the model, check and report the following before
interpreting results: (1) residual deviance and degrees of freedom, (2)
whether the dispersion parameter suggests overdispersion, (3) a residual
plot, and (4) whether any influential observations were detected (Cook's
distance or equivalent). Do not present conclusions until these checks
are done.
:::

**Why:** A Poisson GLM with overdispersion produces standard errors that
are too small and p-values that are too significant. The model will show
you a p-value regardless of whether assumptions are met.

**Watch for:** Residual deviance \>\> degrees of freedom →
overdispersion. Rule of thumb: deviance/df \> 2 → consider switching to
negative binomial.
:::::

------------------------------------------------------------------------

::::: guardian
::: guardian-title
Guardian Prompt 9 — Aggregation Transparency
:::

::: prompt-block
When you aggregate or summarize this data, tell me explicitly: (1) how
many original rows contributed to each summary value, (2) what function
was used (sum, mean, median), and (3) whether any rows were excluded
from the aggregation and why.
:::

**Why:** Every time series plot or summary table involves aggregation.
Making the model describe the aggregation operation lets you verify the
plot shows what you think it shows.

**Watch for:** Whether NAs were excluded or treated as zero before
aggregation — default behavior varies by language and function.
:::::

------------------------------------------------------------------------

::::: guardian
::: guardian-title
Guardian Prompt 10 — Reproducibility Handoff
:::

::: prompt-block
Please compile all the code used in this session into a single,
self-contained R script that I can run locally. The script should: load
the data from file, reproduce every transformation and visualization we
discussed, and include comments explaining each step. After generating
the script, list any steps that were performed in Python that will need
translation to R.
:::

**Why:** Analysis done inside ChatGPT is ephemeral and not reproducible.
This prompt converts the session into a permanent, auditable record. Run
it as the *last thing* you do in every ChatGPT data analysis session.

**Watch for:** Run the exported script in R and verify it produces the
same numerical results ChatGPT reported. Any discrepancy is a bug — not
a rounding difference to ignore.
:::::

------------------------------------------------------------------------

# Part 4 — Statistical Analysis in ChatGPT {#stats}

> Continue working in the browser. Use the Guardian Prompt numbers in
> brackets as reminders of which prompt you are deploying.

## 4.1 Scope confirmation before plotting \[GP 7\]

::: prompt-block
Before creating any plots: restate which rows are included in the
analysis, what the analysis unit is, and whether NA values in COUNT have
been handled. \[Guardian Prompt 7\]
:::

## 4.2 Time series — total count \[GP 9\]

::: prompt-block
Create a time series plot of total fish count across all species, with
YEAR on the x-axis and total COUNT on the y-axis. Use different colors
for each SITE. After creating the plot, tell me how many rows
contributed to each year's total and whether any years had missing data.
\[Guardian Prompt 9\]
:::

**What to notice:** Year-to-year variability vs. directional trend. Are
all sites represented in every year?

## 4.3 Top species composition \[GP 9\]

::: prompt-block
Create a time series plot of the top 5 most abundant species across all
years. Before plotting, report which 5 species were selected and their
total counts so I can verify the ranking. \[Guardian Prompt 9\]
:::

## 4.4 Model selection consultation \[GP 8 — pre-flight\]

::: prompt-block
I want to understand whether total fish count has increased or decreased
over time. Before recommending a model, tell me: (1) whether count data
with many zeros is appropriate for a Poisson GLM or whether another
distribution family is more suitable, and (2) what assumptions that
model makes that I should verify. \[Guardian Prompt 8\]
:::

**What to listen for:** Poisson → check equidispersion; many zeros →
zero-inflation risk; variance \> mean → overdispersion → negative
binomial.

## 4.5 Fitting and checking the model \[GP 8\]

::: prompt-block
Fit a Poisson GLM with total annual count as the response and YEAR as
the predictor. After fitting, report: (1) residual deviance and degrees
of freedom, (2) the dispersion ratio (deviance/df), (3) whether this
suggests we should switch to a negative binomial model. Do not interpret
results until after this check. \[Guardian Prompt 8\]
:::

**Dispersion rule of thumb:** deviance/df \> 2 → overdispersion → refit
with negative binomial.

## 4.6 Refit if needed

::: prompt-block
The dispersion ratio suggests overdispersion. Please refit using a
negative binomial GLM, then show me the coefficient for YEAR and its 95%
confidence interval. Interpret the coefficient in plain language — what
does it mean for how fish counts change per year?
:::

## 4.7 Report-ready narrative

::: prompt-block
Write a methods and results paragraph suitable for a scientific report
describing the GLM analysis. Include: the model family and link
function, the reason for choosing it, the key statistical results
(coefficient, confidence interval, p-value), and a plain-language
interpretation of the trend. Use past tense and avoid jargon where
possible.
:::

**Critical read:** Does the narrative accurately reflect the model you
actually fitted (Poisson or negative binomial)? Does it cite the right
coefficient and p-value? Narrative quality is not evidence that the
numbers are right.

------------------------------------------------------------------------

# Part 5 — Reproducibility: Export and Verify in R {#reproducibility}

> Switch to RStudio. Use Guardian Prompt 10 in ChatGPT first, then paste
> the exported code into the chunks below.

## 5.1 Send the reproducibility handoff prompt \[GP 10\]

In ChatGPT, send Guardian Prompt 10:

::: prompt-block
Please compile all the code used in this session into a single,
self-contained R script that I can run locally. The script should: load
the data from file, reproduce every transformation and visualization we
discussed, and include comments explaining each step. After generating
the script, list any steps that were performed in Python that will need
translation to R. \[Guardian Prompt 10\]
:::

## 5.2 Load and inspect the data in R

```{r}
#| label: load-data
#| eval: false

# Paste ChatGPT's data loading code here
# Then verify dimensions match what ChatGPT reported

fish <- read_csv("Annual_survey_NAPL_only.csv")

# Verification check — run this regardless of what ChatGPT exported
cat("Rows:", nrow(fish), "\n")
cat("Columns:", ncol(fish), "\n")
glimpse(fish)
```

## 5.3 Reproduce the transformations

```{r}
#| label: transformations
#| eval: false

# Paste ChatGPT's transformation code here
# After each step, run the verification pattern below:

# GUARDIAN PROMPT 2 — row accountability
# (run this after every transformation — substitute your object and column names)
verify_step <- function(df, label, response_col) {
  cat("\n---", label, "---\n")
  cat("  Rows:", nrow(df), "\n")
  cat("  NAs in", response_col, ":", sum(is.na(df[[response_col]])), "\n")
  cat("  Zeros in", response_col, ":", sum(df[[response_col]] == 0, na.rm = TRUE), "\n")
}

# Example usage — adapt column names to your data:
# verify_step(fish, "after loading", "COUNT")
# verify_step(fish_clean, "after dropping NAs", "COUNT")
# verify_step(fish_filtered, "after filtering to years 2000+", "COUNT")
```

## 5.4 Reproduce the time series plot

```{r}
#| label: time-series-plot
#| eval: false

# Paste ChatGPT's plotting code here (translated to ggplot2 if needed)
# Verify: does the plot match what ChatGPT showed in the browser?

# Reference ggplot structure for total count by year and site:
# fish_summary <- fish |>
#   group_by(YEAR, SITE) |>
#   summarise(total_count = sum(COUNT, na.rm = TRUE), .groups = "drop")
#
# ggplot(fish_summary, aes(x = YEAR, y = total_count, color = SITE)) +
#   geom_line() + geom_point() +
#   labs(title = "Total fish count by year and site",
#        x = "Year", y = "Total count") +
#   theme_minimal()
```

## 5.5 Reproduce and verify the GLM

```{r}
#| label: glm-poisson
#| eval: false

# Step 1: aggregate to annual totals (adjust as needed)
fish_annual <- fish |>
  group_by(YEAR) |>
  summarise(total_count = sum(COUNT, na.rm = TRUE))

# Step 2: fit Poisson GLM
m_poisson <- glm(total_count ~ YEAR, data = fish_annual, family = poisson(link = "log"))
summary(m_poisson)

# Step 3: Guardian Prompt 8 — dispersion check
cat("\nDispersion ratio (deviance / df):",
    round(m_poisson$deviance / m_poisson$df.residual, 2), "\n")
cat("Rule of thumb: > 2 suggests overdispersion → consider negative binomial\n")

# Optional: formal dispersion test
# performance::check_overdispersion(m_poisson)
```

```{r}
#| label: glm-negbin
#| eval: false

# If dispersion ratio > 2, refit with negative binomial
m_negbin <- MASS::glm.nb(total_count ~ YEAR, data = fish_annual)
summary(m_negbin)

# Tidy output for reporting
broom::tidy(m_negbin, conf.int = TRUE) |>
  filter(term == "YEAR") |>
  mutate(
    exp_estimate = exp(estimate),        # count multiplier per year
    exp_conf.low = exp(conf.low),
    exp_conf.high = exp(conf.high)
  )

# Compare to what ChatGPT reported:
# Coefficient for YEAR: ___
# 95% CI: [___, ___]
# p-value: ___
```

## 5.6 Verification checklist

Run this chunk and fill in the comparison table:

```{r}
#| label: verification-checklist
#| eval: false

# Fill in after running the code above
comparison <- tibble(
  Quantity             = c("Row count after loading",
                            "Row count after cleaning",
                            "YEAR coefficient (log scale)",
                            "YEAR p-value",
                            "Dispersion ratio"),
  ChatGPT_reported     = c(NA, NA, NA, NA, NA),   # fill from ChatGPT output
  R_result             = c(NA, NA, NA, NA, NA)    # fill from your R output above
)

print(comparison)

# Any discrepancy in coefficients or row counts → investigate before trusting either result
```

------------------------------------------------------------------------

# Quick Reference — Data Guardian Prompts {#reference}

Keep this section open during any LLM data analysis session.

| \#  | Name                        | When to use                            |
|-----|-----------------------------|----------------------------------------|
| 1   | Data Ingestion Verification | Immediately after every file upload    |
| 2   | Row Accountability          | After every transformation step        |
| 3   | Duplicate Detection         | Before any aggregation or join         |
| 4   | Join Audit                  | After every merge/join operation       |
| 5   | Zero vs. NA Distinction     | Before any count or abundance analysis |
| 6   | Outlier Flagging            | Before fitting any model               |
| 7   | Scope Confirmation          | Before running any analysis            |
| 8   | Model Assumption Check      | After fitting, before interpreting     |
| 9   | Aggregation Transparency    | After any summary or plot              |
| 10  | Reproducibility Handoff     | Last thing in every session            |

::: {.callout-important icon="false"}
## The core principle

An LLM will produce confident, well-formatted output regardless of
whether it is computing on the right data, the right subset, or using
the right model. The Data Guardian Prompts do not make the LLM smarter —
they force it to show its work so that *you* can catch the errors.
:::
