---
title: "Publishing to the Web, Part 2"
subtitle: "Week 11 Lecture 2 · BIO/CHEM 806 · 75 min"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Today's lecture"
    theme: cosmo
  
    code-tools: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

---

## Section 1 — Opening reflection · `0:00–0:12` {#opening}

::: {.callout-tip}
## Instructor cue · Slides 1–3 · 12 min

**Stay on the slides.** Do not open this document yet.

Post the chat prompt from Slide 2 immediately after your opening line.
Read responses aloud for 5–6 minutes — this time is worth it. Look for
the analytically interesting decisions: the 70 mm threshold, filtering
NAPL, dropping NAs, stacked vs. dodged bars, adding the MPA
designation with `case_when()`. Name those specifically on Slide 3.

**Monday cohort:** genuine debrief — they did the activity.  
**Thursday cohort:** they're hearing what's coming. Ask them to start
forming their own thoughts for their section.

After Slide 3, switch to this document and tell students: *"Open your
own `training_{USERNAME}` project in RStudio and follow along."*
:::

::: {.callout-note}
## 💬 Chat prompt · Slide 2

*What was one analytical decision you made during the lobster activity
that you weren't sure about — a threshold, a filter, a plot choice?
Take 60 seconds and type it in chat.*
:::

**What those decisions reveal — Slide 3**

The decisions students described — threshold at 70 mm, whether to drop
NAs, stacked vs. dodged bars — are not wrong or right. **They are
scientific choices that need to be documented.** Adding the MPA
`DESIGNATION` with `case_when()` is a good example: it changes what
questions you can ask. That is a modeling decision that belongs in your
methods section.

When you publish a report to GitHub Pages, those decisions become
visible. That is a feature, not a bug. Reproducible work means your
choices are traceable.

---

## Section 2 — YAML options · `0:12–0:22` {#yaml-options}

::: {.callout-tip}
## Instructor cue · Slides 4–5 · 10 min

Walk through the YAML block on Slide 4 and the table below it slowly.
Then move to Slide 5 (themes) — work through the themes table, then do
the live demo: change `theme: cosmo` to `theme: sandstone` in this
document's YAML header, render, then change it back. One word, whole
page changes.

After the demo, show your own published GitHub Pages site and point out
which YAML options you used. The connection "this YAML produced that
page" is the payoff.
:::

The YAML header of your `index.qmd` controls how your published page
looks and behaves. A few high-value options:

```yaml
---
title: "My Research"
author: "Your Name"
date: today
toc: true
toc-depth: 2
number-sections: true
theme: cosmo
---
```

| YAML key | What it does |
|---|---|
| `date: today` | Auto-inserts the render date — updates every time you render |
| `toc: true` | Adds a clickable table of contents |
| `toc-depth: 2` | Controls how many heading levels appear in the TOC |
| `number-sections: true` | Numbers every section heading automatically |
| `theme: cosmo` | Applies a Bootstrap theme — one word changes the entire page |

```{r}
#| label: show-date

# date: today in YAML evaluates to this:
Sys.Date()
```

Every time you render and push, the published date updates
automatically. This tells a reader when the analysis was last touched —
important for a living document like a thesis site.

### Themes · Slide 5

Quarto ships with the full Bootswatch theme library. Preview all of
them at [bootswatch.com](https://bootswatch.com). Popular options for
scientific documents:

| Theme | Character |
|---|---|
| `cosmo` | Clean, modern, readable |
| `flatly` | Minimal, professional |
| `journal` | Slightly more editorial |
| `lumen` | Light and airy |
| `sandstone` | Earthy, warm |

::: {.callout-tip}
## Instructor cue — live theme demo

Change `theme: cosmo` to `theme: sandstone` in **this document's YAML
header** and render. Then change it back. Point out: one word, entire
page changes. Students can browse [bootswatch.com](https://bootswatch.com)
on their own time to choose a theme for their thesis site.
:::

---

## Section 3 — `_quarto.yml` · `0:22–0:35` {#quarto-yml}

::: {.callout-tip}
## Instructor cue · Slides 6–7 · 13 min

Return to the slides briefly for Slide 6, then come back here and
**open the `_quarto.yml` file you prepared before class** alongside
this document in RStudio. Show its contents without going deep into
syntax. The concept: one file at the root, shared settings, navbar
appears on every page.

End this section by pointing to the course website:
*"This site — `nceas-learning-hub.github.io/sfsu_data_sci` — is built
exactly this way. Everything you see there was made with the tools you
now know."*

**Remind students:** they do **not** need `_quarto.yml` for this
week's activity. A single `index.qmd` with Markdown links is
completely sufficient. This is the step up for thesis projects with
multiple chapters.
:::

When you have multiple `.qmd` files that should all look the same —
same theme, same TOC settings, same navigation bar — don't repeat the
YAML in each file. Create `_quarto.yml` at the **root of your project**
and Quarto applies it to every document automatically.

```yaml
# _quarto.yml — place this file at the root of your project

project:
  type: website

website:
  title: "My Research Project"
  navbar:
    left:
      - href: index.qmd
        text: Home
      - href: analysis/data-cleaning.qmd
        text: Data Cleaning

format:
  html:
    theme: cosmo
    toc: true
```

### What `_quarto.yml` unlocks · Slide 7

With `_quarto.yml` and `type: website`, your repository becomes a
proper multi-page site:

- A **navbar** that appears on every page automatically
- **Shared theme, font, and layout** across all documents
- `quarto render` in the Terminal builds every page at once
- `quarto preview` lets you see the full site locally before pushing

::: {.callout-note}
## You don't need this for today's activity

A single `index.qmd` with Markdown links is enough for the current
assignment. `_quarto.yml` is for thesis projects with multiple
chapters and a consistent navigation structure across all pages.
:::

---

## Section 4 — Thesis structure and open science · `0:35–0:47` {#thesis}

::: {.callout-tip}
## Instructor cue · Slides 8–10 · 12 min

This is the most important conceptual section — do not rush it.

**Slide 8:** Use the file tree to make the thesis structure concrete.
Narrate each file's purpose as you walk down the list.

**Slide 9:** Work through the three audience types (advisor,
collaborator, future you) — ask students which one they find hardest
to write for.

**Slide 10 (open science):** Post the chat prompt and take 3–4
responses. Listen for: an abstract at the top, figure labels explaining
biological significance, a methods section describing thresholds and
rationale, a link to EDI. Surface 2–3 of these explicitly.

**For Thursday students** who haven't done the activity yet: say
explicitly — *"Keep this question in mind during your section. You'll
have something concrete to think about."*
:::

### Connecting this to your thesis · Slide 8

Here is one way to structure a thesis GitHub repository so it publishes
cleanly to GitHub Pages:

```
thesis-project/
├── _quarto.yml                 ← shared settings and navbar
├── index.qmd                   ← project overview and data description
├── analysis/
│   ├── 01-clean.qmd            ← data cleaning decisions, documented
│   ├── 02-explore.qmd          ← exploratory analysis
│   ├── 03-model.qmd            ← statistical modeling
│   └── 04-figures.qmd          ← publication-ready figures
├── data/                       ← raw data (or link to repository)
└── figs/                       ← saved figure outputs
```

Your committee sees a URL. Your advisor can open it without installing
R. Your collaborator at another institution can check your work. Six
months from now, you can pick this up and understand what you did —
because you documented it.

### Who is your audience? · Slide 9

Different audiences need different things from your site:

**Advisor:** Can they trace your logic from raw data to conclusion? Is
the decision log visible?

**Collaborator who doesn't use R:** Can they see the visualizations and
read your interpretations without running anything?

**Future you:** Can you pick this up in six months and understand what
you did and why?

**The site forces you to write for all three.**

### Open science in practice · Slide 10

Many journals now request or require: data availability statements,
analysis code shared openly, and reproducible workflows. A GitHub Pages
site with your Quarto analyses satisfies all three. Some journals link
directly to your GitHub repo or your published site in supplementary
materials.

You are not learning a tool. **You are building a practice.**

::: {.callout-note}
## 💬 Chat prompt · Slide 10

*For those who have done the activity: if your lobster report were
supplementary material for a journal paper, what's the one thing you'd
add to make it more useful for a reader who can't run your code?*
:::

---

## Section 5 — Troubleshooting · `0:47–0:55` {#troubleshooting}

::: {.callout-tip}
## Instructor cue · Slide 11 · 8 min

Work through the three issues below. After presenting them, ask in
Zoom: *"Is everyone's site live? Thumbs up if yes."* Flag any
non-responders — they may need help before Thursday.

Keep this section brisk. The goal is to send students into Thursday
with working sites, not to cover every possible failure mode.
:::

Three issues come up almost every time:

**1 — Site shows "404" or won't load**

- Is `index.html` committed **and** pushed? Not `index.qmd` — the
  **HTML file** is what GitHub Pages serves.
- Did you activate Pages in **Settings → Pages → Branch: main →
  Folder: root → Save**?
- After pushing, wait 2 minutes and hard-refresh
  (`Cmd+Shift+R` on Mac, `Ctrl+Shift+R` on Windows).
- Try the URL in a private/incognito window first — browsers cache
  aggressively.

**2 — Link goes to `.qmd` instead of `.html`**

```markdown
<!-- Wrong — links to the source file -->
[See the data page](data.qmd)

<!-- Right — links to the rendered file -->
[See the data page](data.html)
```

Render the file first, then commit the `.html`. The `.qmd` is your
source. The `.html` is what the browser serves.

**3 — Old content is still showing**

GitHub Pages caches aggressively. Try the URL in an incognito window.
If that shows updated content, clear your browser cache or wait a few
minutes.

::: {.callout-note}
## The order matters every time

```
Render  →  Stage (add)  →  Commit  →  Pull  →  Push
```

Both the `.qmd` **and** the `.html` must be committed and pushed. The
`.qmd` is version control for your source. The `.html` is what GitHub
Pages publishes.
:::

---

## Section 6 — Thursday setup and bridge to spatial data · `0:55–1:15` {#thursday-bridge}

::: {.callout-tip}
## Instructor cue · Slides 12–14 · 20 min

Give the Thursday checklist directly and explicitly — Slide 12. Then
move into the sf bridge on Slides 13–14. Keep the spatial section
conceptual: no live coding, just orientation. Students are not expected
to understand `sf` yet; they are being shown what it unlocks.

**If running behind:** Compress to 10 minutes — give the Thursday
checklist verbally and shorten the sf teaser to one paragraph. Section
4 (thesis and open science) is the high-value content; protect it
before cutting here.

**If running ahead:** Expand Section 4 — have students sketch their own
thesis project file structure: file names, section titles, intended
audience. Valuable even if rough.
:::

### For Thursday's activity cohort · Slide 12

Everything from Monday's lecture applies to you — the activity is
identical for both cohorts.

Before you arrive at Thursday's section, make sure:

- [ ] GitHub Pages is activated for `training_{USERNAME}`
- [ ] `index.html` is live at your URL
      (`{username}.github.io/{repo-name}/`)
- [ ] You have at least one Markdown link working in `index.qmd`

If you ran into setup problems, **come a few minutes early to the
activity or post in the course discussion board today**. Nicole has
seen all the common errors.

### What's next: spatial data with `sf` · Slides 13–14

Next week we move into geospatial analysis. The publishing workflow is
**identical** to what you learned today — you will render analyses that
produce maps and publish them to GitHub Pages. The `sf` unit builds
directly on the foundation you have now.

**The conceptual shift:**

Right now, when you have location data, you have latitude and longitude
columns — numbers. `sf` treats those numbers as **geometry**: points in
a coordinate reference system. Once you have `sf` objects, you can do
things that are impossible with plain columns: spatial joins, distance
calculations, buffering, overlaying with polygon data.

The coordinate reference system (CRS) is what makes "spatial" different
from "just numbers with labels." The `leaflet` maps from the
visualization unit are a preview of the spatial thinking we'll build
on next.

```{r}
#| label: sf-teaser
#| eval: false
#| echo: true

# sf teaser — Slide 14
# Read this, don't run it yet.

library(sf)

# A data frame becomes spatial
lobster_sf <- st_as_sf(lobster_data,
                       coords = c("longitude", "latitude"),
                       crs = 4326)  # WGS84 — the standard GPS CRS

# Now you can do spatial joins, buffering, distance calculations...
# ...and publish the result to GitHub Pages.
```

---

## Wrap-up · Slide 15

::: {.callout-tip}
## Instructor cue — closing

For the **Monday cohort:** their published lobster report is due.
Confirm the URL works before they leave.

For the **Thursday cohort:** remind them of the pre-activity checklist
above. Post a Canvas reminder after class.

**Post-class checklist:**

- [ ] Confirm Thursday activity logistics with Nicole — pairing, data
      setup
- [ ] Post Canvas reminder: Thursday cohort must have Pages live before
      section
- [ ] Flag students without working sites for office hours
- [ ] Review today's chat responses from the open science discussion —
      strong answers worth surfacing at Thursday's debrief
- [ ] Prep spatial data materials for next week (textbook s15)
:::

This week in one sentence: **you learned to close the loop** between
your analysis and the world.

Write in R → Render with Quarto → Publish with GitHub Pages →
Share a URL.

That loop is reproducible, open, and professional. Next week: the same
loop, with maps.

---

## Going further {#going-further}

- **Textbook s12** — Step-by-step GitHub Pages setup; use as a
  reference if students get stuck on the publish workflow
- **Textbook s13** — What a complete collaborative report includes;
  the lobster report maps directly to their thesis site structure
- **Bootswatch themes** — [bootswatch.com](https://bootswatch.com) —
  all available in Quarto with one YAML word
- **Quarto documentation:**
  [HTML theming](https://quarto.org/docs/output-formats/html-themes.html) ·
  [Websites with `_quarto.yml`](https://quarto.org/docs/websites/) ·
  [YAML reference](https://quarto.org/docs/reference/formats/html.html)
