---
title: "Week 14: Writing Functions in R"
subtitle: "Instructor Live-Coding Companion"
author: "BIO/CHEM 806"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    code-fold: false
    code-tools: true
    theme: cosmo
    embed-resources: true
editor:
  markdown:
    wrap: 72
execute:
  eval: true
  echo: true
  warning: true
  message: true
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
```

::: {.callout-note}
## About this document

This is the instructor live-coding companion for **Week 14: Writing
Functions in R**. It covers both Monday and Wednesday lectures in
sequence. Run each chunk interactively during class — do not knit the
whole document at once. The TOC on the left lets you jump between
sections during lecture.
:::

---

# LECTURE 1 — Monday {#lecture-1}

*DRY principle · Function anatomy · Vectorization · Lists · roxygen2*

---

## 1.1 The Copy-Paste Problem {#copy-paste}

Before we write any functions, let's look at the problem they solve.

```{r}
#| label: copy-paste-trap

airtemps <- c(212, 30.3, 78, 32)

# The copy-paste approach — fragile and verbose
celsius1 <- (airtemps[1] - 32) * 5/9
celsius2 <- (airtemps[2] - 32) * 5/9
celsius3 <- (airtemps[3] - 32) * 5/9

celsius1
celsius2
celsius3
```

::: {.callout-tip}
## Instructor note
Ask students: "What's the problem with this code? What happens if the
formula changes? What if airtemps has 200 values?"
:::

---

## 1.2 Function Anatomy {#anatomy}

Every R function has four parts: name, arguments, body, return value.

```{r}
#| label: anatomy-demo

# Generic structure — walk through each part
function_name <- function(argument1, argument2) {
  
  # body: computation goes here
  result <- argument1 + argument2
  
  return(result)
}

# Call it
function_name(3, 7)
```

---

## 1.3 🌡️ Temperature Conversion {#temperature}

*Ecological context: field sensors often record in Fahrenheit; analysis
and publication require Celsius.*

### Build `fahr_to_celsius()`

```{r}
#| label: fahr-to-celsius

fahr_to_celsius <- function(fahr) {
  celsius <- (fahr-32) * 5/9
  return (celsius)
}
  
```

### Test on a single value

```{r}
#| label: test-single

fahr_to_celsius(212)    # expect 100
fahr_to_celsius(32)     # expect 0
```

### Test on a vector — vectorization for free

```{r}
#| label: test-vector

airtemps <- c(212, 30.3, 78, 32)
fahr_to_celsius(airtemps)
```

### ✏️ Chat exercise: `celsius_to_fahr()`

*Post in chat, give students ~4 minutes, then live-code solution below.*

```{r}
#| label: celsius-to-fahr

celsius_to_fahr <- function(celsius) {
  fahr <- celsius * 9/5 + 32
  return(fahr)
}
```

### Verify round-trip correctness

```{r}
#| label: round-trip

celsius  <- fahr_to_celsius(airtemps)
result   <- celsius_to_fahr(celsius)

airtemps == result    # all TRUE?
```

---

## 1.4 🔋 Battery Chemistry: Coulombic Efficiency {#coulombic}

*Battery chemistry context: CE is calculated for every charge-discharge
cycle. A researcher running 200-cycle experiments repeats this formula
hundreds of times — an ideal DRY problem.*

### The biology/chemistry slide moment

> **Coulombic Efficiency (CE)** = discharge capacity ÷ charge capacity × 100
>
> Li-ion batteries typically achieve >99% CE, but the difference between
> 99.0% and 99.96% predicts whether a cell survives 200 or 500+ cycles.

### Build `calc_coulombic_efficiency()`

```{r}
#| label: calc-ce

calc_coulombic_efficiency <- function(discharge_cap_mAh, charge_cap_mAh) {
  ce <- (discharge_cap_mAh / charge_cap_mAh) *100
  return (ce)
}

```

### Test it

```{r}
#| label: test-ce

calc_coulombic_efficiency(2.98, 3.01)   # single cycle
```

### Apply across a full cycling dataset

```{r}
#| label: cycling-data

cycling_data <- data.frame(
  cycle         = 1:5,
  charge_mAh    = c(3.01, 3.00, 2.99, 2.98, 2.97),
  discharge_mAh = c(2.98, 2.97, 2.96, 2.95, 2.94)
)

cycling_data$CE <- calc_coulombic_efficiency(
  cycling_data$discharge_mAh,
  cycling_data$charge_mAh
)

cycling_data
```

---

## 1.5 🧬 Molecular Biology: GC Content & Primer Tm {#molbio}

*Molecular biology context: GC content determines primer melting
temperature. Every primer must be screened before synthesis. Designing
20 primers means 20 calculations — a function makes it instant.*

### The biology/chemistry slide moment

> **GC content** = % of bases that are G or C
>
> G–C pairs form three hydrogen bonds; A–T pairs form two. GC-rich
> primers bind more tightly. GC content directly determines melting
> temperature (Tm) via **Wallace's Rule**:
>
> `Tm = 2°C × (A + T) + 4°C × (G + C)`

### Build `calc_gc_content()`

```{r}
#| label: calc-gc

calc_gc_content <- function(sequence) {
  sequence <-toupper(sequence)
  n_total <- nchar(sequence) # total number of bases
  n_gc <- nchar(gsub("[^GC]", "", sequence))  #counts G and C
  return ((n_gc/ n_total) * 100)
}
```

### Test it

```{r}
#| label: test-gc

calc_gc_content("ATGCGCATTAGC")   # 50%
calc_gc_content("GCGCGCGC")       # 100%
calc_gc_content("aaaaattttt")       # 0%
```

### Build `calc_primer_tm()` — Wallace's Rule

```{r}
#| label: calc-tm

calc_primer_tm <- function(sequence) {
  sequence <-toupper(sequence)
  n_a <- nchar(gsub("[^A]", "", sequence))
  n_t <- nchar(gsub("[^T]", "", sequence))
  n_g <- nchar(gsub("[^G]", "", sequence))
  n_c <- nchar(gsub("[^C]", "", sequence))
  
  tm <- 2 * (n_a + n_t) + 4 * (n_g + n_c)
  
  return(tm)
}

```

### Test it

```{r}
#| label: test-tm

calc_primer_tm("ATGCGCATTAGC")
calc_primer_tm("GCGCGCGC")        # all GC — high Tm
calc_primer_tm("AAAATTTT")        # all AT — low Tm
```

---

## 1.6 Returning Multiple Outputs: Lists {#lists}

*What if we want both GC content and Tm from one function call?
Return a named list.*

### Build `calc_primer_stats()` — the function family payoff

```{r}
#| label: calc-primer-stats

calc_primer_stats <- function(sequence) {
  gc <- calc_gc_content(sequence)
  tm <- calc_primer_tm(sequence)
  
  return(list(
    sequence   = sequence,
    gc_percent = gc,
    tm_celsius = tm
  ))
}
```

### Test on one primer

```{r}
#| label: test-primer-stats

calc_primer_stats("ATGCGCATTAGC")
```

### Apply across many primers with `lapply()`

```{r}
#| label: many-primers

primers <- c(
  "ATGCGCATTAGC",
  "GCGCTTAAGCCT",
  "AATATTTAATGG"
)

results   <- lapply(primers, calc_primer_stats)
primer_df <- do.call(rbind, lapply(results, as.data.frame))
primer_df
```

### The function family

```{r}
#| label: function-family
#| eval: false

# Our calc_ family — notice how autocomplete helps
calc_gc_content()           # mol bio
calc_primer_tm()            # mol bio
calc_primer_stats()         # mol bio

calc_coulombic_efficiency() # battery chem
```

---

## 1.7 Documenting with roxygen2 {#roxygen2}

*Documentation is part of the function — not an afterthought.*

### Documented `calc_gc_content()`

```{r}
#| label: documented-gc

#' Calculate GC content of a DNA sequence
#'
#' @param sequence Character string of nucleotide bases (A, T, G, C).
#'   Case-insensitive.
#'
#' @return Numeric: percentage of bases that are G or C (0–100)
#' @export
#'
#' @examples
#' calc_gc_content("ATGCGCATTAGC")
#' calc_gc_content("AAAAAA")

calc_gc_content <- function(sequence) 
```

### ✏️ Chat exercise: document `calc_coulombic_efficiency()`

*Students write the roxygen2 block; model answer below.*

```{r}
#| label: documented-ce

#' Calculate coulombic efficiency for a battery charge-discharge cycle
#'
#' @param discharge_cap_mAh Numeric. Discharge capacity in
#'   milliamp-hours (mAh). Must be positive.
#' @param charge_cap_mAh Numeric. Charge capacity in milliamp-hours
#'   (mAh). Must be positive.
#'
#' @return Numeric. Coulombic efficiency as a percentage. Values near
#'   100 indicate efficient charge storage; values substantially below
#'   100 indicate capacity loss.
#' @export
#'
#' @examples
#' calc_coulombic_efficiency(2.98, 3.01)
#' calc_coulombic_efficiency(
#'   discharge_cap_mAh = 2.94,
#'   charge_cap_mAh    = 2.97
#' )

calc_coulombic_efficiency <- function(discharge_cap_mAh,
                                      charge_cap_mAh) {
  ce <- (discharge_cap_mAh / charge_cap_mAh) * 100
  return(ce)
}
```


in a function it would look like:

lebuhn/
├── DESCRIPTION
├── NAMESPACE
├── R/
│   └── calc_gc_content.R
└── man/






---

# LECTURE 2 — Wednesday {#lecture-2}

*Conditionals · stop() · warning() · Default arguments · ggplot wrapping*

::: {.callout-important}
## Before Wednesday's class
Paste the functions from Lecture 1 into your R session or source this
document. All of these are needed: `calc_gc_content()`,
`calc_primer_tm()`, `calc_primer_stats()`, `calc_coulombic_efficiency()`,
and `cycling_data`.
:::

---

## 2.1 Warm-Up: Functions with Conditionals {#conditionals}

*A function that branches based on input — the warm-up chat exercise.*

```{r}
#| label: classify-temp

classify_temp <- function(celsius) {
  if (celsius > 30) {
    return("hot")
  } else if (celsius < 10) {
    return("cold")
  } else {
    return("moderate")
  }
}
```

### ✏️ Chat warm-up: predict before running

```{r}
#| label: test-classify

classify_temp(35)    # predict before running
classify_temp(5)     # predict before running
classify_temp(20)    # predict before running
classify_temp(15)    # the chat exercise question
```

::: {.callout-tip}
## Instructor note
Run each one separately. Ask "what do you predict?" and wait for chat
before hitting Run. The key insight: `return()` exits immediately — the
function never evaluates the remaining branches.
:::

---

## 2.2 🔋 Battery Chemistry: Specific Capacity {#specific-capacity}

*Specific capacity (mAh/g) = how much charge stored per gram of active
electrode material. The primary metric for comparing materials — and
a function where bad inputs produce silently wrong results.*

### Stage 1 — naïve version: what goes wrong?

```{r}
#| label: naive-specific-cap

calc_specific_capacity <- function(capacity_mAh, mass_g) {
  return(capacity_mAh / mass_g)
}
```

```{r}
#| label: bad-inputs-naive

# What happens?
calc_specific_capacity(3.0, 0)           # Inf — wrong, no error
calc_specific_capacity(3.0, 0.015)       # correct: 200 mAh/g
```

```{r}
#| label: bad-type-naive
#| error: true

calc_specific_capacity(3.0, "graphite")  # error — but cryptic
```

::: {.callout-tip}
## Instructor note
The `Inf` case is the most dangerous: it passes silently and produces a
meaningless result that could make it into a paper. This motivates
`stop()`.
:::

---

### Stage 2 — add `stop()` for impossible inputs

```{r}
#| label: with-stop

calc_specific_capacity <- function(capacity_mAh, mass_g) {
  
  if (!is.numeric(capacity_mAh) | !is.numeric(mass_g)) {
    stop("Both capacity_mAh and mass_g must be numeric.")
  }
  
  if (mass_g <= 0) {
    stop("mass_g must be a positive value — check your units.")
  }
  
  return(capacity_mAh / mass_g)
}
```

```{r}
#| label: test-stop
#| error: true

calc_specific_capacity(3.0, "graphite")  # stop: not numeric
calc_specific_capacity(3.0, 0)           # stop: mass <= 0
calc_specific_capacity(3.0, 0.015)       # works: 200 mAh/g
```

---

### Stage 3 — add `warning()` for suspicious-but-computable inputs

```{r}
#| label: with-warning

calc_specific_capacity <- function(capacity_mAh, mass_g) {
  
  if (!is.numeric(capacity_mAh) | !is.numeric(mass_g)) {
    stop("Both capacity_mAh and mass_g must be numeric.")
  }
  
  if (mass_g <= 0) {
    stop("mass_g must be a positive value — check your units.")
  }
  
  if (capacity_mAh < 0) {
    warning("Negative capacity detected — verify instrument output.")
  }
  
  return(capacity_mAh / mass_g)
}
```

```{r}
#| label: test-warning

calc_specific_capacity(-0.5, 0.015)   # warning, but returns -33.3
calc_specific_capacity(3.0, 0.015)    # clean: 200 mAh/g
```

::: {.callout-tip}
## Instructor note — the stop() vs warning() decision
Ask in chat: "Why `warning()` for negative capacity but `stop()` for
zero mass?"

- Zero mass: division by zero → `Inf` → meaningless result → `stop()`
- Negative capacity: odd but computable — might be real instrument noise
  worth investigating → `warning()`, let the user decide
:::

---

## 2.3 Default Arguments {#defaults}

*Arguments with defaults make functions flexible without burdening the
common case.*

```{r}
#| label: with-defaults

calc_specific_capacity <- function(capacity_mAh,
                                   mass_g,
                                   units = "mAh/g") {
  
  if (!is.numeric(capacity_mAh) | !is.numeric(mass_g)) {
    stop("Both capacity_mAh and mass_g must be numeric.")
  }
  if (mass_g <= 0) {
    stop("mass_g must be a positive value — check your units.")
  }
  if (capacity_mAh < 0) {
    warning("Negative capacity detected — verify instrument output.")
  }
  
  result <- capacity_mAh / mass_g
  message("Returning specific capacity in ", units)
  return(result)
}
```

```{r}
#| label: test-defaults

# Uses default units
calc_specific_capacity(3.0, 0.015)

# Override default
calc_specific_capacity(3.0, 0.015, units = "Ah/g")
```

---

## 2.4 🧬 Molecular Biology: Primer Quality Check {#primer-quality}

*Putting it all together: conditionals + stop() + warning() + defaults,
applied to a real primer screening workflow.*

### The biology/chemistry slide moment

> Primers for standard PCR should have GC content of **40–60%**. Outside
> this range: GC < 40% → may not bind reliably; GC > 60% → may form
> secondary structures. These thresholds are checked for every primer
> before ordering synthesis.

### Build `calc_primer_quality()`

```{r}
#| label: calc-primer-quality

calc_primer_quality <- function(sequence,
                                gc_min = 40,
                                gc_max = 60) {
  
  # Input validation
  if (!is.character(sequence)) {
    stop("sequence must be a character string.")
  }
  
  # Compute stats using our Lecture 1 functions
  gc <- calc_gc_content(sequence)
  tm <- calc_primer_tm(sequence)
  
  # Warn if GC is outside the specified range
  if (gc < gc_min | gc > gc_max) {
    warning("GC content (", round(gc, 1), "%) is outside the ",
            gc_min, "-", gc_max, "% optimal range.")
  }
  
  return(list(
    gc_percent  = gc,
    tm_celsius  = tm,
    gc_in_range = (gc >= gc_min & gc <= gc_max)
  ))
}
```

### Test it

```{r}
#| label: test-primer-quality

# Standard PCR — uses defaults
calc_primer_quality("ATGCGCATTAGC")      # GC = 50%, in range

# Low GC primer — should warn
calc_primer_quality("AATATTTAATGG")      # GC = 8%, warning

# High GC primer — should warn
calc_primer_quality("GCGCGGCATGCC")      # GC = 83%, warning
```

### Default arguments enable scientific flexibility

```{r}
#| label: flexible-thresholds

# GC-rich organism (e.g. Mycobacterium tuberculosis genome ~65% GC)
calc_primer_quality("GCGCGGCATGCC",
                    gc_min = 50,
                    gc_max = 75)

# High-stringency assay — tighter window
calc_primer_quality("ATGCGCATTAGC",
                    gc_min = 45,
                    gc_max = 55)
```

::: {.callout-tip}
## Instructor note
Point out that `calc_primer_quality()` calls `calc_gc_content()` and
`calc_primer_tm()` — functions we already wrote. Functions calling
functions is normal and encouraged. This is why we gave them sensible
names with a common prefix.
:::

---

## 2.5 Wrapping ggplot in a Function {#ggplot-function}

*The DRY principle applies equally to visualization. If you make the
same style of plot repeatedly, wrap it.*

### Step 1 — build a custom theme function

```{r}
#| label: custom-theme

custom_theme <- function(base_size = 9) {
  ggplot2::theme(
    text             = ggplot2::element_text(family = "Helvetica",
                                             color  = "gray30",
                                             size   = base_size),
    plot.title       = ggplot2::element_text(size  = ggplot2::rel(1.25),
                                             hjust = 0.5,
                                             face  = "bold"),
    panel.background = ggplot2::element_blank(),
    panel.border     = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(colour    = "grey90",
                                             linewidth = 0.25),
    legend.position  = "right",
    legend.key       = ggplot2::element_rect(colour = NA, fill = NA),
    axis.ticks       = ggplot2::element_blank(),
    axis.line        = ggplot2::element_blank()
  )
}
```

### Step 2 — wrap the full cycling plot

```{r}
#| label: plot-cycling

plot_cycling <- function(df,
                         point_size    = 2,
                         font_size     = 10,
                         ce_threshold  = 99) {
  
  ggplot(df, aes(x = cycle, y = CE)) +
    geom_point(size = point_size, color = "#167B62") +
    geom_line(color = "#167B62", alpha = 0.6) +
    geom_hline(yintercept = ce_threshold,
               linetype   = "dashed",
               color      = "#A23923") +
    labs(
      title = "Coulombic Efficiency by Cycle",
      x     = "Cycle Number",
      y     = "Coulombic Efficiency (%)"
    ) +
    custom_theme(font_size)
}
```

### Call it

```{r}
#| label: call-plot-cycling

# Default arguments
plot_cycling(cycling_data)
```

```{r}
#| label: call-plot-cycling-custom

# Override point size and threshold line
plot_cycling(cycling_data, point_size = 3, ce_threshold = 98)
```

::: {.callout-tip}
## Instructor note — the thesis argument
"Your advisor asks you to change the CE threshold line from 99% to 98%
across all 15 cycling plots in your thesis.

- **Without the function:** find and edit 15 ggplot blocks.
- **With the function:** change one default argument. Done."
:::

---

## 2.6 Putting It All Together {#synthesis}

*A complete pipeline function: validate input → compute → plot.*

```{r}
#| label: full-pipeline

analyze_cycling_data <- function(df,
                                 ce_threshold = 99,
                                 font_size    = 10) {
  
  # Validate required columns
  required_cols <- c("cycle", "charge_mAh", "discharge_mAh")
  if (!all(required_cols %in% names(df))) {
    stop("df must contain columns: cycle, charge_mAh, discharge_mAh")
  }
  
  # Compute CE
  df$CE <- calc_coulombic_efficiency(
    df$discharge_mAh,
    df$charge_mAh
  )
  
  # Warn if any CE exceeds 100%
  if (any(df$CE > 100)) {
    warning("Some CE values exceed 100% — check charge/discharge convention.")
  }
  
  # Plot
  plot_cycling(df,
               ce_threshold = ce_threshold,
               font_size    = font_size)
}
```

```{r}
#| label: test-pipeline

analyze_cycling_data(cycling_data)
```

```{r}
#| label: test-pipeline-bad
#| error: true

# What happens with a bad data frame?
bad_df <- data.frame(x = 1:3, y = 4:6)
analyze_cycling_data(bad_df)
```

---

## 2.7 Bridge to Activity {#bridge}

The Thursday in-person activity (s18) uses `find_max()` — here's how
today's tools map onto those exercises:

```{r}
#| label: bridge-map
#| eval: false

# Activity exercise 1-2: function anatomy (Monday)
double_it   <- function(x)   { return(2 * x) }
exclaim_age <- function(age) { return(paste("I am", age, "years old!")) }

# Activity exercise 3: conditionals (Wednesday)
find_max <- function(value_1, value_2) {
  if (value_1 > value_2) {
    return(value_1)
  } else if (value_2 > value_1) {
    return(value_2)
  }
}

# Activity exercise 5-6: stop() — same pattern as calc_specific_capacity()
find_max <- function(value_1, value_2) {
  if (!is.numeric(value_1) | !is.numeric(value_2)) {
    stop("Values must be numeric.")
  }
  if (value_1 == value_2) {
    stop("Values must be different from each other.")
  }
  if (value_1 > value_2) return(value_1)
  if (value_2 > value_1) return(value_2)
}

# Activity exercise 7: default arguments — same pattern as gc_min/gc_max
find_max <- function(value_1, value_2, equal_ok = FALSE) {
  if (!is.numeric(value_1) | !is.numeric(value_2)) {
    stop("Values must be numeric.")
  }
  if (value_1 == value_2) {
    if (equal_ok) return(value_1)
    stop("Values must be different from each other.")
  }
  if (value_1 > value_2) return(value_1)
  if (value_2 > value_1) return(value_2)
}
```

::: {.callout-note}
## For students
Everything in Thursday's activity follows the exact same patterns from
this week's lectures. Keep this companion document open as a reference.
The textbook page s18 has the full exercise set.
:::

---

## Summary: All Functions Written This Week {#summary}

```{r}
#| label: full-function-library
#| eval: false

# ── Lecture 1 ──────────────────────────────────────────────────────────

# Temperature (ecology)
fahr_to_celsius <- function(fahr) { ... }
celsius_to_fahr <- function(celsius) { ... }

# Battery chemistry
calc_coulombic_efficiency <- function(discharge_cap_mAh,
                                      charge_cap_mAh) { ... }

# Molecular biology — a function family
calc_gc_content   <- function(sequence) { ... }
calc_primer_tm    <- function(sequence) { ... }
calc_primer_stats <- function(sequence) { ... }   # returns list

# ── Lecture 2 ──────────────────────────────────────────────────────────

# Battery chemistry — with validation
calc_specific_capacity <- function(capacity_mAh,
                                   mass_g,
                                   units = "mAh/g") { ... }

# Molecular biology — with validation + defaults
calc_primer_quality <- function(sequence,
                                gc_min = 40,
                                gc_max = 60) { ... }

# Visualization
custom_theme  <- function(base_size = 9) { ... }
plot_cycling  <- function(df,
                          point_size   = 2,
                          font_size    = 10,
                          ce_threshold = 99) { ... }

# Pipeline
analyze_cycling_data <- function(df,
                                 ce_threshold = 99,
                                 font_size    = 10) { ... }
```
