---
title: "Week 8: Data Visualization"
subtitle: "Instructor Live-Coding Companion"
author: "BIO/CHEM 806 · Gretchen LeBuhn"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Jump to"
    theme: cosmo
    code-tools: true
    embed-resources: true
    df-print: paged
execute:
  echo: true
  eval: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

::: {.callout-note appearance="minimal"}
**How to use this document**

This is your live-coding script for Week 8. Run chunks in order —
objects created in one section are used in later sections. Open it in
RStudio alongside your slides. The table of contents on the left mirrors
the lecture structure.

-   **Lecture 1** covers Tufte, the Grammar of Graphics, data prep, and
    geom types
-   **Lecture 2** covers customization, themes, color, faceting, and
    interactive outputs

Render to HTML at any time to see the full annotated output.
:::

------------------------------------------------------------------------

# Lecture 1 {.unnumbered}

## Setup: Load Packages

```{r}
#| label: load-packages

library(readr)
library(dplyr)
library(tidyr)
library(forcats)   # working with categorical variables / factors
library(ggplot2)   # visualization
library(leaflet)   # interactive maps
library(DT)        # interactive tables
library(scales)    # scale functions for visualization
library(janitor)   # clean column names
library(viridis)   # colorblind-friendly color palettes
library(here)      # file paths relative to project root
```

------------------------------------------------------------------------

## Section 1: Edward Tufte

::: {.callout-tip appearance="minimal"}
**Slide cue: Slides 3–6 (Tufte section)**

No code to run here — this section is entirely conceptual. Use the slide
time to discuss Tufte's three key ideas before moving on.

Tufte → `ggplot2` connections to mention:

-   Data-Ink Ratio → `theme_minimal()`, `theme_classic()`
-   Chartjunk → avoid 3D charts, rainbow fills, heavy gridlines
-   Small Multiples → `facet_wrap()`
:::

------------------------------------------------------------------------

## Section 2: The Dataset

### Load and Explore

::: {.callout-tip appearance="minimal"}
**Slide cue: Slides 12–13 (Dataset + Explore the Data)**

Run these chunks live. Ask students to run along in their own session.
Read the output with them — the glimpse() output is particularly useful
for spotting the tidy data problem.
:::

```{r}
#| label: load-data

# delta_visits_raw <- read_csv(
#   "https://portal.edirepository.org/nis/dataviewer?packageid=edi.587.1&entityid=d0dd56e44e3ab9f5ec60f787b46f5bf0",
#   show_col_types = FALSE
# )



delta_visits_raw <- read_csv(
   here("data/r_data_visualization", "Socioecological_monitoring_data.csv")
 )
```

```{r}
#| label: explore-data

# What are the columns?
colnames(delta_visits_raw)
```

```{r}
# Shape and data types — note that visitor counts are in separate columns
glimpse(delta_visits_raw)
```

```{r}
# What time period does this data cover?
range(delta_visits_raw$Date)
```

```{r}
# What times of day were surveys conducted?
unique(delta_visits_raw$Time_of_Day)
```

### Clean Column Names

::: {.callout-tip appearance="minimal"}
**Slide cue: "Step 1: Explore Before You Plot"**

Pause after running clean_names() and show the before/after column name
comparison. Students appreciate how much cleaner snake_case is.
:::

```{r}
#| label: clean-names

delta_visits <- delta_visits_raw %>%
  janitor::clean_names()

# Compare before and after
colnames(delta_visits)
```

------------------------------------------------------------------------

## Section 3: Tidy Data for Plotting

::: {.callout-tip appearance="minimal"}
**Slide cue: Slides 14–16 (Is This Tidy? → pivot_longer)**

Before running pivot_longer(), pause and ask the chat question: "Is this
data tidy? Which columns make you think it might not be?"

Key insight to drive home: ggplot2 needs visitor_type to be a *variable*
(column) so it can be mapped to color. While it's spread across sm_boat,
med_boat, lrg_boat etc., that mapping is impossible.
:::

### Pivot to Long Format

```{r}
#| label: pivot-longer

visits_long <- delta_visits %>%
  pivot_longer(
    cols      = c(sm_boat, med_boat, lrg_boat,
                  bank_angler, scientist, cars),
    names_to  = "visitor_type",
    values_to = "quantity"
  ) %>%
  rename(restore_loc = eco_restore_approximate_location) %>%
  select(-notes)

# How did the shape change?
cat("Original: ", nrow(delta_visits), "rows ×", ncol(delta_visits), "cols\n")
cat("Long:     ", nrow(visits_long),  "rows ×", ncol(visits_long),  "cols\n")

head(visits_long)
```

### Exercise 1: Calculate Daily Visit Totals

::: {.callout-tip appearance="minimal"}
**Slide cue: Slide 17 (Exercise 1)**

Give students 2 minutes to attempt this before revealing the answer. The
result — daily_visits_loc — is the data frame used in every plotting
example for the rest of the week.
:::

```{r}
#| label: exercise-1-daily-totals

daily_visits_loc <- visits_long %>%
  group_by(restore_loc, date, visitor_type) %>%
  summarise(daily_visits = sum(quantity), .groups = "drop")

head(daily_visits_loc)
```

------------------------------------------------------------------------

## Section 4: The Four Essential Components

::: {.callout-tip appearance="minimal"}
**Slide cue: Slides 18–19 (Four Components + First Plot)**

Show all three options on the slide, then run Option 1 live. The
overlapping x-axis labels are intentional — let students see the
"before" state.
:::

### Three Ways to Write the Same Plot

```{r}
#| label: three-options

# Option 1 — data and mapping in ggplot() [PREFERRED]
ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits)) +
  geom_col()
```

```{r}
# Option 2 — mapping called in geom
ggplot(data = daily_visits_loc) +
  geom_col(aes(x = restore_loc, y = daily_visits))
```

```{r}
# Option 3 — everything in geom
ggplot() +
  geom_col(data = daily_visits_loc,
           aes(x = restore_loc, y = daily_visits))
```

::: {.callout-note appearance="minimal"}
All three produce the same plot. Option 1 is the convention — put shared
data and aesthetics in `ggplot()`, put geometry-specific arguments in
the `geom_*()` layer.
:::

------------------------------------------------------------------------

## Section 5: Geom Types

::: {.callout-tip appearance="minimal"}
**Slide cue: Slides 21–23 (Geom Types)**

Run each chunk and read the output with students. Note that the violin
and boxplot use the same filtered data — same data, different geometry,
different insight.
:::

### Boxplot — Distributions Across Categories

```{r}
#| label: geom-boxplot

daily_visits_loc %>%
  filter(
    daily_visits < 30,
    visitor_type %in% c("sm_boat", "med_boat", "lrg_boat")
  ) %>%
  ggplot(aes(x = visitor_type, y = daily_visits)) +
  geom_boxplot()
```

### Violin Plot — Shape of the Distribution

```{r}
#| label: geom-violin

daily_visits_loc %>%
  filter(
    daily_visits < 30,
    visitor_type %in% c("sm_boat", "med_boat", "lrg_boat")
  ) %>%
  ggplot(aes(x = visitor_type, y = daily_visits)) +
  geom_violin()
```

### Line + Point — Time Series (Two Geoms, One Plot)

```{r}
#| label: geom-line-point

daily_visits_loc %>%
  filter(
    restore_loc == "Decker Island",
    visitor_type == "med_boat"
  ) %>%
  ggplot(aes(x = date, y = daily_visits)) +
  geom_line() +
  geom_point()
```

::: {.callout-note appearance="minimal"}
`geom_line()` shows the trend; `geom_point()` marks where the actual
data points are. Combining them is better than either alone for time
series.
:::

------------------------------------------------------------------------

------------------------------------------------------------------------

# Lecture 2 {.unnumbered}

------------------------------------------------------------------------

## Section 1: The aes() Boundary

::: {.callout-tip appearance="minimal"}
**Slide cue: Slides 3–6 (aes() section)**

Post the chat question BEFORE running Version A: "What's the difference
between Version A and Version B?"

Run Version A first. Let students see the salmon bars. Then explain why,
then run the fix.

This is the most important teaching moment of Lecture 2.
:::

### Version A — The "Wrong" Way (The Salmon Experiment)

```{r}
#| label: aes-wrong-fill

# fill = "blue" INSIDE aes() — this does NOT make blue bars
ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits,
           fill = "blue")) +
  geom_col()
```

::: {.callout-important appearance="minimal"}
**Why salmon?** Putting `fill = "blue"` inside `aes()` tells ggplot2
that `"blue"` is a variable in your data. It creates a new column filled
with the word "blue" and maps it to fill using the default color palette
— hence salmon, not blue.
:::

### The Fix — Constant Outside aes()

```{r}
#| label: aes-correct-constant

# fill = "blue" OUTSIDE aes() — now all bars are actually blue
ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits)) +
  geom_col(fill = "blue")
```

### The Right Way — Map a Variable to Color

```{r}
#| label: aes-correct-variable

# fill = visitor_type INSIDE aes() — correct because it IS a variable
ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits,
           fill = visitor_type)) +
  geom_col()
```

::: {.callout-note appearance="minimal"}
**The rule:**

-   **Inside `aes()`** → maps a *variable* to an aesthetic (value
    changes per data point based on the data)
-   **Outside `aes()`** → sets a *constant* for all data points
:::

------------------------------------------------------------------------

## Section 2: Labels and Built-in Themes

::: {.callout-tip appearance="minimal"}
**Slide cue: Slides 7–8 (labs + themes)**

Note the axis flip — x and y are swapped from the earlier version. This
solves the overlapping label problem in one line.

After showing this plot, open the RStudio autocomplete and type theme\_
then Tab to show the full list of built-in themes live.
:::

### labs() + Flipped Axes + theme_bw()

```{r}
#| label: labs-and-theme

ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits,
           fill = visitor_type)) +
  geom_col() +
  labs(
    x        = "Number of Visits",
    y        = "Restoration Location",
    fill     = "Type of Visitor",
    title    = "Total Visits to Delta Restoration Areas by Visitor Type",
    subtitle = "Sum of all visits during July 2017 and March 2018"
  ) +
  theme_bw()
```

### Built-in Theme Comparison

```{r}
#| label: themes-demo

# Build the base plot once so we can quickly swap themes
base_plot <- ggplot(data = daily_visits_loc,
                    aes(y = restore_loc, x = daily_visits,
                        fill = visitor_type)) +
  geom_col() +
  labs(x = "Number of Visits", y = "Restoration Location",
       fill = "Type of Visitor",
       title = "Total Visits to Delta Restoration Areas")

base_plot + theme_bw()       # white background, dark grid
base_plot + theme_minimal()  # no background panel
base_plot + theme_light()    # light gray grid
base_plot + theme_classic()  # no grid — publication classic
```

------------------------------------------------------------------------

## Section 3: Custom theme() and Reusable Theme Objects

::: {.callout-tip appearance="minimal"}
**Slide cue: Slides 9–10 (theme() + my_theme)**

Emphasize the ORDER WARNING: theme() must come AFTER theme_bw() or the
built-in theme will override your custom settings.

After building my_theme, show students how it applies to multiple plots
with a single + my_theme call.
:::

### Fine-Tuning with theme()

```{r}
#| label: custom-theme

ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits,
           fill = visitor_type)) +
  geom_col() +
  labs(
    x     = "Number of Visits",
    y     = "Restoration Location",
    fill  = "Type of Visitor",
    title = "Total Visits to Delta Restoration Areas by Visitor Type"
  ) +
  theme_bw() +                         # built-in theme FIRST
  theme(
    legend.position = "bottom",        # move legend
    axis.ticks.y    = element_blank()  # remove y-axis tick marks
  )
```

### Save Theme as a Reusable Object

```{r}
#| label: my-theme

my_theme <- theme_bw(base_size = 16) +
  theme(
    legend.position = "bottom",
    axis.ticks.y    = element_blank()
  )
```

```{r}
#| label: plot-with-my-theme

ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits,
           fill = visitor_type)) +
  geom_col() +
  labs(
    x        = "Number of Visits",
    y        = "Restoration Location",
    fill     = "Type of Visitor",
    title    = "Total Visits to Delta Restoration Areas by Visitor Type",
    subtitle = "Sum of all visits during study period"
  ) +
  my_theme
```

------------------------------------------------------------------------

## Section 4: Exercise 2 — scale_x_continuous()

::: {.callout-tip appearance="minimal"}
**Slide cue: Slide 11 (Exercise 2)**

Ask the chat prediction question BEFORE running. "What do you think
scale_x_continuous(breaks = seq(0, 120, 20)) will do?"

Then run once without expand, then once with expand = c(0,0) so students
can see exactly what each argument changes.
:::

```{r}
#| label: exercise-2-scale

# Step 1: Add axis breaks
ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits,
           fill = visitor_type)) +
  geom_col() +
  labs(x = "Number of Visits", y = "Restoration Location",
       fill = "Type of Visitor",
       title = "Total Visits to Delta Restoration Areas by Visitor Type",
       subtitle = "Sum of all visits during study period") +
  scale_x_continuous(breaks = seq(0, 120, 20)) +
  my_theme
```

```{r}
# Step 2: Also remove the gap between bars and axis (expand)
ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits,
           fill = visitor_type)) +
  geom_col() +
  labs(x = "Number of Visits", y = "Restoration Location",
       fill = "Type of Visitor",
       title = "Total Visits to Delta Restoration Areas by Visitor Type",
       subtitle = "Sum of all visits during study period") +
  scale_x_continuous(breaks = seq(0, 120, 20), expand = c(0, 0)) +
  my_theme
```

------------------------------------------------------------------------

## Section 5: Reordering with fct_reorder()

::: {.callout-tip appearance="minimal"}
**Slide cue: Slides 12–13 (fct_reorder)**

After creating daily_visits_totals, run levels() and show students that
the factor levels are now ordered by total visits, not alphabetically.
This is the key to understanding what fct_reorder actually does — it
doesn't sort the data frame rows; it reorders the factor levels that
ggplot2 uses for axis ordering.
:::

### Build Totals and Reorder

```{r}
#| label: fct-reorder-ascending

daily_visits_totals <- daily_visits_loc %>%
  group_by(restore_loc) %>%
  mutate(n = sum(daily_visits)) %>%
  ungroup() %>%
  mutate(restore_loc = fct_reorder(restore_loc, n))

# Confirm: no longer alphabetical
levels(daily_visits_totals$restore_loc)
```

```{r}
#| label: plot-sorted-ascending

ggplot(data = daily_visits_totals,
       aes(x = daily_visits, y = restore_loc,
           fill = visitor_type)) +
  geom_col() +
  labs(x = "Number of Visits", y = "Restoration Location",
       fill = "Type of Visitor",
       title = "Total Visits to Delta Restoration Areas by Visitor Type",
       subtitle = "Sum of all visits during study period") +
  scale_x_continuous(breaks = seq(0, 120, 20), expand = c(0, 0)) +
  my_theme
```

### Reverse Order — Most Visited at Top

```{r}
#| label: fct-reorder-descending

daily_visits_totals <- daily_visits_loc %>%
  group_by(restore_loc) %>%
  mutate(n = sum(daily_visits)) %>%
  ungroup() %>%
  mutate(restore_loc = fct_reorder(restore_loc, desc(n)))

ggplot(data = daily_visits_totals,
       aes(x = daily_visits, y = restore_loc,
           fill = visitor_type)) +
  geom_col() +
  labs(x = "Number of Visits", y = "Restoration Location",
       fill = "Type of Visitor",
       title = "Total Visits to Delta Restoration Areas by Visitor Type",
       subtitle = "Sum of all visits during study period") +
  scale_x_continuous(breaks = seq(0, 120, 20), expand = c(0, 0)) +
  my_theme
```

------------------------------------------------------------------------

## Section 6: Accessible Color with viridis + ggsave()

::: {.callout-tip appearance="minimal"}
**Slide cue: Slides 14–15 (viridis + ggsave)**

Mention the accessibility rationale before running: \~1 in 12 men and 1
in 200 women have some form of color vision deficiency.

After saving, navigate to the figures/ folder in RStudio's Files pane
and click the file to open it — show students the output.
:::

### scale_fill_viridis_d()

```{r}
#| label: viridis-color

ggplot(data = daily_visits_totals,
       aes(x = daily_visits, y = restore_loc,
           fill = visitor_type)) +
  geom_col() +
  scale_fill_viridis_d() +
  labs(
    x        = "Number of Visits",
    y        = "Restoration Location",
    fill     = "Type of Visitor",
    title    = "Total Visits to Delta Restoration Areas by Visitor Type",
    subtitle = "Sum of all visits during study period"
  ) +
  scale_x_continuous(breaks = seq(0, 120, 20), expand = c(0, 0)) +
  my_theme
```

### ggsave()

```{r}
#| label: ggsave-demo
#| eval: false

# Create the figures directory if it doesn't exist
dir.create("figures", showWarnings = FALSE)

# Save the last rendered plot
ggsave("figures/visit_restore_site_delta.jpg",
       width = 12, height = 6, units = "in")

# Save a named object at publication resolution
ggsave("figures/visit_restore_site_delta.png",
       plot   = last_plot(),
       width  = 12,
       height = 6,
       units  = "in",
       dpi    = 300)
```

------------------------------------------------------------------------

## Section 7: facet_wrap() — Small Multiples

::: {.callout-tip appearance="minimal"}
**Slide cue: Slides 16–17 (facet_wrap)**

Connect back to Tufte before running: "This is literally Tufte's small
multiples concept implemented in one function call."

Point out scales = "free_y" explicitly — run once with and once without
so students can see the difference.
:::

### Build the Faceted Plot

```{r}
#| label: facet-wrap-plot
#| fig-width: 12
#| fig-height: 8

facet_plot <- ggplot(data = daily_visits_totals,
                     aes(x = visitor_type, y = daily_visits,
                         fill = visitor_type)) +
  geom_col() +
  facet_wrap(
    ~restore_loc,
    scales = "free_y",   # each panel gets its own y-axis scale
    ncol   = 5,
    nrow   = 2
  ) +
  scale_fill_viridis_d() +
  labs(
    x        = "Type of Visitor",
    y        = "Number of Visits",
    title    = "Total Visits to Delta Restoration Areas",
    subtitle = "Sum of all visits during study period"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.ticks.x    = element_blank(),
    axis.text.x     = element_blank()   # legend handles labeling
  )

facet_plot
```

```{r}
#| label: save-facet-plot
#| eval: false

ggsave("figures/visit_restore_site_facet.jpg",
       plot   = facet_plot,
       width  = 12,
       height = 8,
       units  = "in")
```

------------------------------------------------------------------------

## Section 8: Interactive Visualization

::: {.callout-tip appearance="minimal"}
**Slide cue: Slides 18–21 (Interactive viz)**

Remind students that these outputs are HTML-only — they won't appear in
a static PDF. Render this document to HTML to see them.

The leaflet maps will display interactively in the rendered HTML.
Students can pan, zoom, and click the popups.
:::

### Unique Locations Data Frame

```{r}
#| label: locations-df

locations <- visits_long %>%
  distinct(restore_loc, .keep_all = TRUE) %>%
  select(restore_loc, latitude, longitude)

locations
```

### Interactive Table with DT

```{r}
#| label: dt-table

datatable(locations)
```

### Interactive Map — OpenStreetMap Markers

```{r}
#| label: leaflet-basic

leaflet(locations) %>%
  addTiles() %>%                   # OpenStreetMap base layer
  addMarkers(
    lng   = ~longitude,
    lat   = ~latitude,
    popup = ~restore_loc           # clickable label
  )
```

### Interactive Map — Styled Circle Markers

```{r}
#| label: leaflet-circle-markers

leaflet(locations) %>%
  addTiles() %>%
  addCircleMarkers(
    lng         = ~longitude,
    lat         = ~latitude,
    popup       = ~restore_loc,
    radius      = 5,
    fillColor   = "salmon",
    fillOpacity = 1,
    stroke      = TRUE,
    weight      = 0.5,
    color       = "white",
    opacity     = 1
  )
```

### Interactive Map — Layered USGS Tiles

```{r}
#| label: leaflet-usgs-tiles

leaflet(locations) %>%
  addWMSTiles(
    "https://basemap.nationalmap.gov/arcgis/services/USGSImageryTopo/MapServer/WmsServer",
    layers  = "0",
    options = WMSTileOptions(format = "image/png", transparent = TRUE)
  ) %>%
  addWMSTiles(
    "https://basemap.nationalmap.gov/arcgis/services/USGSHydroCached/MapServer/WmsServer",
    layers  = "0",
    options = WMSTileOptions(format = "image/png", transparent = TRUE)
  ) %>%
  addCircleMarkers(
    lng         = ~longitude,
    lat         = ~latitude,
    popup       = ~restore_loc,
    radius      = 5,
    fillColor   = "salmon",
    fillOpacity = 1,
    stroke      = TRUE,
    weight      = 0.5,
    color       = "white",
    opacity     = 1
  )
```

------------------------------------------------------------------------

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
