---
title: "Interactive Visualization — Live Coding Companion"
subtitle: "DT and leaflet with the Sacramento Delta Dataset"
author: "BIO/CHEM 806"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: false
    theme: cosmo
execute:
  echo: true
  eval: true
editor:
  markdown:
    wrap: 72
---

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(leaflet)
library(DT)
library(scales)
library(janitor)
library(viridis)
```

::: callout-tip
## Instructor Note — Before Class (Slide 1)

Have students open their existing Quarto document from the ggplot2
activity, or create a new one inside the same project. They should
already have `visits_long` and `daily_visits_totals` in their
environment from last time. If not, the data-prep section below rebuilds
everything from scratch.

Confirm that all packages in the setup chunk above are installed. `DT`
and `leaflet` are the new ones — students may need to run
`install.packages(c("DT", "leaflet"))` in the console.
:::

------------------------------------------------------------------------

## Data Preparation

Rebuild `visits_long` if students don't have it from the ggplot2
session. If they do, skip to the next section.

```{r load-data, message=FALSE, warning=FALSE}
delta_visits_raw <- read_csv(
  here::here("data/r_data_visualization/Socioecological_monitoring_data.csv")
)

delta_visits <- delta_visits_raw %>%
  janitor::clean_names()

visits_long <- delta_visits %>%
  pivot_longer(
    cols = c(sm_boat, med_boat, lrg_boat,
             bank_angler, scientist, cars),
    names_to = "visitor_type",
    values_to = "quantity"
  ) %>%
  rename(restore_loc = eco_restore_approximate_location) %>%
  select(-notes)
```

::: callout-tip
## Instructor Note — Data Setup (Slide 5)

If anyone gets a file-not-found error, they either haven't downloaded
the EDI data yet, or they're not in their R project. The most common
fix: File → Open Project → navigate to their `import_template_USERNAME`
folder and open the `.Rproj` file.

Don't spend more than 3 minutes on this — have students pair up if
they're stuck.
:::

------------------------------------------------------------------------

## Part 1: Interactive Tables with DT

### Getting one row per site

The `visits_long` data has many rows per site — one for each visit ×
visitor type combination. Before building a location table, we need
exactly one row per site.

```{r distinct-locations}
locations <- visits_long %>%
  distinct(restore_loc, .keep_all = TRUE) %>%
  select(restore_loc, latitude, longitude)

locations
```

::: callout-tip
## Instructor Note — `distinct()` (Slide 9)

Ask the chat: "What happens if we run `distinct(restore_loc)` without
`.keep_all = TRUE`? Try it." Give 30 seconds.

The answer: you only get one column back. `.keep_all = TRUE` is what
preserves the latitude and longitude columns. This is a genuinely common
mistake — worth surfacing before the activity.
:::

### Basic `datatable()`

```{r basic-datatable}
datatable(locations)
```

### Customized for scientific communication

```{r custom-datatable}
datatable(
  locations,
  options = list(
    pageLength = 5,
    dom = 'ftp'
  ),
  colnames = c("Restoration Site", "Latitude", "Longitude"),
  caption = "Sampling locations, Sacramento–San Joaquin Delta."
)
```

::: callout-tip
## Instructor Note — DT Demo (Slides 10–11)

Run the basic version first, then the customized one. Point out:

1.  Click a column header to sort — ask "what does sorting by latitude
    tell us about the north–south arrangement of these sites?"
2.  Type "Island" in the search box — instant filter
3.  The `dom = 'ftp'` argument: `f` = filter, `t` = table, `p` =
    pagination. Removing the default buttons keeps it clean for a
    supplementary document.

Chat prompt (Slide 11): "Think about a table in your thesis or a paper
you've read. What would have been easier to find if it was searchable?"
Give 90 seconds, call on 2–3 people.
:::

------------------------------------------------------------------------

## Part 2: Interactive Maps with leaflet

### Simplest possible map

```{r basic-map}
leaflet(locations) %>%
  addTiles() %>%
  addMarkers(
    lng = ~ longitude,
    lat = ~ latitude,
    popup = ~ restore_loc
  )
```

::: callout-tip
## Instructor Note — First leaflet Map (Slide 15)

Run this live, click one of the markers to show the popup working. Ask:
"How does the `~longitude` syntax feel familiar? Where have you seen `~`
used to reference a column in data?" (Answer: `aes()` in ggplot2,
`facet_wrap(~variable)` — the formula interface.)

Zoom in on the Delta region to show the geographic context. Most
students won't have spent much time looking at this area.
:::

### Switching base maps

```{r topo-map}
leaflet(locations) %>%
  addProviderTiles(providers$Esri.WorldTopoMap) %>%
  addCircleMarkers(
    lng = ~ longitude,
    lat = ~ latitude,
    popup = ~ restore_loc,
    radius = 5,
    fillColor = "#167B62",
    fillOpacity = 0.8,
    stroke = FALSE
  )
```

```{r satellite-map}
leaflet(locations) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addCircleMarkers(
    lng = ~ longitude,
    lat = ~ latitude,
    popup = ~ restore_loc,
    radius = 5,
    fillColor = "salmon",
    fillOpacity = 0.9,
    stroke = FALSE
  )
```

::: callout-tip
## Instructor Note — Provider Tiles (Slide 17)

Run both maps back to back. Ask: "For studying restoration ecology in
the Delta, which base map tells you more — topographic or satellite
imagery?" Pause for chat responses.

This is not a trick question — satellite is often more useful for
habitat context. But for sites spanning an elevation gradient (e.g., a
mountain watershed), topographic would win. This connects to the
decision log framing: base map choice is a defensible scientific
decision.

Mention: there are hundreds of provider tiles available.
`leaflet::providers` is a list you can explore.
:::

### Layered base maps (imagery + labels)

```{r layered-map}
leaflet(locations) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addProviderTiles(providers$CartoDB.PositronOnlyLabels) %>%
  addCircleMarkers(
    lng = ~ longitude,
    lat = ~ latitude,
    popup = ~ restore_loc,
    radius = 5,
    fillColor = "yellow",
    fillOpacity = 1,
    stroke = TRUE,
    weight = 0.5,
    color = "white",
    opacity = 1
  )
```

::: callout-tip
## Instructor Note — Layered Tiles (Slide 18)

Run this. Note that you're calling `addProviderTiles()` twice — the
second call stacks on top of the first. The white-on-dark circle markers
are the most legible against satellite imagery.
:::

### Informative popups

```{r popup-map}
leaflet(locations) %>%
  addProviderTiles(providers$Esri.WorldTopoMap) %>%
  addCircleMarkers(
    lng = ~ longitude,
    lat = ~ latitude,
    popup = ~ paste0(
      "<b>", restore_loc, "</b><br>",
      "Lat: ", round(latitude, 3), "<br>",
      "Lon: ", round(longitude, 3)
    ),
    radius = 6,
    fillColor = "#167B62",
    fillOpacity = 0.85,
    stroke = TRUE,
    weight = 1,
    color = "white",
    opacity = 1
  )
```

::: callout-tip
## Instructor Note — HTML Popups (Slide 20)

Walk through the `paste0()` call:

-   `<b>...</b>` = bold in HTML
-   `<br>` = line break
-   `round(latitude, 3)` = 3 decimal places is appropriate precision for
    field GPS coordinates

Chat prompt (Slide 20): "What three variables from your thesis dataset
would be most valuable in a popup?" Give 2 minutes. Call on 2–3 students
— this usually surfaces really interesting domain variation between the
bio and chem students.
:::

------------------------------------------------------------------------

## Putting It Together

A complete workflow: wrangle → table → map, all in one document.

```{r full-workflow}
# Step 1: Prepare unique locations
locations <- visits_long %>%
  distinct(restore_loc, .keep_all = TRUE) %>%
  select(restore_loc, latitude, longitude)

# Step 2: Searchable table for supplementary methods
datatable(
  locations,
  colnames = c("Site", "Latitude", "Longitude"),
  caption = "Table S1. Restoration monitoring locations."
)

# Step 3: Interactive map for GitHub Pages / supplementary report
leaflet(locations) %>%
  addProviderTiles(providers$Esri.WorldImagery) %>%
  addProviderTiles(providers$CartoDB.PositronOnlyLabels) %>%
  addCircleMarkers(
    lng = ~ longitude,
    lat = ~ latitude,
    popup = ~ paste0("<b>", restore_loc, "</b>"),
    radius = 5,
    fillColor = "salmon",
    fillOpacity = 1,
    stroke = TRUE, weight = 0.5,
    color = "white", opacity = 1
  )
```

------------------------------------------------------------------------

## Breakout Activity (8 min)

::: callout-tip
## Instructor Note — Breakout (Slide 22)

Put students in pairs in breakout rooms. Prompt:

"Thinking about your own thesis data or a dataset you'd plausibly use:
(1) What are your coordinate columns, or how would you get coordinates?
(2) What one additional variable — beyond just location — would most
useful to see in a popup? (3) Which base map — topo, satellite, or
street — would best show the scientific context of your sites?"

After 7 minutes, bring everyone back and call on 2–3 pairs to share.
This typically takes 3–4 more minutes. Budget 12 minutes total. Cut to 6
minutes if you're running behind; this can continue as an activity
warm-up.
:::

------------------------------------------------------------------------

## Key Concepts Recap

| Concept | What to remember |
|-------------------------|----------------------------------------------|
| `distinct(.keep_all = TRUE)` | One row per site; keep lat/lon |
| `datatable()` | Searchable/sortable with one function call |
| `leaflet() %>% addTiles()` | Canvas + base map |
| `addProviderTiles()` | Swap base map for scientific context |
| `addCircleMarkers()` | Styleable markers: radius, color, popup |
| `~column` | References data frame column (same as `aes()` logic) |
| `paste0()` in popup | HTML string — bold, line breaks, rounded values |

::: callout-tip
## Instructor Note — Wrap-Up (Slide 24)

The most important conceptual takeaway for thesis students: ggplot2 is
for your chapters; DT + leaflet are for your GitHub Pages supplementary
materials and advisor-facing exploratory reports. They serve different
audiences and different stages of the workflow.

Remind them that in Week 13 they'll revisit spatial data with `sf`
objects and `geom_sf()` — today's lat/lon columns will become properly
georeferenced spatial objects with CRS metadata.
:::
