---
title: "Week 6 Lecture 2 — Reshaping and Its Consequences (Code Companion)"
format: html
execute:
  echo: true
  warning: true
  message: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(comment = "#>")
```

## Libraries

Note: the tidyverse is a set of packages built around a shared design philosophy — consistent function names, consistent argument order, consistent use of data frames as inputs and outputs, and readable pipelines. Once you know one tidyverse package, the others feel familiar. That consistency is the point. Connect it to what they already know from Week 5: tidy data is a principle for organizing data, not a tool. The tidyverse is a collection of tools designed to work naturally with data that's already tidy. The two things reinforce each other — if your data is tidy, tidyverse functions are easy to apply; if you use tidyverse functions, you naturally produce tidy output.

library(tidyverse) loads eight or so packages at once. It's convenient but worth explaining what it actually does — it loads ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, and forcats together.

-   reading, cleaning, and reshaping data, `readr`, `dplyr`, and `tidyr`

-   plotting, `ggplot2`.

-   working with strings, `stringr`

-   working with dates, `lubridate`.

-   working with categorical variables, `forcats`

In a course context that's fine. In production code or a package, loading the specific packages you need is better practice because it **makes dependencies explicit and avoids masking problems.**

```{r}
#| label: libraries
library(tidyverse)  # dplyr + tidyr + ggplot2 + friends
library(readr)
```

# Data sources

## Read the data

Read the data in from KNB Package Alaska commercial salmon catches by management region (1886-1997)

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

------------------------------------------------------------------------

## `select()` — keeping what you need

```{r}
#| label: select-keep
## keep what you want
catch_data <- catch_original %>%
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

## OR drop what you don't want
catch_data <- catch_original %>%
  select(-All, -notesRegCode)
```

## Quality check with `glimpse()`

```{r}
#| label: quality-check
dplyr::glimpse(catch_data)
```

------------------------------------------------------------------------

## `mutate()` — first attempt at type conversion

```{r}
#| label: mutate-as-numeric-first
catch_clean <- catch_data %>%
  mutate(Chinook = as.numeric(Chinook))
```

Find the row(s) that became `NA` after coercion:

```{r}
#| label: find-na
i <- which(is.na(catch_clean$Chinook))
i

catch_data[i, ]
```

------------------------------------------------------------------------

## `mutate()` with `if_else()` — fixing the `"I"` problem

```{r}
#| label: mutate-if-else-fix
catch_clean <- catch_data %>%
  mutate(
    Chinook = if_else(
      condition = Chinook == "I",
      true = "1",
      false = Chinook
    ),
    Chinook = as.numeric(Chinook)
  )
```

Verify:

```{r}
#| label: verify-fix
catch_clean[i, ]
```

------------------------------------------------------------------------

## `pivot_longer()` — wide → long

```{r}
#| label: pivot-longer
catch_long <- catch_clean %>%
  pivot_longer(
    cols = -c(Region, Year),
    names_to = "species",
    values_to = "catch"
  )

catch_long
```

## `pivot_wider()` — long → wide (reverse)

```{r}
#| label: pivot-wider
catch_wide <- catch_long %>%
  pivot_wider(
    names_from = species,
    values_from = catch
  )

catch_wide
```

------------------------------------------------------------------------

## `rename()` — defensive coding

```{r}
#| label: rename-catch
catch_long <- catch_long %>%
  rename(catch_thousands = catch)

catch_long
```

------------------------------------------------------------------------

## Unit conversion with `mutate()` + `select()`

```{r}
#| label: unit-conversion
catch_long <- catch_long %>%
  mutate(catch = catch_thousands * 1000) %>%
  select(-catch_thousands)

catch_long
```

------------------------------------------------------------------------

## Split–Apply–Combine with `group_by()` + `summarize()`

```{r}
#| label: group-summarize
mean_region <- catch_long %>%
  group_by(Region) %>%
  summarize(mean_catch = mean(catch))

mean_region
```

### Counting with `n()` and `count()`

```{r}
#| label: counting
n_region <- catch_long %>%
  group_by(Region) %>%
  summarize(n = n())

n_region

## Shortcut:
catch_long %>% count(Region)
```

------------------------------------------------------------------------

## `filter()` — keeping rows by condition

```{r}
#| label: filter-examples
sse_catch <- catch_long %>%
  filter(Region == "SSE")

chinook_sse <- catch_long %>%
  filter(
    Region == "SSE",
    species == "Chinook"
  )

sse_catch
chinook_sse
```

------------------------------------------------------------------------

## `arrange()` — sorting rows

```{r}
#| label: arrange-examples
## ascending (default)
mean_region %>% arrange(mean_catch)

## descending
mean_region %>% arrange(desc(mean_catch))
```

------------------------------------------------------------------------

## `separate()` + `unite()` — splitting and combining columns

These are standalone examples from lecture (they use example data frames).

```{r}
#| label: separate-example
# Example: sites_df has a column called site, like "HAW-101"
# sites_df %>%
#   separate(site, c("island", "site_number"), "-")

# Tiny reproducible example:
sites_df <- tibble(site = c("HAW-101", "MOL-023"))
sites_df %>%
  separate(site, c("island", "site_number"), "-")
```

```{r}
#| label: unite-example
# Example: dates_df has year/month/day columns
# dates_df %>%
#   unite(date, year, month, day, sep = "-")

# Tiny reproducible example:
dates_df <- tibble(year = 1930, month = 12, day = 14)
dates_df %>%
  unite(date, year, month, day, sep = "-")
```

------------------------------------------------------------------------

## All together — the complete pipeline

```{r}
#| label: complete-pipeline
mean_region_pipeline <- catch_original %>%
  select(-All, -notesRegCode) %>%
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>% # data entry erro
  mutate(Chinook = as.numeric(Chinook)) %>%
  pivot_longer(
    -c(Region, Year),
    names_to = "species",
    values_to = "catch"
  ) %>%
  mutate(catch = catch * 1000) %>%
  group_by(Region) %>%
  summarize(mean_catch = mean(catch)) %>%
  arrange(desc(mean_catch))

mean_region_pipeline
```
