---
title: "Week 13 Live Demo: Working with Spatial Data in R"
subtitle: "BIO/CHEM 806 · Instructor Companion"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-title: "Demo sections"
    number-sections: false
    theme: cosmo
    code-tools: true
    self-contained: true
execute:
  echo: true
  eval: false
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| eval: false
library(readr)
library(here)
library(sf)
library(ggplot2)
library(leaflet)
library(scales)
library(ggspatial)
library(dplyr)
```

::: callout-tip
## How to use this document

This is your live-coding companion for the Week 13 demo and activity.
Each section maps to a part of the student assignment and the textbook
(s15). Green callout boxes are instructor cues — things to say or do
before running the code. Run code chunks top to bottom; everything
builds on what came before.

**Before class:** download the KNB data (run the setup chunk below in
your Console), confirm all packages load, and render this document once
to check it works.
:::

------------------------------------------------------------------------

## 0 · Setup — Get the data

::: callout-tip
## Before running this in class

Run the chunk below **once in your Console** before the demo — not live,
because the download takes time. Have the `data/` folder ready with the
three files already in it.

If a student asks how to download: share the code block below and tell
them to run it in their Console, not inside their Quarto file.
:::

```{r}
#| label: download-data
#| eval: false

# Run once in your Console — not inside your Quarto file
knb_url <- "https://dev.nceas.ucsb.edu/knb/d1/mn/v2/object/urn%3Auuid%3Aaceaecb2-1ce0-4d41-a839-d3607d32bb58"

download.file(url = knb_url, destfile = "shapefile_demo_data.zip")
unzip("shapefile_demo_data.zip", exdir = "data")
file.remove("shapefile_demo_data.zip")
```

After running, your `data/` folder should contain:

- `ak_regions_simp.shp` — Alaska regional boundary polygons
- `alaska_population.csv` — population by community (point data)
- `ak_rivers_simp.shp` — Alaska river network (line data)

------------------------------------------------------------------------

## 1 · Read and explore the shapefile

::: callout-tip
## Say this

"We read shapefiles with `read_sf()`. Notice the `_sf` suffix on the
object name — that's our convention to remind us this object has a
geometry column attached.

A **shapefile** is a common format for storing vector spatial data.
Think of it like a CSV for spatial objects — except instead of rows of
numbers, it stores points, lines, or polygons. The `.shp` file always
comes with companion files (`.dbf` for attribute data, `.prj` for
projection info) — keep them together.

The `plot()` call gives us a quick look at every attribute column mapped
spatially. Does anything look off?"

Let students respond. The map will look distorted — that's a CRS issue.
Don't explain it yet; let the distortion sit as a question.
:::

```{r}
#| label: read-shapefile

ak_rgns_sf <- read_sf(here("data/ak_regions_simp.shp"))

# Quick visual — one panel per attribute column, each shown as a map
plot(ak_rgns_sf)
```

::: callout-tip
## Say this

"What two classes does this object have? Let's check."
:::

```{r}
#| label: class-check

class(ak_rgns_sf)
```

::: callout-tip
## Say this

"`sf` and `data.frame`. Two classes at once. That means everything you
already know — `head()`, `glimpse()`, `select()`, `filter()` — works
here, because it *is* a data frame. The `sf` class adds spatial
awareness on top. Let's look at the structure."
:::

```{r}
#| label: explore-sf

head(ak_rgns_sf)
glimpse(ak_rgns_sf)
```

::: callout-tip
## Point out

Highlight the `geometry` column at the end of the `head()` output.

"This is what makes an `sf` object different from a plain data frame.
The `geometry` column holds the actual shapes — lists of coordinates
that define each polygon boundary. And it's **sticky** — if you
`select()` just three of ten columns, the geometry column comes along
anyway, even if you didn't ask for it. R makes it hard to accidentally
lose your spatial information."

Also point to the header lines printed above the data frame:
`geometry type`, `bbox` (bounding box — the spatial extent, like the min
and max x and y coordinates of the dataset), and `CRS`.
:::

------------------------------------------------------------------------

## 2 · Check and transform the CRS

::: callout-tip
## Say this

"Before we do anything spatial, we check the CRS. A **CRS — Coordinate
Reference System** — is the system that tells R how to interpret the
coordinate numbers. Think of it as the map's coordinate language. Two
datasets speaking different coordinate languages can't be compared until
you translate one to match the other.

Make checking the CRS a reflex — always know what coordinate system your
data is in before you start working with it."
:::

```{r}
#| label: check-crs

st_crs(ak_rgns_sf)
```

::: callout-tip
## Say this

"The output looks dense, but the key things to find: the CRS name
(WGS84) and the EPSG code (4326). **WGS84** is a geographic CRS —
coordinates are in degrees of latitude and longitude, not meters. No
flattening has been applied to the sphere. That's why the plot looked
distorted — Alaska spans a huge longitude range and the naïve lat/lon
display squashes it.

For Alaska, a much better choice is **Alaska Albers — EPSG 3338** — an
equal-area projected CRS designed for Alaska. Equal-area means every
polygon has the right relative size, which matters if you're ever
calculating areas. Let's transform."
:::

```{r}
#| label: transform-crs

ak_rgns_3338_sf <- ak_rgns_sf %>%
    st_transform(crs = 3338)

# Plot again — notice the difference
plot(ak_rgns_3338_sf)
```

::: callout-tip
## Point out

"Much better — Alaska isn't squashed across the dateline anymore. The
proportions look right. Notice the naming convention: `_3338` in the
object name tells anyone reading the code exactly what coordinate system
this object is in. When you have multiple versions of the same data in
different projections, this convention prevents confusion."
:::

------------------------------------------------------------------------

## 3 · sf + tidyverse

::: callout-tip
## Say this

"Because `sf` objects are data frames, your tidyverse skills transfer
directly. Watch what happens to the geometry column."
:::

```{r}
#| label: select-demo

# select() — geometry stays even though we didn't ask for it
ak_rgns_3338_sf %>%
    select(region)
```

```{r}
#| label: filter-demo

# filter() works exactly as expected
unique(ak_rgns_3338_sf$region)

ak_rgns_3338_sf %>%
    filter(region == "Southeast")
```

::: callout-tip
## Chat prompt (optional)

"Drop in chat: how many distinct regions does `unique()` return?"
Answer: 13.
:::

------------------------------------------------------------------------

## 4 · CSV with coordinates → sf object

::: callout-tip
## Say this

"Most scientific data doesn't arrive as a shapefile — it arrives as a
CSV. GPS exports from field instruments, GBIF species occurrence
downloads, water quality monitoring tables, disease surveillance records
from public health databases — if it has latitude and longitude columns,
`st_as_sf()` promotes it to a spatial object in one step.

This is often the first line of your own spatial analysis: you have a
spreadsheet of locations, and you want to put them on a map."
:::

```{r}
#| label: read-population

# _df suffix: plain data frame for now — no geometry yet
pop_df <- read_csv(here("data/alaska_population.csv"))

head(pop_df)
```

```{r}
#| label: csv-to-sf

# _sf suffix: now spatial; _4326 tells us the CRS
pop_4326_sf <- st_as_sf(pop_df,
                        coords = c('lng', 'lat'),  # longitude first, then latitude
                        crs = 4326,
                        remove = FALSE)  # keep the raw lat/lng columns

head(pop_4326_sf)
```

::: callout-tip
## Point out

"Three things to notice:

1.  `coords` must be **longitude first, then latitude** — x before y,
    even though we usually say 'lat/lon.' This trips people up.

2.  `crs = 4326` tells R what coordinate system the numbers are in.
    Almost all GPS data and most downloaded data uses WGS84 (4326)
    unless told otherwise.

3.  `remove = FALSE` keeps the original `lat` and `lng` columns in the
    data frame. We need them later for `leaflet`, which wants raw
    numeric columns, not just the geometry column."
:::

------------------------------------------------------------------------

## 5 · Spatial join — hit the error on purpose

::: callout-tip
## Say this

"Now we want to assign each community to its Alaska region — a spatial
join. Instead of matching on a shared key column, we'll match by
geography: which polygon does each point fall inside?

Let's try it directly." Then run the code and let it fail.
:::

```{r}
#| label: join-attempt-fails

# This will fail — run it anyway so students see the error
pop_joined_sf <- st_join(pop_4326_sf,
                         ak_rgns_3338_sf,
                         join = st_within)
```

::: callout-tip
## Say this

"`Error: st_crs(x) == st_crs(y) is not TRUE` — R is protecting us.

We tried to spatially join two datasets in different coordinate systems:
one in degrees (4326), one in meters (3338). Joining them directly would
be like overlaying two maps printed at different scales — the result
would be nonsense.

This is not a bug. R caught a real problem. The message tells you
exactly what went wrong and how to think about the fix: the CRS of x and
y must be equal. The fix is always `st_transform()` — translate one to
match the other, then join."
:::

```{r}
#| label: transform-before-join

pop_3338_sf <- st_transform(pop_4326_sf, crs = 3338)
```

```{r}
#| label: join-success

pop_joined_sf <- st_join(pop_3338_sf,
                         ak_rgns_3338_sf,
                         join = st_within)

head(pop_joined_sf)
```

::: callout-tip
## Point out

"Each community row now has a `region` column — the Alaska region it
falls inside. That's the spatial join. Now we can aggregate by region to
get totals."
:::

------------------------------------------------------------------------

## 6 · Aggregate by region

::: callout-tip
## Say this

"We want total population per region. If we call `group_by()` and
`summarize()` directly on an `sf` object, R tries to aggregate the
geometries too — combining all the city points within each region into a
single MULTIPOINT geometry. That's not what we want; we want a plain
numeric sum. So we drop the geometry first, then summarize."
:::

```{r}
#| label: drop-geometry-and-summarize

# _df suffix: no longer spatial after st_drop_geometry()
pop_rgn_df <- pop_joined_sf %>%
    st_drop_geometry() %>%
    group_by(region) %>%
    summarize(total_pop = sum(population))

head(pop_rgn_df)
```

```{r}
#| label: rejoin-to-polygons

# Regular left_join — not spatial
# Join totals back to the polygon sf object so we can map them
pop_rgn_3338_sf <- left_join(ak_rgns_3338_sf,
                             pop_rgn_df,
                             by = "region")

# Quick check — color each polygon by total population
plot(pop_rgn_3338_sf["total_pop"])
```

::: callout-tip
## Summarize the pattern

"Three steps — you'll use this constantly in your own spatial work:

1.  `st_join()` — points get polygon attributes (the region they belong
    to)
2.  `st_drop_geometry()` + `group_by()` + `summarize()` — aggregate as a
    plain data frame, avoiding geometry aggregation
3.  `left_join()` — reattach the aggregated values to the polygon
    geometry for mapping"
:::

------------------------------------------------------------------------

## 7 · Save the spatial object

```{r}
#| label: write-sf

write_sf(pop_rgn_3338_sf, here("data/ak_regions_population.shp"))
```

::: callout-tip
## Say this

"Save your processed spatial objects so you don't have to rebuild them
every session. This belongs in your numbered script sequence — the same
reproducibility practice we've followed all semester.

A `.shp` file is the standard interoperable format — readable in QGIS,
ArcGIS, and by collaborators who don't use R. A more modern alternative
is **GeoPackage** (`.gpkg`) — a single file (no companion files needed)
with no column-name length restrictions. Both read back with
`read_sf()`."
:::

------------------------------------------------------------------------

## 8 · Static map with ggplot2

::: callout-tip
## Say this

"`geom_sf()` reads the geometry column automatically — no `x` and `y`
aesthetics needed. R knows where the shapes are. Everything else —
`aes()`, `scale_fill_*`, `labs()`, `theme_*` — works exactly like any
other ggplot."
:::

```{r}
#| label: ggplot-choropleth

ggplot(pop_rgn_3338_sf) +
    geom_sf(aes(fill = total_pop)) +
    labs(fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high = "firebrick",
                          labels = comma) +
    theme_bw()
```

::: callout-tip
## Say this

"Now let's add the rivers and community locations. Three `geom_sf()`
layers, three geometry types — polygons, lines, points — all in one
plot. **Layer order matters: bottom layer renders first.** Polygons go
first (they'd cover everything else), then lines, then points on top."
:::

```{r}
#| label: read-rivers

rivers_3338_sf <- read_sf(here("data/ak_rivers_simp.shp"))

st_crs(rivers_3338_sf)
```

::: callout-tip
## Point out

"No explicit EPSG code is set on the rivers file — a common real-world
situation. But the proj4 string contains enough information to identify
it. Search the string on [epsg.io](https://epsg.io) and it confirms this
is Alaska Albers (3338). We check before layering — same CRS rule as
always."
:::

```{r}
#| label: three-layer-map

ggplot() +
    geom_sf(data = pop_rgn_3338_sf,
            aes(fill = total_pop)) +
    geom_sf(data = pop_3338_sf,
            size = 0.5) +
    geom_sf(data = rivers_3338_sf,
            aes(linewidth = StrOrder)) +
    scale_linewidth(range = c(0.05, 0.5),
                    guide = "none") +
    labs(title = "Total Population by Alaska Region",
         fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high = "firebrick",
                          labels = comma) +
    theme_bw()
```

::: callout-tip
## Decision log moment

"This map is a set of choices — which color scale, whether to show
rivers at all, how to scale river width by stream order, whether to
label the points. These are methods decisions. Note them in your
decision log: not just *what* you chose, but *why*. 'Used a sequential
color scale (khaki to firebrick) to show population magnitude; no
diverging scale because there is no meaningful zero or reference
point.'"
:::

------------------------------------------------------------------------

## 9 · Basemap with ggspatial (optional)

::: callout-tip
## Say this

"Web tile basemaps — like OpenStreetMap — use **Pseudo-Mercator (EPSG
3857)**, the projection behind every major web mapping service including
Google Maps. We need to transform our data to that same CRS before
layering tiles on top of it.

This step takes a minute or two to load tiles over the network — I'll
run it now and explain while it loads."
:::

```{r}
#| label: ggspatial-basemap

pop_3857_sf <- st_transform(pop_3338_sf, crs = 3857)

ggplot(data = pop_3857_sf) +
    ggspatial::annotation_map_tile(type = "osm",
                                   zoom = 4,
                                   progress = "none") +
    geom_sf(aes(color = population), fill = NA) +
    scale_color_continuous(low = "darkkhaki",
                           high = "firebrick",
                           labels = comma)
```

::: callout-tip
## Say this

"Does the basemap add value here — or does it compete with the data?
That's a genuine question and a design decision.

A heuristic: use a basemap when your audience needs geographic context
(roads, coastlines, city names help them orient) or when point locations
relative to physical features matter for interpretation. Skip it when
the basemap competes visually with your data or when you're making a
publication figure that will be printed.

For your activity: Part 5 is optional stretch. If tile loading hangs,
skip it and note that in your decision log."
:::

------------------------------------------------------------------------

## 10 · Interactive map with leaflet

::: callout-tip
## Say this

"`leaflet` creates browser-rendered interactive maps — your audience can
zoom, click, and hover over features. Two key differences from ggplot:
it uses `%>%` pipes not `+`, and it needs your data in WGS84 (EPSG
4326).

For the Alaska Albers projection specifically, we define a custom CRS
object so leaflet renders the map with the correct regional projection
on the client side. This is boilerplate — copy it from the textbook, you
don't need to memorize the proj4 string."
:::

```{r}
#| label: leaflet-crs-setup

epsg3338 <- leaflet::leafletCRS(
    crsClass = "L.Proj.CRS",
    code = "EPSG:3338",
    proj4def = "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154
                +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0
                +units=m +no_defs",
    resolutions = 2 ^ (16:7)
)

pop_rgn_4326_sf <- pop_rgn_3338_sf %>%
    st_transform(crs = 4326)
```

::: callout-tip
## Say this

"Build leaflet maps in stages — start simple to confirm the plumbing
works before adding visual complexity."
:::

```{r}
#| label: leaflet-stage-1

# Stage 1: confirm polygons render
m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(data = pop_rgn_4326_sf,
                fillColor = "gray",
                weight = 1)
m
```

```{r}
#| label: leaflet-stage-2

# Stage 2: add color scale and legend
pal <- colorNumeric(palette = "Reds",
                    domain = pop_rgn_4326_sf$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_rgn_4326_sf,
        fillColor = ~pal(total_pop),   # ~ is leaflet's version of aes()
        weight = 1,
        color = "black",
        fillOpacity = 1,
        label = ~region
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_rgn_4326_sf$total_pop),
        title = "Total Population"
    )
m
```

```{r}
#| label: leaflet-stage-3

# Stage 3: add community markers sized by population
m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_rgn_4326_sf,
        fillColor = ~pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1
    ) %>%
    addCircleMarkers(
        data = pop_4326_sf,
        lat = ~lat,
        lng = ~lng,
        radius = ~log(population / 500),   # log scale prevents huge outliers
        fillColor = "gray",
        fillOpacity = 1,
        weight = 0.25,
        color = "black",
        label = ~paste0(city, ", population ", comma(population))
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_rgn_4326_sf$total_pop),
        title = "Total Population"
    )
m
```

::: callout-tip
## Point out

"`radius = ~log(population / 500)` scales circle size logarithmically.
Without the log, Anchorage would be enormous and tiny communities would
be invisible — a common visualization problem with right-skewed data.
That's an intentional design choice that belongs in your decision log.

This interactive map, embedded in a Quarto HTML document and published
to GitHub Pages, is a shareable portfolio piece. Your committee, a
hiring manager, or a funding agency can open the link and explore it.
That's what reproducible, open science communication looks like."
:::

------------------------------------------------------------------------

## Wrap-up

::: callout-tip
## Closing remarks

"You just ran the complete spatial workflow:

- Read shapefiles and CSVs with coordinates
- Check and transform CRS — and hit the mismatch error and fix it
- Spatial join to assign points to polygons
- Aggregate without geometry, then reattach to polygons for mapping
- Three map types: static ggplot, basemap overlay, interactive leaflet

The activity asks you to do exactly this sequence, following the
textbook. The textbook has all the code. Your job is to run it,
understand each step, and document at least two decisions in your
decision log.

Remember the submission convention — add your last name to each file:
`Lastname_intro-to-spatial-data.html`,
`Lastname_intro-to-spatial-data.qmd`, `Lastname_decision_log.html`."
:::
