---
title: "Week 14, Lecture 2: Conditionals, Errors, and ggplot Functions"
subtitle: "Integrated live-demo script with visible teaching cues"
author: "BIO/CHEM 806"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    code-fold: false
    code-tools: true
    theme: cosmo
    embed-resources: true
editor:
  markdown:
    wrap: 72
execute:
  eval: true
  echo: true
  warning: true
  message: true
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(ggplot2)
```

# Wednesday setup: Monday functions and data

::: callout-tip
## Teaching cue

Run this once at the start of class. Do not reteach Monday's lecture
here; this simply restores the function family and example data needed
for today's examples.
:::

```{r}
#| label: monday-functions-and-data

calc_coulombic_efficiency <- function(discharge_cap_mAh, charge_cap_mAh) {
  ce <- (discharge_cap_mAh / charge_cap_mAh) * 100
  return(ce)
}

cycling_data <- data.frame(
  cycle         = 1:5,
  charge_mAh    = c(3.01, 3.00, 2.99, 2.98, 2.97),
  discharge_mAh = c(2.98, 2.97, 2.96, 2.95, 2.94)
)

cycling_data$CE <- calc_coulombic_efficiency(
  cycling_data$discharge_mAh,
  cycling_data$charge_mAh
)

calc_gc_content <- function(sequence) {
  sequence <- toupper(sequence)
  n_total <- nchar(sequence)
  n_gc <- nchar(gsub("[^GC]", "", sequence))
  return((n_gc / n_total) * 100)
}

calc_primer_tm <- function(sequence) {
  sequence <- toupper(sequence)
  n_a <- nchar(gsub("[^A]", "", sequence))
  n_t <- nchar(gsub("[^T]", "", sequence))
  n_g <- nchar(gsub("[^G]", "", sequence))
  n_c <- nchar(gsub("[^C]", "", sequence))
  tm <- 2 * (n_a + n_t) + 4 * (n_g + n_c)
  return(tm)
}

calc_primer_stats <- function(sequence) {
  gc <- calc_gc_content(sequence)
  tm <- calc_primer_tm(sequence)
  return(list(
    sequence   = sequence,
    gc_percent = gc,
    tm_celsius = tm
  ))
}
```

```{r}
#| label: check-setup

cycling_data
calc_primer_stats("ATGCGCATTAGC")
```

# Segment 1 - Warm-up: mental model check (5 min)

::: callout-note
## Read aloud

"Welcome back. Before we add new tools, let's make sure Monday's mental
model is solid."
:::

::: callout-tip
## Chat prompt

Post in chat:

"What does `classify_temp(15)` return, and why? Take 60 seconds - type
your answer before reading others."
:::

```{r}
#| label: classify-temp

classify_temp <- function(celsius) {
  if (celsius > 30) {
    return("hot")
  } else if (celsius < 10) {
    return("cold")
  } else {
    return("moderate")
  }
}
```

```{r}
#| label: test-classify-temp

classify_temp(35)    # "hot"
classify_temp(5)     # "cold"
classify_temp(20)    # "moderate"
classify_temp(15)    # chat warm-up answer
```

::: callout-tip
## Teaching cue

Expected answer: `"moderate"`. The value 15 is not greater than 30 and
not less than 10, so the function reaches the final `else` branch.

Run each call one at a time. Before each call, ask: "What do you
predict?" Wait a few seconds for chat responses, then run the line.

Follow-up question if students answer correctly but cannot explain the
logic: "At which line does the function stop executing? Why doesn't it
reach the later branches?"
:::

# Segment 2 - Functions with conditionals (10 min)

::: callout-note
## Former slides 3-4: `if`/`else` inside functions

Monday's functions all did the same thing every time. Today's functions
make decisions.

The key insight is that `return()` exits the function immediately. The
first true branch wins.
:::

::: callout-tip
## Teaching cue

Say this after the prediction examples:

"Notice that once the function hits `return()`, it is done. It never
checks the remaining conditions. This early-exit pattern is exactly how
we will write our error checks."
:::

# Segment 3 - Battery chemistry context: specific capacity (5 min)

::: callout-note
## Former slide 5: why inputs need validation

"Before we write our next function, I want to motivate the
error-handling tools we are about to learn. And I want to do it through
a real research scenario in battery chemistry."

Read these points aloud and pause on each:

- Specific capacity is measured in mAh/g: how much charge is stored per
  gram of active electrode material.
- It is a primary metric for comparing battery materials.
- The formula is simple: `specific_capacity = capacity_mAh / mass_g`.
- Mistakes happen when mass is entered in milligrams instead of grams,
  inactive material is included, zero is entered by mistake, or negative
  capacity appears from instrument noise.
- A silent wrong answer is worse than a loud error.

"A function with proper input validation catches these mistakes before
they become a wrong number in a table. Let's build it."
:::

# Segment 4 - `stop()` and `warning()` (15 min)

## Stage 1 - naive version

::: callout-tip
## Teaching cue

Build the function in three stages. Start with the naive version so
students can see what goes wrong before you add protection.
:::

```{r}
#| label: naive-specific-capacity

calc_specific_capacity <- function(capacity_mAh, mass_g) {
  return(capacity_mAh / mass_g)
}
```

```{r}
#| label: naive-specific-capacity-zero

calc_specific_capacity(3.0, 0)       # Inf - wrong, no error
calc_specific_capacity(3.0, 0.015)   # works: 200 mAh/g
```

```{r}
#| label: naive-specific-capacity-bad-type
#| error: true

calc_specific_capacity(3.0, "graphite")  # error, but cryptic
```

::: callout-tip
## Teaching cue

"Both examples fail in different ways. The text input fails with an
unhelpful error. The `Inf` case is worse: it passes without complaint
and produces a wrong answer that could make it into a paper."
:::

## Stage 2 - add `stop()` for impossible inputs

::: callout-note
## Former slides 6-7: hard errors

Use `stop()` when the input is so wrong that the function cannot produce
a meaningful result. The function halts and prints your message.

A good `stop()` message tells the user what was wrong and, when
possible, how to fix it.
:::

```{r}
#| label: specific-capacity-with-stop

calc_specific_capacity <- function(capacity_mAh, mass_g) {
  if (!is.numeric(capacity_mAh) | !is.numeric(mass_g)) {
    stop("Both capacity_mAh and mass_g must be numeric.")
  }

  if (mass_g <= 0) {
    stop("mass_g must be a positive value - check your units.")
  }

  return(capacity_mAh / mass_g)
}
```

```{r}
#| label: test-specific-capacity-stop-text
#| error: true

calc_specific_capacity(3.0, "graphite")  # stop: not numeric
```

```{r}
#| label: test-specific-capacity-stop-zero
#| error: true

calc_specific_capacity(3.0, 0)           # stop: mass <= 0
```

```{r}
#| label: test-specific-capacity-stop-good

calc_specific_capacity(3.0, 0.015)       # works: 200 mAh/g
```

## Stage 3 - add `warning()` for suspicious but computable inputs

::: callout-note
## Former slide 8: soft warnings

Use `warning()` when something is suspicious, but the calculation can
still proceed. The user should see a flag, but the function does not
need to halt.

Here, zero mass is impossible for the calculation, so it gets `stop()`.
Negative capacity is unusual, but it may reflect instrument behavior or
a case the researcher needs to inspect, so it gets `warning()`.
:::

```{r}
#| label: specific-capacity-with-warning

calc_specific_capacity <- function(capacity_mAh, mass_g) {
  if (!is.numeric(capacity_mAh) | !is.numeric(mass_g)) {
    stop("Both capacity_mAh and mass_g must be numeric.")
  }

  if (mass_g <= 0) {
    stop("mass_g must be a positive value - check your units.")
  }

  if (capacity_mAh < 0) {
    warning("Negative capacity detected - verify instrument output.")
  }

  return(capacity_mAh / mass_g)
}
```

```{r}
#| label: test-specific-capacity-warning

calc_specific_capacity(-0.5, 0.015)   # warning, but returns -33.3
calc_specific_capacity(3.0, 0.015)    # clean: 200 mAh/g
```

::: callout-note
## `stop()` versus `warning()` decision table

| Situation | Use | Reason |
|----|----|----|
| Input type is wrong | `stop()` | The function cannot compute the intended result. |
| Value is physically impossible | `stop()` | The result would be meaningless. |
| Value is unexpected but computable | `warning()` | Flag it and let the user decide. |
| Value is near the edge of a reasonable range | `warning()` | Worth noting, not fatal. |

Rule of thumb: would the result mislead someone? If yes, stop. If it is
suspicious but still interpretable, warn.
:::

::: callout-tip
## Chat prompt

Ask in chat:

"Why did we use `warning()` for negative capacity but `stop()` for zero
mass? What is the reasoning? 30 seconds."

Expected answer: negative capacity is physically strange but may be real
instrument noise worth investigating. Zero mass makes division
impossible; no result is better than a wrong one.
:::

# Segment 5 - Default arguments (7 min)

::: callout-note
## Former slides 9-10: defaults make functions reusable

"Arguments can have default values, set in the function definition. The
user only provides them when they want something different. This is what
makes functions reusable across research contexts."
:::

```{r}
#| label: specific-capacity-with-defaults

calc_specific_capacity <- function(capacity_mAh,
                                   mass_g,
                                   units = "mAh/g") {
  if (!is.numeric(capacity_mAh) | !is.numeric(mass_g)) {
    stop("Both capacity_mAh and mass_g must be numeric.")
  }

  if (mass_g <= 0) {
    stop("mass_g must be positive.")
  }

  if (capacity_mAh < 0) {
    warning("Negative capacity detected - verify instrument output.")
  }

  result <- capacity_mAh / mass_g
  message("Returning specific capacity in ", units)
  return(result)
}
```

```{r}
#| label: test-specific-capacity-defaults

calc_specific_capacity(3.0, 0.015)                  # uses default
calc_specific_capacity(3.0, 0.015, units = "Ah/g") # override
```

::: callout-tip
## Teaching cue

"Sensible defaults mean your function works out of the box for the most
common case. The override lets it adapt. This is the same pattern you've
been using all semester: `ggplot2` geoms have defaults for color, size,
and alpha. You only specify them when you need something different."

Technical note: in this teaching example, `units` is only a label in the
message. The function does not convert mAh/g to Ah/g.
:::

# Segment 6 - Molecular biology: `calc_primer_quality()` (8 min)

::: callout-note
## Former slides 11-12: primer quality context

"Let's bring everything we've learned together in one function:
conditionals, `warning()`, `stop()`, and default arguments. We'll do it
in molecular biology."

Read these points aloud:

- GC content should often be about 40-60 percent for standard PCR.
- GC below 40 percent: primer may not bind reliably.
- GC above 60 percent: primer may form secondary structures or
  non-specific products.
- These thresholds are checked for every primer, which makes them a good
  candidate for a function with built-in checks.
:::

```{r}
#| label: calc-primer-quality

calc_primer_quality <- function(sequence,
                                gc_min = 40,
                                gc_max = 60) {
  if (!is.character(sequence)) {
    stop("sequence must be a character string.")
  }

  gc <- calc_gc_content(sequence)
  tm <- calc_primer_tm(sequence)

  if (gc < gc_min | gc > gc_max) {
    warning("GC content (", round(gc, 1), "%) is outside ",
            gc_min, "-", gc_max, "% optimal range.")
  }

  return(list(
    gc_percent  = gc,
    tm_celsius  = tm,
    gc_in_range = (gc >= gc_min & gc <= gc_max)
  ))
}
```

```{r}
#| label: test-primer-quality

# Standard PCR - use defaults.
calc_primer_quality("ATGCGCATTAGC")

# High-GC primer under standard PCR defaults - should warn.
calc_primer_quality("GCGCGGCATGCC")

# GC-rich organism or assay - override the acceptable range.
calc_primer_quality("GCGCGATATGCC", gc_min = 50, gc_max = 70)
```

::: callout-tip
## Teaching cue

"Look at what this function does: it validates input, calls two
functions we already wrote, checks the result against thresholds with a
warning, and returns a named list. Every tool from today and Monday is
in here."

Thesis connection: "This is a real primer screening workflow. If you
work in a molecular biology lab, add this pattern to your project
repository. It replaces a manual spreadsheet."

Note on the third example: the sequence was chosen so the 50-70 percent
GC override actually demonstrates the value of changing the defaults.
:::

# Segment 7 - Wrapping `ggplot` in a function (8 min)

::: callout-note
## Former slides 13-15: DRY applies to plots too

"The DRY principle applies equally to visualization code. If you are
making the same style of plot 15 times in your thesis, you should write
it once."
:::

```{r}
#| label: custom-theme

custom_theme <- function(base_size = 9) {
  theme(
    text = element_text(family = "Helvetica",
                        color = "gray30",
                        size = base_size),
    plot.title = element_text(size = rel(1.25),
                              hjust = 0.5,
                              face = "bold"),
    panel.background = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "grey90",
                                    linewidth = 0.25),
    axis.ticks = element_blank()
  )
}
```

```{r}
#| label: plot-cycling

plot_cycling <- function(df,
                         point_size = 2,
                         font_size = 10,
                         ce_threshold = 99) {
  ggplot(df, aes(x = cycle, y = CE)) +
    geom_point(size = point_size, color = "#167B62") +
    geom_line(color = "#167B62", alpha = 0.6) +
    geom_hline(yintercept = ce_threshold,
               linetype = "dashed",
               color = "#A23923") +
    labs(title = "Coulombic Efficiency by Cycle",
         x = "Cycle Number",
         y = "Coulombic Efficiency (%)") +
    custom_theme(font_size)
}
```

```{r}
#| label: call-plot-cycling-default

plot_cycling(cycling_data)
```

```{r}
#| label: call-plot-cycling-custom

plot_cycling(cycling_data, point_size = 3, ce_threshold = 98)
```

::: callout-tip
## Teaching cue

Ask students which parts are fixed and which parts are flexible.

- Fixed: the columns, title, geoms, and overall style.
- Flexible: point size, font size, and CE threshold.

Then say:

"Your advisor asks you to change the threshold line from 99 percent to
98 percent across all 15 cycling plots in your thesis. Without the
function: find and edit 15 plot blocks. With the function: change one
default argument. Done."
:::

# Segment 8 - Synthesis and bridge to activity (5 min)

::: callout-note
## Former slide 16: bridge to Thursday activity

"Thursday's activity walks you through a set of exercises from the NCEAS
textbook - s18 in your course website. The exercises use a function
called `find_max()` that you will build from scratch, break
intentionally, and then harden with the exact same `stop()` and
`warning()` patterns we used today."

Today's lecture maps to the activity this way:

- `double_it()` and `exclaim_age()` review Monday's function anatomy.
- `find_max()` with conditionals reviews today's `if`/`else` pattern.
- Adding `stop()` to `find_max()` follows the same pattern as
  `calc_specific_capacity()`.
- The default argument `equal_ok = FALSE` follows the same pattern as
  `gc_min`, `gc_max`, and `ce_threshold`.

"Everything from this week maps directly onto the activity. Keep this
file open as a reference. The textbook page s18 has the full exercise
set."
:::

```{r}
#| label: bridge-to-activity
#| eval: false

# Function anatomy review.
double_it <- function(x) {
  return(2 * x)
}

exclaim_age <- function(age) {
  return(paste("I am", age, "years old!"))
}

# Conditionals.
find_max <- function(value_1, value_2) {
  if (value_1 > value_2) {
    return(value_1)
  } else if (value_2 > value_1) {
    return(value_2)
  }
}

# Add validation.
find_max <- function(value_1, value_2) {
  if (!is.numeric(value_1) | !is.numeric(value_2)) {
    stop("Values must be numeric.")
  }

  if (value_1 == value_2) {
    stop("Values must be different from each other.")
  }

  if (value_1 > value_2) return(value_1)
  if (value_2 > value_1) return(value_2)
}

# Add a default argument.
find_max <- function(value_1, value_2, equal_ok = FALSE) {
  if (!is.numeric(value_1) | !is.numeric(value_2)) {
    stop("Values must be numeric.")
  }

  if (value_1 == value_2) {
    if (equal_ok) return(value_1)
    stop("Values must be different from each other.")
  }

  if (value_1 > value_2) return(value_1)
  if (value_2 > value_1) return(value_2)
}
```

# Segment 9 - Wrap-up (2 min)

::: callout-note
## Former slide 17: summary

"The complete function toolkit is now in your hands. Write it once.
Protect it with `stop()` and `warning()`. Make it flexible with default
arguments. Apply it everywhere, including inside `ggplot`. That's
reproducible, defensible, reusable science."

Today we added:

- conditionals: `if`, `else if`, and `else`;
- hard errors: `stop()`;
- soft warnings: `warning()`;
- default arguments;
- functions that call other functions;
- plotting functions.

The central judgment is this: would a bad input create a result that
could mislead someone? If yes, stop. If the result is suspicious but
still worth seeing, warn.
:::

```{r}
#| label: final-function-list
#| eval: false

# Today's main functions.
classify_temp()
calc_specific_capacity()
calc_primer_quality()
custom_theme()
plot_cycling()
```

# Exit ticket (1 minute)

::: callout-tip
## Before you leave

Copy this into chat or Canvas and fill in the blanks:

**One tool I can use from today:** `stop()` / `warning()` / default
arguments / ggplot function

**One thing I want to remember:**
\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
:::

::: callout-note
## Teaching cue

Keep this deliberately easy. The goal is not to grade them; it is to
give students a tiny retrieval practice moment and give you a quick read
on what stuck.
:::
