---
title: "Data Visualization"
subtitle: "Week 8 · Lecture 2"
format: 
  revealjs:
    slide-number: true
    code-link: true
    highlight-style: a11y
    chalkboard: true
    theme:
      - lh_slides_style.scss
execute:
  echo: true
  eval: false
  code-line-numbers: false
editor: 
  markdown: 
    wrap: 72
---

## {#title-slide data-menu-title="Title Slide"}

[Data Visualization]{.custom-title}

[Customization, color & interactive outputs · Week 8 · Lecture 2]{.custom-subtitle}

<hr class="hr-red">

[**BIO/CHEM 806 · Gretchen LeBuhn**]{.custom-subtitle2}

---

## {#recap data-menu-title="Recap"}

[Recap: Where We Are]{.slide-title}

<br>

:::{.body-text}
**Lecture 1** gave us the foundation:

- Tufte's principles of graphical excellence and integrity
- The Grammar of Graphics mental model
- Tidy data for plotting — `pivot_longer()` + `summarise()`
- The four essential `ggplot2` components
- Geom types: `geom_col()`, `geom_boxplot()`, `geom_line()`, `geom_point()`

<br>

**Today** we make those plots *great*:  
labels · themes · color · factor ordering · facets · saving · interactive outputs
:::

---

## {#aes-question data-menu-title="aes() Question"}

[Quick Recall]{.slide-title}

<br>

:::{.body-text}
Before we start — what's the difference between:

```{r}
# Version A
ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits,
           fill = "blue")) +
    geom_col()
```

```{r}
# Version B
ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits)) +
    geom_col(fill = "blue")
```

Type your prediction in **chat** before we run either one.
:::

---

## {#salmon-moment data-menu-title="The Salmon Moment"}

[The "Why Is It Salmon?" Moment]{.slide-title}

<br>

:::{.body-text}
Version A produces **salmon-colored** bars with a legend that says "blue."

**Why?**

Putting `fill = "blue"` inside `aes()` tells `ggplot2` that `"blue"` is a *variable in your data*. ggplot2 creates a new column filled with the word "blue" and maps it to fill — using its **default color** for that category.
:::

. . .

:::{.body-text}
**The fix:** move `fill` *outside* of `aes()`:

```{r}
ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits)) +
    geom_col(fill = "blue")
```

This is now a **constant** — all bars are blue.
:::

---

## {#aes-rule data-menu-title="The aes() Rule"}

[The `aes()` Boundary — The Most Important Rule]{.slide-title}

<br>

:::{.body-text}
**Inside `aes()`** → maps a **variable** to an aesthetic  
The value changes per data point based on your data.

```{r}
aes(fill = visitor_type)   # different color per visitor type ✓
```

<br>

**Outside `aes()`** → sets a **constant** for all data points

```{r}
geom_col(fill = "blue")    # all bars are blue ✓
```
:::

. . .

:::{.body-text}
*"If you want to map a variable onto a graph aesthetic, put it within* `aes()`. *If you want to update your plot base with a constant, add it directly to the relevant* `geom_` *layer outside the* `aes()` *call."*

— NCEAS Learning Hub textbook
:::

---

## {#map-variable data-menu-title="Mapping a Variable"}

[Mapping a Variable to Color]{.slide-title}

<br>

:::{.body-text}
`fill = visitor_type` **inside** `aes()` — correct, because it's a real column:
:::

```{r}
ggplot(data = daily_visits_loc,
       aes(x = restore_loc, y = daily_visits,
           fill = visitor_type)) +
    geom_col()
```

:::{.body-text}
`ggplot2` assigns a different color to each visitor type and adds a legend **automatically**.

That's the Grammar of Graphics at work.
:::

---

## {#labs data-menu-title="labs()"}

[Labels with `labs()` + Flip the Axes]{.slide-title}

<br>

```{r}
ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits,
           fill = visitor_type)) +
    geom_col() +
    labs(
        x        = "Number of Visits",
        y        = "Restoration Location",
        fill     = "Type of Visitor",
        title    = "Total Visits to Delta Restoration Areas by Visitor Type",
        subtitle = "Sum of all visits during July 2017 and March 2018"
    ) +
    theme_bw()
```

:::{.body-text}
Swapping `x` and `y` → **horizontal bars** → overlapping labels solved in one line.
:::

---

## {#builtin-themes data-menu-title="Built-in Themes"}

[Built-in Themes]{.slide-title}

<br>

:::{.body-text}
`ggplot2` ships with several complete themes. Try them with auto-complete:  
type `theme_` then **Tab** in RStudio.
:::

```{r}
theme_bw()       # white background, dark grid — textbook standard
theme_minimal()  # no background panel, just grid lines
theme_light()    # light gray grid, white background
theme_classic()  # no grid at all — publication classic
```

. . .


---

## {#custom-theme data-menu-title="custom theme()"}

[Fine-Tuning with `theme()`]{.slide-title}

<br>

:::{.body-text}
`theme()` gives fine-grained control over every visual element.  
⚠️ Must come **after** built-in themes — otherwise `theme_bw()` overrides your settings.
:::

```{r}
ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits,
           fill = visitor_type)) +
    geom_col() +
    labs(x = "Number of Visits", y = "Restoration Location",
         fill = "Type of Visitor",
         title = "Total Visits to Delta Restoration Areas by Visitor Type") +
    theme_bw() +
    theme(
        legend.position = "bottom",
        axis.ticks.y    = element_blank()
    )
```

---

## {#mytheme data-menu-title="my_theme object"}

[Save Your Theme as an Object]{.slide-title}

<br>

:::{.body-text}
When you have a set of theme settings you like — **save them once, use everywhere:**
:::

```{r}
my_theme <- theme_bw(base_size = 16) +
    theme(
        legend.position = "bottom",
        axis.ticks.y    = element_blank()
    )
```

. . .

```{r}
ggplot(data = daily_visits_loc,
       aes(y = restore_loc, x = daily_visits,
           fill = visitor_type)) +
    geom_col() +
    labs(x = "Number of Visits", y = "Restoration Location",
         fill = "Type of Visitor",
         title = "Total Visits to Delta Restoration Areas by Visitor Type",
         subtitle = "Sum of all visits during study period") +
    my_theme
```

:::{.body-text}
`base_size = 16` scales **all text** proportionally — change in one place.
:::

---

## {#exercise-2 data-menu-title="Exercise 2"}

[Exercise 2: `scale_x_continuous()`]{.slide-title}

<br>

:::{.body-text}
**Predict** what this line adds before running it:

```{r}
scale_x_continuous(breaks = seq(0, 120, 20))
```

Then add `expand = c(0, 0)` — what changes?

```{r}
scale_x_continuous(breaks = seq(0, 120, 20), expand = c(0, 0))
```
:::

. . .

:::{.body-text}
- `breaks = seq(0, 120, 20)` → tick marks at 0, 20, 40, 60, 80, 100, 120  
- `expand = c(0, 0)` → removes the gap between bars and the axis line  

These polish details matter for publication figures.
:::

---

## {#fct-reorder data-menu-title="fct_reorder()"}

[Reordering — Defeating Alphabetical Order]{.slide-title}

<br>

:::{.body-text}
`ggplot2` loves alphabetical order. That's almost never the order you want.

**`forcats::fct_reorder()`** reorders a factor by another variable.
:::

```{r}
daily_visits_totals <- daily_visits_loc %>%
    group_by(restore_loc) %>%
    mutate(n = sum(daily_visits)) %>%
    ungroup() %>%
    mutate(restore_loc = fct_reorder(restore_loc, n))

levels(daily_visits_totals$restore_loc)   # no longer alphabetical!
```

. . .

:::{.body-text}
To reverse — most visited at top — wrap `n` in `desc()`:

```{r}
mutate(restore_loc = fct_reorder(restore_loc, desc(n)))
```

**Pattern:** `mutate + fct_reorder` is one of the most frequent pre-plot steps.
:::

---

## {#viridis data-menu-title="viridis colors"}

[Accessible Color with `viridis`]{.slide-title}

<br>

:::{.body-text}
~1 in 12 men and 1 in 200 women have some form of **color vision deficiency.**

If your visualization relies on red-green contrast, a significant portion of your audience can't read it.

The `viridis` package provides palettes that are:

- Perceptually uniform
- Accessible to the most common forms of color blindness
- Print well in grayscale
:::

. . .

```{r}
ggplot(data = daily_visits_totals,
       aes(x = daily_visits, y = restore_loc,
           fill = visitor_type)) +
    geom_col() +
    scale_fill_viridis_d() +   # _d = discrete variable
    scale_x_continuous(breaks = seq(0, 120, 20), expand = c(0, 0)) +
    labs(x = "Number of Visits", y = "Restoration Location",
         fill = "Type of Visitor",
         title = "Total Visits to Delta Restoration Areas by Visitor Type",
         subtitle = "Sum of all visits during study period") +
    my_theme
```

---

## {#ggsave data-menu-title="ggsave()"}

[Saving Plots with `ggsave()`]{.slide-title}

<br>

:::{.body-text}
`ggsave()` saves the last plot you created — or any named plot object:
:::

```{r}
# Save the last plot rendered
ggsave("figures/visit_restore_site_delta.jpg",
       width = 12, height = 6, units = "in")

# Save a named object at publication resolution
ggsave("figures/visit_restore_site_delta.png",
       plot   = my_named_plot,
       width  = 12,
       height = 6,
       units  = "in",
       dpi    = 300)
```

:::{.body-text}
`ggsave()` infers file format from the extension: `.jpg`, `.png`, `.pdf`, `.svg` all work.
:::

---

## {#facet-wrap data-menu-title="facet_wrap()"}

[Small Multiples with `facet_wrap()`]{.slide-title}

<br>

:::{.body-text}
The `~variable` syntax creates **one panel per unique value** of that variable.

This is Tufte's *small multiples* — same chart repeated across subgroups.
:::

```{r}
facet_plot <- ggplot(data = daily_visits_totals,
       aes(x = visitor_type, y = daily_visits,
           fill = visitor_type)) +
    geom_col() +
    facet_wrap(~restore_loc,
               scales = "free_y",
               ncol   = 5,
               nrow   = 2) +
    scale_fill_viridis_d() +
    labs(x = "Type of Visitor", y = "Number of Visits",
         title = "Total Visits to Delta Restoration Areas") +
    theme_bw() +
    theme(legend.position = "bottom",
          axis.ticks.x    = element_blank(),
          axis.text.x     = element_blank())

ggsave("figures/visit_restore_site_facet.jpg",
       plot = facet_plot, width = 12, height = 8, units = "in")
```

---

## {#facet-args data-menu-title="facet_wrap() Arguments"}

[`facet_wrap()` Key Arguments]{.slide-title}

<br>

:::{.body-text}
| Argument | What it does |
|---|---|
| `~variable` | One panel per unique value of `variable` |
| `scales = "free_y"` | Each panel gets its own y-axis scale |
| `scales = "free"` | Both axes free per panel |
| `ncol =` / `nrow =` | Control grid layout |

<br>

`scales = "free_y"` is important when sites have very different total counts — a shared scale would make small-count sites unreadable.

<br>

*When to use faceting:* across species, sites, treatments, time periods, experimental conditions...
:::

---

## {#interactive-intro data-menu-title="Interactive Viz"}

[From Static to Interactive]{.slide-title}

<br>

:::{.body-text}
Two packages for interactive outputs — powerful when published to **GitHub Pages** (Week 11):

<br>

**`DT`** — interactive data tables  
Sortable · searchable · paginated · built on the DataTables JavaScript library

<br>

**`leaflet`** — interactive maps  
Pan · zoom · clickable markers · layerable tile sources

<br>

⚠️ These only render as **interactive HTML** — not in static PDFs or the RStudio viewer.  
Render your Quarto document to HTML to see them.
:::

---

## {#dt-tables data-menu-title="DT Tables"}

[Interactive Tables with `DT`]{.slide-title}

<br>

:::{.body-text}
First, create a data frame of unique sampling locations using `dplyr::distinct()`:
:::

```{r}
locations <- visits_long %>%
    distinct(restore_loc, .keep_all = TRUE) %>%
    select(restore_loc, latitude, longitude)
```

. . .

:::{.body-text}
`.keep_all = TRUE` retains all columns (without it, only `restore_loc` is returned).

Then one function call creates a fully interactive table:
:::

```{r}
datatable(locations)
```

:::{.body-text}
Readers can **sort by any column, search, and paginate** — far more useful than a static table for a monitoring dataset with many sites.
:::

---

## {#leaflet-basic data-menu-title="leaflet Maps"}

[Interactive Maps with `leaflet`]{.slide-title}

<br>

:::{.body-text}
`leaflet` uses **`%>%`** — not `+` like ggplot2. The `~` models columns to the map (similar to `facet_wrap()`).
:::

```{r}
leaflet(locations) %>%
    addTiles() %>%               # OpenStreetMap base layer
    addMarkers(
        lng   = ~longitude,
        lat   = ~latitude,
        popup = ~restore_loc     # clickable label
    )
```

. . .

```{r}
# Circle markers with custom styling
leaflet(locations) %>%
    addTiles() %>%
    addCircleMarkers(
        lng         = ~longitude,
        lat         = ~latitude,
        popup       = ~restore_loc,
        radius      = 5,
        fillColor   = "salmon",
        fillOpacity = 1,
        stroke      = TRUE,
        weight      = 0.5,
        color       = "white",
        opacity     = 1
    )
```

---

## {#leaflet-layers data-menu-title="Layering Base Maps"}

[Layering Tile Sources]{.slide-title}

<br>

:::{.body-text}
You can layer **USGS base maps** using `addWMSTiles()` — the same layering logic as ggplot2 geoms:
:::

```{r}
leaflet(locations) %>%
    addWMSTiles(
        "https://basemap.nationalmap.gov/arcgis/services/USGSImageryTopo/MapServer/WmsServer",
        layers  = "0",
        options = WMSTileOptions(format = "image/png", transparent = TRUE)
    ) %>%
    addWMSTiles(
        "https://basemap.nationalmap.gov/arcgis/services/USGSHydroCached/MapServer/WmsServer",
        layers  = "0",
        options = WMSTileOptions(format = "image/png", transparent = TRUE)
    ) %>%
    addCircleMarkers(
        lng = ~longitude, lat = ~latitude, popup = ~restore_loc,
        radius = 5, fillColor = "salmon", fillOpacity = 1,
        stroke = TRUE, weight = 0.5, color = "white", opacity = 1
    )
```

---

## {#resources data-menu-title="Resources"}

[Resources]{.slide-title}

<br>

:::{.body-text}
**From the textbook — bookmark these:**

[Why not to use two axes](https://blog.datawrapper.de/dualaxis/) — Lisa Charlotte Rost  
*A compelling argument against dual-axis charts*

[Customized Data Visualization in ggplot2](https://allisonhorst.github.io/eds221-day8-comp/data_viz.html) — Allison Horst  
*Extremely detailed, beautiful examples*

[A ggplot2 tutorial for beautiful plotting in R](https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/) — Cédric Scherer  
*An epic reference — bookmark it*

[R Color Cheatsheet](https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf) — NCEAS  
*Discrete vs. continuous palettes, viridis and beyond*

<br>

**And always:** search *"ggplot cheat sheet"* — your best friend during the activity.
:::

---

## {#activity-preview data-menu-title="Activity Preview"}

[Activity Preview]{.slide-title}

<br>

:::{.body-text}
In Nicole's 2-hour section, you'll execute the **full pipeline**:

1. `clean_names()` → `pivot_longer()` → `group_by()` + `summarise()`
2. Build a bar chart — discover the `aes()` boundary firsthand
3. `labs()`, `theme_bw()`, `theme()`, save `my_theme` object
4. `fct_reorder()` to sort locations by total visits
5. `scale_fill_viridis_d()` for accessible color
6. `facet_wrap()` for small multiples
7. `ggsave()` for publication-ready output
8. `datatable()` and `leaflet()` for interactive outputs

<br>

**Textbook:** use Section 3 as your reference throughout — it mirrors the pipeline exactly.
:::
