---
title: "Week 13 Lecture 1: Vector Spatial Data and CRS"
format:
  revealjs:
    slide-number: true
    code-link: true
    highlight-style: a11y
    chalkboard: true
    theme:
      - lh_slides_style.scss
execute:
  echo: true
  eval: false
  code-line-numbers: false
editor:
  markdown:
    wrap: 72
---

##  {#title-slide data-menu-title="Title Slide"}

[Working with Spatial Data]{.custom-title}

[Part 1: Vector Data & Coordinate Reference Systems]{.custom-subtitle}

<hr class="hr-red">

[**BIO/CHEM 806 · Week 13**]{.custom-subtitle2}

------------------------------------------------------------------------

##  {#learning-objectives data-menu-title="Learning Objectives"}

[By the end of today you will be able to...]{.slide-title}

<hr>

::: body-text
- Explain what **vector spatial data** is and how R stores it using the
  `sf` package
- Identify the components of an `sf` object: geometry, attributes, and
  spatial metadata
- Describe what a **Coordinate Reference System (CRS)** is and why it
  matters
- Distinguish between **geographic** and **projected** coordinate
  systems
- Check and transform CRS using `st_crs()` and `st_transform()`
- Convert a CSV with lat/lon columns into an `sf` object using
  `st_as_sf()`

📖 [Textbook
s15](https://nceas-learning-hub.github.io/sfsu_data_sci/s15_r_geospatial_vector_analysis.html)
· sections 1–3.1
:::

------------------------------------------------------------------------

##  {#why-spatial data-menu-title="Why Spatial Data"}

[Spatial data across biology and chemistry]{.slide-title}

<hr>

::: body-text
Almost every field question has a **"where"** — not just ecology:

1.  **Molecular/cell biology:** where in the cell is this protein
    localized? Which nucleus contains this fluorescence signal?
2.  **Microbiology:** where did this pathogen outbreak originate? How
    does transmission spread across a region?
3.  **Marine/organismal biology:** which habitat type does this species
    occurrence fall in?
4.  **Environmental chemistry:** which monitoring station is nearest to
    this contamination point? Which county does each water sample belong
    to?
5.  **Public health / epidemiology:** what is the disease burden by
    census tract or zip code?

Spatial tools let us answer these questions **quantitatively** and
**reproducibly** — in the same R workflow you already know.
:::

------------------------------------------------------------------------

##  {#vector-vs-raster data-menu-title="Vector vs Raster"}

[Two types of spatial data]{.slide-title}

<hr>

::::::: columns
:::: {.column width="50%"}
::: body-text
**Vector data**\
Discrete locations or boundaries with defined geometry

- Sampling sites, field stations → **points**
- Rivers, transects, roads → **lines**
- Watershed boundaries, counties, habitats → **polygons**

Today's focus ↓
:::
::::

:::: {.column width="50%"}
::: body-text
**Raster data**\
A regular grid of cells, each holding a value

- Elevation (DEM), temperature surfaces
- Satellite imagery, land cover
- Concentration gradient surfaces

(Future topic — many `sf` workflows pair with raster data)
:::
::::
:::::::

------------------------------------------------------------------------

##  {#simple-features data-menu-title="Simple Features"}

[The `sf` package: Simple Features in R]{.slide-title}

<hr>

::: body-text
**Simple Features** is an international standard (ISO 19125) for
representing real-world objects by their location and shape.

> Think of it as a shared language for spatial data — just as CSV is a
> shared language for tabular data.

The `sf` package is R's implementation. It:

- introduces a spatial data class that **extends** the data frame
- provides functions for reading and writing spatial files
- enables spatial operations (joins, intersections, distances)

Almost every `sf` function starts with `st_` — for **s**patial and
**t**emporal. `st_join()`, `st_crs()`, `st_transform()` — you're in `sf`
territory whenever you see that prefix.
:::

------------------------------------------------------------------------

##  {#read-sf data-menu-title="Reading Shapefiles"}

[Reading spatial data with `read_sf()`]{.slide-title}

<hr>

::: body-text
We add a `_sf` suffix to signal this is a spatial data frame:
:::

```{r}
library(sf)
library(here)

ak_rgns_sf <- read_sf(here("data/ak_regions_simp.shp"))

# quick look at each attribute column mapped spatially
plot(ak_rgns_sf)
```

::: body-text
`plot()` gives a fast sanity check — one panel per attribute column,
each displayed as a map. Notice the maps may look distorted — that's a
CRS issue we'll fix shortly.

> **What is a shapefile?** A `.shp` file is a common format for storing
> vector spatial data. It always comes with companion files (`.dbf`,
> `.prj`) — keep them together.
:::

------------------------------------------------------------------------

##  {#sf-as-dataframe data-menu-title="sf as Data Frame"}

[An `sf` object is a data frame — plus a geometry column]{.slide-title}

<hr>

```{r}
class(ak_rgns_sf)
# [1] "sf"         "data.frame"

head(ak_rgns_sf)
glimpse(ak_rgns_sf)
```

::: body-text
`sf` objects have two classes: `sf` **and** `data.frame`. All your
familiar tools — `head()`, `glimpse()`, `select()`, `filter()`,
`group_by()` — work here. In addition, an `sf` object carries:

- **Geometry type** — the shape kind: POINT, LINESTRING, MULTIPOLYGON…
- **Bounding box** — the geographic extent of the data *(think: x-min,
  x-max, y-min, y-max for the dataset)*
- **CRS** — the coordinate reference system *(which map system are we
  using? — more on this next)*
- A **`geometry`** column containing the actual shapes

The `geometry` column is **sticky** — it travels with the object even
when you `select()` other columns.
:::

------------------------------------------------------------------------

##  {#crs-intro data-menu-title="CRS Introduction"}

[What is a Coordinate Reference System?]{.slide-title}

<hr>

::: body-text
A CRS answers: **"How do we translate locations on a 3D Earth onto a
flat screen?"**

It has two components:

**1. Datum** — a model of Earth's shape (it's not a perfect sphere!)

> *"Is the Earth shaped like a lemon, a lime, or an orange?"*\
> Different datums model the bulges and flattening differently. Using
> the wrong datum can shift your points by meters or kilometers.

**2. Projection** — the math that flattens the curved surface into 2D

> *"How do you peel the fruit and flatten the peel?"*\
> Every projection distorts *something* — area, shape, distance, or
> direction. There is no perfect flat map.
:::

------------------------------------------------------------------------

##  {#geographic-vs-projected data-menu-title="Geographic vs Projected"}

[Geographic vs. Projected CRS]{.slide-title}

<hr>

::: body-text
**Geographic (unprojected) CRS**\
Coordinates are in *degrees* of latitude and longitude. No flattening
has happened — you're still on a sphere.

- WGS84 (EPSG **4326**) — used by GPS, the most common global standard;
  this is what your phone records

**Projected CRS**\
Coordinates are in *meters* on a flat plane. The Earth has been
mathematically flattened. Optimized for specific regions.

- Alaska Albers (EPSG **3338**) — minimizes distortion across Alaska
- Pseudo-Mercator (EPSG **3857**) — Google Maps, OpenStreetMap

> **Key practical rule:** You cannot calculate meaningful areas or
> distances in a geographic CRS (degrees aren't meters). Use a
> *projected* CRS for any area or distance calculation.
:::

------------------------------------------------------------------------

##  {#epsg-codes data-menu-title="EPSG Codes"}

[EPSG codes: a shorthand for CRS]{.slide-title}

<hr>

::: body-text
Every standard CRS has a numeric **EPSG code** — an agreed-upon
shorthand so you don't have to type a long projection definition string.
Three you'll use in this unit:

| EPSG     | Name            | Use case                                       |
|-----------------|-----------------|---------------------------------------|
| **4326** | WGS84           | GPS coordinates, lat/lon CSVs, `leaflet` input |
| **3338** | Alaska Albers   | Regional mapping in Alaska (equal-area)        |
| **3857** | Pseudo-Mercator | Web tile basemaps (OSM, ggspatial)             |

> Need a code you don't know? Search [epsg.io](https://epsg.io) by name
> — e.g., "California Albers", "UTM Zone 10N", "NAD83". It returns the
> EPSG number, the proj4 string, and a preview map.
:::

------------------------------------------------------------------------

##  {#st-crs data-menu-title="Checking CRS"}

[Checking CRS with `st_crs()`]{.slide-title}

<hr>

::: body-text
Make checking the CRS a reflex — before doing anything spatial:
:::

```{r}
st_crs(ak_rgns_sf)
```

::: body-text
The output looks dense, but look for the EPSG code in the `ID` field.
Here it will show **WGS84 (EPSG 4326)** — a geographic (unprojected) CRS
with coordinates in degrees.

**The most important rule in spatial data:**

> Before any operation involving two spatial datasets, they **must share
> the same CRS**. If they don't, R will throw an error — and that error
> is protecting you from a silent, wrong answer.

See [spatialreference.org](http://spatialreference.org/ref/epsg/) for a
searchable catalog of all EPSG codes.
:::

------------------------------------------------------------------------

##  {#st-transform data-menu-title="Transforming CRS"}

[Transforming CRS with `st_transform()`]{.slide-title}

<hr>

::: body-text
The Alaska Albers projection (EPSG 3338) is a much better choice for
mapping Alaska — it preserves area and avoids dateline distortion:
:::

```{r}
ak_rgns_3338_sf <- ak_rgns_sf %>%
    st_transform(crs = 3338)

st_crs(ak_rgns_3338_sf)

plot(ak_rgns_3338_sf)
```

::: body-text
💡 **Naming convention:** encode the CRS in the object name (`_3338`,
`_4326`). When you have multiple versions of the same data in different
coordinate systems, the suffix tells you — and anyone reading your code
— exactly what each object is in.
:::

------------------------------------------------------------------------

##  {#sf-tidyverse data-menu-title="sf + Tidyverse"}

[`sf` plays nicely with the tidyverse]{.slide-title}

<hr>

::: body-text
Because `sf` objects are data frames, your existing skills transfer
directly:
:::

```{r}
# select() — geometry stays automatically
ak_rgns_3338_sf %>%
    select(region)

# filter() works exactly as expected
# use %in% when testing multiple values
unique(ak_rgns_3338_sf$region)

ak_rgns_3338_sf %>%
    filter(region == "Southeast")
```

::: body-text
The sticky geometry column means you never have to call it explicitly —
you can't accidentally lose your spatial information with a `select()`.
:::

------------------------------------------------------------------------

##  {#csv-to-sf data-menu-title="CSV to sf"}

[Turning a CSV with lat/lon into an `sf` object]{.slide-title}

<hr>

::: body-text
Most scientific data doesn't arrive as a shapefile — it arrives as a
CSV. If it has latitude and longitude columns, `st_as_sf()` promotes it
to spatial:
:::

```{r}
# _df suffix: just a data frame for now
pop_df <- read_csv(here("data/alaska_population.csv"))

# _sf suffix: now a spatial data frame
pop_4326_sf <- st_as_sf(pop_df,
                        coords = c('lng', 'lat'),
                        crs = 4326,
                        remove = F)

head(pop_4326_sf)
```

::: body-text
`remove = F` keeps the original `lat`/`lng` columns — you'll need them
later for `leaflet`. `coords` must be in **longitude first, latitude
second** order (x before y).
:::

------------------------------------------------------------------------

##  {#crs-mismatch data-menu-title="CRS Mismatch Error"}

[The most common spatial error — and its fix]{.slide-title}

<hr>

::: body-text
Trying to join `pop_4326_sf` (WGS84, degrees) with `ak_rgns_3338_sf`
(Alaska Albers, meters):

``` r
pop_joined_sf <- st_join(pop_4326_sf, ak_rgns_3338_sf, join = st_within)
# Error: st_crs(x) == st_crs(y) is not TRUE
```

**R is protecting you.** Joining datasets in different coordinate
systems is like overlaying two maps printed at different scales — the
result would be nonsense. The fix is always `st_transform()`:
:::

```{r}
pop_3338_sf <- st_transform(pop_4326_sf, crs = 3338)

# Now both are in 3338 — safe to join
pop_joined_sf <- st_join(pop_3338_sf, ak_rgns_3338_sf, join = st_within)
```

::: body-text
Make `st_crs()` checks a habit before any spatial join.
:::

------------------------------------------------------------------------

##  {#decision-log-spatial data-menu-title="Decision Log"}

[Spatial choices belong in your decision log]{.slide-title}

<hr>

::: body-text
Spatial analysis is full of decisions that affect your results —
document them just as you would in a lab notebook:

- Which CRS did you choose for analysis, and why?
- What join type did you use — `st_within` vs. `st_intersects`?
- Did you simplify geometries? At what tolerance?
- Did you clip to a study area boundary?

Example entry:

> *"Transformed to EPSG 3338 (Alaska Albers) because the study area
> spans Alaska and an equal-area projection is needed for valid area
> calculations and comparisons across regions."*

These are methods statements — they belong both in your decision log and
in a methods section of any paper or report.
:::

------------------------------------------------------------------------

##  {#exit-ticket data-menu-title="Exit Ticket"}

[Chat Exit Ticket — write in chat]{.slide-title}

<hr>

::: body-text
**In pseudocode, describe these two steps:**

1.  You have a CSV with `latitude` and `longitude` columns. Describe how
    to read it in and convert it to an `sf` object in WGS84 (EPSG 4326).

2.  You need to do a spatial join with a polygon boundary file that is
    in Alaska Albers (EPSG 3338). What do you check first, and what do
    you do before joining?

No R syntax required — describe the logic clearly enough that a
collaborator could follow your steps.
:::

------------------------------------------------------------------------

##  {#looking-ahead data-menu-title="Looking Ahead"}

[Coming up]{.slide-title}

<hr>

::: body-text
**Wednesday Lecture:**\
Spatial joins with `st_join()`, aggregating with `st_drop_geometry()`,
static maps with `geom_sf()`, basemaps with `ggspatial`, and interactive
maps with `leaflet`

**Activity:**\
Working with Alaska population and region data — the textbook s15
exercise

📖 [Textbook
s15](https://nceas-learning-hub.github.io/sfsu_data_sci/s15_r_geospatial_vector_analysis.html)
— sections 1 through 3.1 as your reference
:::
