---
title: "Week 13 Lecture 2: Spatial Joins, Maps & Mapping Choices"
format:
  revealjs:
    slide-number: true
    code-link: true
    highlight-style: a11y
    chalkboard: true
    theme:
      - lh_slides_style.scss
execute:
  echo: true
  eval: false
  code-line-numbers: false
editor:
  markdown:
    wrap: 72
---

## {#title-slide data-menu-title="Title Slide"}

[Working with Spatial Data]{.custom-title}

[Part 2: Spatial Joins, Maps & Mapping Choices]{.custom-subtitle}

<hr class="hr-red">

[**BIO/CHEM 806 · Week 13**]{.custom-subtitle2}

---

## {#recap data-menu-title="Recap"}

[Quick recap from Monday]{.slide-title}

<hr>

:::{.body-text}
- `sf` objects are data frames with a sticky `geometry` column
- Every `sf` object has a **CRS** — always check with `st_crs()` before
  doing anything spatial
- CRS must match before any spatial operation — transform with
  `st_transform()`
- CSV with lat/lon → `sf` object with `st_as_sf()`
- Naming convention: `_sf` suffix for class, `_4326`/`_3338` for CRS

**Today:** spatial joins + three mapping approaches

📖 [Textbook s15](https://nceas-learning-hub.github.io/sfsu_data_sci/s15_r_geospatial_vector_analysis.html) · sections 3–8
:::

---

## {#learning-obj data-menu-title="Learning Objectives"}

[By the end of today you will be able to...]{.slide-title}

<hr>

:::{.body-text}
- Perform a **spatial join** with `st_join()` to relate points to
  polygons
- Use `st_drop_geometry()` to aggregate spatially joined data as a
  plain data frame
- Combine geometries with `group_by()` + `summarize()` on `sf` objects
- Save spatial data with `write_sf()`
- Build a **static map** with `ggplot2` + `geom_sf()`
- Add a **basemap** with `ggspatial::annotation_map_tile()`
- Build an **interactive map** with `leaflet`
:::

---

## {#spatial-join-concept data-menu-title="Spatial Join Concept"}

[What is a spatial join?]{.slide-title}

<hr>

:::{.body-text}
A **spatial join** uses *geometry* — not a shared key column — to match
rows from two datasets.

> **In other words:** instead of `left_join(df1, df2, by = "id")`, you
> join by asking: *"Does this point fall inside that polygon?"*

The question we'll answer with the Alaska data:

> **"How many people live in each Alaska region?"**

We have population data by *city* point, not by *region* polygon. The
spatial join assigns each city to its region based on geography — then
we can aggregate.
:::

---

## {#spatial-join-examples data-menu-title="Spatial Join Examples"}

[Spatial joins across biology and chemistry]{.slide-title}

<hr>

:::{.body-text}
The same logic applies across disciplines — using location to create a
match that no shared key column can provide:

| Question | Points | Polygons |
|----------|--------|----------|
| Which county does each water sampling station fall in? | monitoring stations | county boundaries |
| Which habitat type does each bird observation belong to? | occurrence records | land cover polygons |
| Which watershed drains to each stream chemistry site? | sample sites | watershed boundaries |
| Which census tract does each disease case belong to? | case addresses | census tracts |
| Which marine protected area contains each dive survey? | survey sites | MPA boundaries |

Think of your own data: do you have points and boundaries you'd want to
connect?
:::

---

## {#st-join data-menu-title="st_join()"}

[Step 1: `st_join()` — the spatial left join]{.slide-title}

<hr>

:::{.body-text}
First, transform population points to match the regions CRS:
:::

```{r}
pop_3338_sf <- st_transform(pop_4326_sf, crs = 3338)
```

:::{.body-text}
Now join — each city row gets the attributes of the region it falls
within:
:::

```{r}
pop_joined_sf <- st_join(pop_3338_sf,
                         ak_rgns_3338_sf,
                         join = st_within)

head(pop_joined_sf)
```

:::{.body-text}
The `join =` argument specifies the geometric relationship.
`st_within` asks: *is this point completely inside that polygon?*
:::

---

## {#join-types data-menu-title="Join Types"}

[Choosing a join type]{.slide-title}

<hr>

:::{.body-text}
The `join =` argument accepts any **spatial predicate** — a function
that tests a geometric relationship between two objects:

| Function | Asks... | Example use |
|----------|---------|-------------|
| `st_within` | Is geometry A completely inside B? | Points in polygons |
| `st_intersects` | Do geometries share any space at all? | Overlapping polygons |
| `st_crosses` | Does A cross through B? | Lines crossing polygons |
| `st_is_within_distance` | Is A within a buffer distance of B? | Proximity analysis |

The right choice depends on your **geometry types** and your **scientific
question**. This is a methods decision — document it.

Run `?st_within` to see the full list.
:::

---

## {#drop-geometry data-menu-title="Drop Geometry"}

[Step 2: `st_drop_geometry()` before aggregating]{.slide-title}

<hr>

:::{.body-text}
We want to `summarize()` across cities — but if we run `group_by()` on
an `sf` object, R tries to aggregate the geometries too (combining all
the city points into one MULTIPOINT per region). That's not what we
want. Drop the geometry first:
:::

```{r}
pop_rgn_df <- pop_joined_sf %>%
    st_drop_geometry() %>%       # back to a plain data frame
    group_by(region) %>%
    summarize(total_pop = sum(population))

head(pop_rgn_df)
```

:::{.body-text}
Now `pop_rgn_df` is a plain data frame — `_df` suffix, not `_sf` —
with one row per region and a total population column. No geometry. We
rejoin it to the regions polygons next.
:::

---

## {#left-join-back data-menu-title="Rejoin Geometry"}

[Step 3: `left_join()` to reattach polygon geometry]{.slide-title}

<hr>

```{r}
pop_rgn_3338_sf <- left_join(ak_rgns_3338_sf,
                             pop_rgn_df,
                             by = "region")

# quick check
plot(pop_rgn_3338_sf["total_pop"])
```

:::{.body-text}
The three-step pattern — you'll use it constantly:

1. `st_join()` → points get polygon attributes
2. `st_drop_geometry()` + `group_by()` + `summarize()` → plain data
   frame with aggregated values
3. `left_join()` back to polygons → `sf` object ready to map
:::

---

## {#summarize-geometries data-menu-title="Summarize Geometries"}

[Bonus: `group_by()` + `summarize()` can also dissolve geometries]{.slide-title}

<hr>

:::{.body-text}
You can also `summarize()` *directly on an `sf` object* to combine
(dissolve) geometries within each group:
:::

```{r}
# Dissolve regions into management areas (combines polygons)
pop_mgmt_3338_sf <- pop_rgn_3338_sf %>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop))

plot(pop_mgmt_3338_sf["total_pop"])
```

:::{.body-text}
The region polygons are merged into one larger polygon per management
area — this is called a **union** or **dissolve**. To keep the original
polygon boundaries (group by label only, don't merge shapes), add
`do_union = F`:
:::

```{r}
pop_mgmt_3338_sf <- pop_rgn_3338_sf %>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop), do_union = F)
```

---

## {#write-sf data-menu-title="Saving Spatial Data"}

[Step 4: Save with `write_sf()`]{.slide-title}

<hr>

:::{.body-text}
Save your spatial object so you don't have to rebuild it every session:
:::

```{r}
write_sf(pop_rgn_3338_sf, here("data/ak_regions_population.shp"))
```

:::{.body-text}
`.shp` writes an **ESRI Shapefile** — the most widely compatible format,
readable in QGIS, ArcGIS, and by collaborators. It creates several
companion files; keep them together.

A more modern alternative is **GeoPackage** (`.gpkg`) — a single file,
no column-name length limit, handles larger datasets:

```r
write_sf(pop_rgn_3338_sf, here("data/ak_regions_population.gpkg"))
```

Read either back with `read_sf()`. This save step belongs in your
numbered script sequence — the same reproducibility practice we've
used all semester.
:::

---

## {#three-mapping-approaches data-menu-title="Three Mapping Approaches"}

[Three tools for mapping spatial data]{.slide-title}

<hr>

:::{.body-text}
| Tool | Output | Best for |
|------|--------|----------|
| `ggplot2` + `geom_sf()` | Static image | Publication figures, full control |
| `ggspatial` + `annotation_map_tile()` | Static + basemap | Static maps needing geographic context |
| `leaflet` | Interactive (HTML) | Exploration, reports, web communication |

The right choice is a **communication decision** — not just what's
easiest to code. Who is your audience? What do they need to *do* with
the map? A hiring manager exploring your portfolio map needs something
different than a journal reviewer looking at a methods figure.

Document your choice in your decision log.
:::

---

## {#ggplot-geom-sf data-menu-title="ggplot + geom_sf"}

[Static maps: `ggplot2` + `geom_sf()`]{.slide-title}

<hr>

:::{.body-text}
`geom_sf()` reads the geometry column automatically — you don't specify
`x` and `y` aesthetics. Everything else works like any other ggplot
layer:
:::

```{r}
ggplot(pop_rgn_3338_sf) +
    geom_sf(aes(fill = total_pop)) +
    labs(fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high = "firebrick",
                          labels = comma) +
    theme_bw()
```

---

## {#layering-geoms data-menu-title="Layering geoms"}

[Layering multiple `geom_sf()` calls]{.slide-title}

<hr>

```{r}
ggplot() +
    geom_sf(data = pop_rgn_3338_sf,       # polygons: fill = total_pop
            aes(fill = total_pop)) +
    geom_sf(data = pop_3338_sf,            # points: city locations
            size = 0.5) +
    geom_sf(data = rivers_3338_sf,         # lines: rivers
            aes(linewidth = StrOrder)) +
    scale_linewidth(range = c(0.05, 0.5),
                    guide = "none") +
    labs(title = "Total Population by Alaska Region",
         fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high = "firebrick",
                          labels = comma) +
    theme_bw()
```

:::{.body-text}
**Layer order matters** — bottom to top. Polygons render first
(background), then lines, then points (foreground). Note that
`rivers_3338_sf` may not have an explicit EPSG set in its metadata —
check [epsg.io](https://epsg.io) with the proj4 string to confirm it
matches.
:::

---

## {#ggspatial data-menu-title="Basemaps with ggspatial"}

[Adding basemaps with `ggspatial`]{.slide-title}

<hr>

:::{.body-text}
Web tile basemaps (OpenStreetMap, etc.) use **Pseudo-Mercator (EPSG
3857)** — the projection behind every major web mapping service.
Transform your data to match before adding tiles:
:::

```{r}
pop_3857_sf <- st_transform(pop_3338_sf, crs = 3857)

ggplot(data = pop_3857_sf) +
    ggspatial::annotation_map_tile(type = "osm",
                                   zoom = 4,
                                   progress = "none") +
    geom_sf(aes(color = population), fill = NA) +
    scale_color_continuous(low = "darkkhaki",
                           high = "firebrick",
                           labels = comma)
```

:::{.body-text}
`annotation_map_tile()` **must go first** so it renders behind your
data. Higher `zoom` = more detail, slower to load (tiles download over
the network).
:::

---

## {#when-basemaps data-menu-title="When to Use Basemaps"}

[When does a basemap help — and when does it clutter?]{.slide-title}

<hr>

:::{.body-text}
**Use a basemap when:**

- Your audience doesn't know the geography and roads, coastlines, or
  city locations help them orient
- Point locations relative to physical features matter for
  interpretation (e.g., a contamination site relative to a river)

**Skip the basemap when:**

- Your boundaries are well-known (states, national parks, standard
  regions)
- The basemap visually competes with your data
- You're making a publication figure for print — `theme_bw()` is
  cleaner

This is a **design decision** — document it in your decision log. If
tile loading times out on your network, note that too.
:::

---

## {#leaflet-intro data-menu-title="Leaflet Introduction"}

[Interactive maps with `leaflet`]{.slide-title}

<hr>

:::{.body-text}
`leaflet` creates **browser-rendered interactive maps** — zoom, click,
hover, popup. Key differences from `ggplot`:

- Uses `%>%` pipes, not `+`
- Renders in the browser (or Viewer pane), not the Plots pane
- Expects your data in **WGS84 (EPSG 4326)**; it handles the visual
  projection client-side

For the Alaska Albers projection, we define a custom CRS object so
leaflet renders it with the correct regional projection:
:::

```{r}
epsg3338 <- leaflet::leafletCRS(
    crsClass = "L.Proj.CRS",
    code = "EPSG:3338",
    proj4def = "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154
                +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0
                +units=m +no_defs",
    resolutions = 2 ^ (16:7)
)

pop_rgn_4326_sf <- pop_rgn_3338_sf %>%
    st_transform(crs = 4326)
```

---

## {#leaflet-polygons data-menu-title="Leaflet Polygons"}

[Building up a `leaflet` map: polygons + legend]{.slide-title}

<hr>

```{r}
pal <- colorNumeric(palette = "Reds",
                    domain = pop_rgn_4326_sf$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_rgn_4326_sf,
        fillColor = ~pal(total_pop),   # ~ is leaflet's version of aes()
        weight = 1,
        color = "black",
        fillOpacity = 1,
        label = ~region
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_rgn_4326_sf$total_pop),
        title = "Total Population"
    )

m
```

:::{.body-text}
Build leaflet maps in stages — start with gray polygons to confirm they
render, then add color and labels.
:::

---

## {#leaflet-markers data-menu-title="Leaflet Circle Markers"}

[Adding point layers to `leaflet`]{.slide-title}

<hr>

```{r}
m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(data = pop_rgn_4326_sf,
                fillColor = ~pal(total_pop),
                weight = 1, color = "black", fillOpacity = 1) %>%
    addCircleMarkers(
        data = pop_4326_sf,
        lat = ~lat,
        lng = ~lng,
        radius = ~log(population / 500),   # log scale: prevents huge outliers
        fillColor = "gray",
        fillOpacity = 1,
        weight = 0.25,
        color = "black",
        label = ~paste0(city, ", population ", comma(population))
    ) %>%
    addLegend(position = "bottomleft", pal = pal,
              values = range(pop_rgn_4326_sf$total_pop),
              title = "Total Population")
m
```

---

## {#mapping-choices data-menu-title="Mapping Choices"}

[Mapping is communication — make intentional choices]{.slide-title}

<hr>

:::{.body-text}
Every map encodes decisions that affect what your audience sees and
concludes:

- **Color scale** — sequential vs. diverging? Perceptually uniform?
  Color-blind accessible?
- **Point size** — proportional to value (`log(population/500)`) or
  uniform?
- **Projection** — does it distort the area or shape you're trying to
  convey?
- **Basemap** — geographic context, or visual noise?
- **Static vs. interactive** — who's your audience, and what do they
  need to *do* with the map?

There is no single correct map. There are only more or less defensible
choices — and your **decision log** is where you defend them.
:::

---

## {#exit-ticket data-menu-title="Exit Ticket"}

[Exit Ticket — write in chat]{.slide-title}

<hr>

:::{.body-text}
In pseudocode, describe a complete spatial analysis pipeline for your
own field or domain:

1. Read in a CSV of sampling locations (lat/lon) → convert to `sf`
2. Transform to match a polygon boundary file's CRS
3. Spatial join: assign each sample point to the polygon it falls within
4. Summarize: count or sum a value per polygon
5. Re-attach results to polygon geometry

Then: which of the three mapping approaches would you use to present
these results — and why? Write one sentence justifying your choice.
:::

---

## {#looking-ahead data-menu-title="Looking Ahead"}

[Activity time]{.slide-title}

<hr>

:::{.body-text}
**Monday cohort:** Complete the activity as homework before Thursday

**Thursday cohort:** Work through this in-person with Nicole

**What you'll need:**

- Alaska KNB data (download code in the activity) — shapefiles +
  population CSV
- Packages: `sf`, `ggplot2`, `leaflet`, `ggspatial`, `dplyr`, `scales`,
  `here`

📖 [Textbook s15](https://nceas-learning-hub.github.io/sfsu_data_sci/s15_r_geospatial_vector_analysis.html) — your reference throughout the activity
:::
