---
title: "AI for Coding Support"
subtitle: "BIO/CHEM 806 — Live-Coding Companion"
author: "Gretchen LeBuhn"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: false
    theme: flatly
    highlight-style: github
execute:
  echo: true
  eval: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

::: {.callout-note icon="false"}
## 📌 How to use this document

This is your live-coding companion. Type (or accept Copilot suggestions)
directly into the code chunks below — don't paste in the answers ahead
of time. The callout boxes link each exercise back to the lecture
slides. Set `eval: true` on any chunk you want to run and see output.
:::

------------------------------------------------------------------------

# Part 1 — GitHub Copilot {#copilot}

> **Concept:** Copilot is a real-time, in-editor pair programmer. It
> reads everything in your document — prose, comments, previous code —
> and predicts the next line.

::: {.callout-tip icon="false"}
## 🔧 Setup check (do this before the session)

``` r
# Run this in your R console to confirm the package is installed
install.packages("palmerpenguins")   # if needed
library(palmerpenguins)
head(penguins)
```
:::

------------------------------------------------------------------------

## Exercise 1 — tidyverse penguin analysis {#ex1}

::: {.callout-tip icon="false"}
## 🎯 Instructor cue — Slide: "Using Copilot to assist with coding tasks"

Write the pseudocode below as Markdown prose *first*, then open the code
chunk and let Copilot suggest each line. Accept with **Tab**. Point out
how Copilot picks up package and function names from the prose you wrote
above the chunk.
:::

Load the `palmerpenguins` package and the `penguins` dataset. Drop any
rows with missing values. Filter to keep only Chinstrap and Adelie
species. Create a scatter plot of bill length vs. bill depth, colored by
species. Then run a Welch two-sample t-test comparing bill length
between the two species and print the results.

```{r}
#| label: ex1-copilot-tidyverse

#load the palmerpenguins package and load the penguins data
#drop any incomplete observations (any NA values)
#filter to just keep the Chinstrap and Adelie species
#generate a scatter plot of bill length vs. bill depth for each species then, for the two species, 
# run a t-test to determine whether the bill lengths are statistically different between the two species. 
#Print a summary of the t test results.





```

::: {.callout-note collapse="true" icon="false"}
## ✅ Reference answer (expand only if needed)

``` r
library(tidyverse)
library(palmerpenguins)

penguins_filtered <- penguins |>
  drop_na() |>
  filter(species %in% c("Chinstrap", "Adelie"))

ggplot(penguins_filtered, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "Bill Length vs. Bill Depth",
       x = "Bill Length (mm)", y = "Bill Depth (mm)") +
  theme_minimal()

t_test_result <- t.test(bill_length_mm ~ species, data = penguins_filtered)
print(t_test_result)
broom::tidy(t_test_result)
```
:::

------------------------------------------------------------------------

## Exercise 2 — Base R version {#ex2}

::: {.callout-tip icon="false"}
## 🎯 Instructor cue — Slide: "Exercise 2"

Below the previous chunk, add a sentence saying "Repeat using only base
R functions and plotting." Then open a new chunk and let Copilot figure
out the style shift from context.
:::

Repeat the previous analysis using only base R functions and plotting.
No tidyverse packages.

```{r}
#| label: ex2-copilot-baseR

# use only base R — load palmerpenguins


# drop NA rows


# filter to Chinstrap and Adelie


# base R scatter plot with legend


# t-test

```

::: {.callout-note collapse="true" icon="false"}
## ✅ Reference answer

``` r
library(palmerpenguins)

penguins_clean <- na.omit(penguins)
penguins_clean <- penguins_clean[penguins_clean$species %in% c("Chinstrap", "Adelie"), ]

plot(penguins_clean$bill_length_mm, penguins_clean$bill_depth_mm,
     col = ifelse(penguins_clean$species == "Chinstrap", "steelblue", "tomato"),
     pch = 19, xlab = "Bill Length (mm)", ylab = "Bill Depth (mm)",
     main = "Bill Length vs. Bill Depth")
legend("topright", legend = c("Chinstrap", "Adelie"),
       col = c("steelblue", "tomato"), pch = 19)

t.test(bill_length_mm ~ species, data = penguins_clean)
```
:::

------------------------------------------------------------------------

## Exercise 3 — Writing a function with Copilot {#ex3}

::: {.callout-tip icon="false"}
## 🎯 Instructor cue — Slide: "Copilot to write a function"

Write the pseudocode as prose, open a chunk, write a comment naming the
function, and let Copilot build it line by line. After it's done,
discuss: Does it pull the right p-value (slope, not intercept)? Would
you structure the output differently?
:::

Write a function called `calc_lm()` that:

- Takes a dataframe `data` and column name strings `y` (outcome) and `x`
  (predictor)
- Fits a linear regression `y ~ x`
- Returns the slope, intercept, and p-value of the slope in a dataframe

```{r}
#| label: ex3-calc-lm-function

# function to calculate slope, intercept, and p-value for a linear regression


# test on the penguins data

```

::: {.callout-note collapse="true" icon="false"}
## ✅ Reference answer

``` r
calc_lm <- function(data, y, x) {
  y_vals <- data[[y]]
  x_vals <- data[[x]]
  lm_model <- lm(y_vals ~ x_vals)
  data.frame(
    slope     = coef(lm_model)[2],
    intercept = coef(lm_model)[1],
    p_value   = summary(lm_model)$coefficients[2, 4]
  )
}

# test
calc_lm(penguins_filtered, "bill_length_mm", "bill_depth_mm")

# alternative: broom is cleaner
# broom::tidy(lm(bill_length_mm ~ bill_depth_mm, data = penguins_filtered))
```
:::

------------------------------------------------------------------------

## Exercise 4 — Nested loop with Copilot {#ex4}

::: {.callout-tip icon="false"}
## 🎯 Instructor cue — Slide: "Copilot to write a loop"

This one Copilot will struggle with. Let it try, then help it along —
this is intentional. Point out the specific bugs as you encounter them
(single vs. double brackets, missing index increment, incorrect filter
syntax). Emphasize: for complex logic, ChatGPT (Part 2) is often the
better tool.
:::

For each combination of species and sex in `penguins_filtered`, apply
`calc_lm()` to model `bill_length_mm ~ bill_depth_mm`. Store results in
a list, then combine into a single dataframe.

```{r}
#| label: ex4-nested-loop

# create empty list to store results


# iterate over species
  # iterate over sex
    # apply calc_lm() to the filtered subset
    # add species and sex to the result
    # store in results list


# combine all results into a single dataframe

```

::: {.callout-note collapse="true" icon="false"}
## ✅ Reference answer (manual version — Copilot-assisted)

``` r
results <- list()
i <- 1

for (sp in unique(penguins_filtered$species)) {
  for (sx in unique(penguins_filtered$sex)) {
    subset_df <- penguins_filtered[
      penguins_filtered$species == sp & penguins_filtered$sex == sx, ]
    result <- calc_lm(subset_df, "bill_length_mm", "bill_depth_mm")
    result$species <- sp
    result$sex <- sx
    results[[i]] <- result
    i <- i + 1
  }
}

results_df <- do.call(rbind, results)
results_df
```
:::

------------------------------------------------------------------------

# Part 2 — Iterative Coding with ChatGPT {#chatgpt}

> **Concept:** ChatGPT (or Claude) is a conversational coding
> collaborator. You write prompts, it returns complete code blocks, you
> test them, then iterate through follow-up messages. Unlike Copilot, it
> can revise and rewrite earlier code.

------------------------------------------------------------------------

## Exercise 5 — Pseudocode → code {#ex5}

::: {.callout-tip icon="false"}
## 🎯 Instructor cue — Slide: "Pseudocode into code"

Switch to the browser (chat.openai.com or claude.ai). Paste the prompt
below into a NEW chat. Copy the response back here and run it. Compare
to Exercise 1.
:::

**Prompt to paste into ChatGPT:**

```         
Here is a sequence of steps I want to carry out in R using a tidyverse workflow.
Please write the R code:

- Load the palmerpenguins package and the penguins dataset
- Drop any rows with NA values
- Filter to keep only Chinstrap and Adelie species
- Create a scatter plot of bill length vs. bill depth, colored by species
- Run a Welch t-test comparing bill length between the two species and print the results
```

```{r}
#| label: ex5-chatgpt-code

# Paste the code ChatGPT returns here and run it


```

------------------------------------------------------------------------

## Exercise 6 — Follow-up: no tidyverse {#ex6}

::: {.callout-tip icon="false"}
## 🎯 Instructor cue — Slide: "Exercise 6"

In the SAME ChatGPT chat (do NOT start a new one), send the follow-up
below. Point out that ChatGPT remembers the previous context — it knows
the dataset, species, and variables without being told again.
:::

**Follow-up message in the same chat:**

```         
Now rewrite that code using only base R functions — no tidyverse packages.
```

```{r}
#| label: ex6-chatgpt-baseR

# Paste ChatGPT's base R response here

```

------------------------------------------------------------------------

## Exercise 7 — Commenting unfamiliar code {#ex7}

::: {.callout-tip icon="false"}
## 🎯 Instructor cue — Slide: "Commenting existing code"

Open a NEW chat. Navigate to the URL below and copy the raw Python
script. Paste it into ChatGPT with the prompt below. Scroll through the
commented version together — discuss what ChatGPT added and whether any
code was changed.

URL:
https://raw.githubusercontent.com/convei-wwf/NLParty/refs/heads/main/abstract_classifier.py
:::

**Prompt to paste into a NEW ChatGPT chat:**

```         
Please add detailed inline comments to the following Python script to explain what each
section does. Assume I am an R user who is familiar with data analysis but new to Python.

[paste the Python script here]
```

```{python}
#| label: ex7-commented-python
#| eval: false

# Paste the commented Python code here for reference
# (We are not running this in R — it's Python, displayed for reference)

```

------------------------------------------------------------------------

## Exercise 8 — Refactoring code {#ex8}

::: {.callout-tip icon="false"}
## 🎯 Instructor cue — Slide: "Suggesting improvements to code (refactoring)"

Open a new chat. Paste the nested loop from Exercise 4 and use the
prompt below. After ChatGPT responds, copy the refactored code here and
verify it gives the same result as the original nested loop.
:::

**Prompt to paste into a NEW ChatGPT chat:**

```         
Here is some R code that works but could be improved. Can you suggest improvements
for readability and efficiency? Please explain the changes before showing the code.

[paste your nested loop code from Exercise 4 here]
```

```{r}
#| label: ex8-refactored

# Paste ChatGPT's refactored code here and run it
# Verify it matches the results_df from Exercise 4

```

::: {.callout-note collapse="true" icon="false"}
## ✅ What ChatGPT typically suggests

``` r
library(dplyr)

results_df <- penguins_filtered |>
  split(list(.$species, .$sex), drop = TRUE) |>
  lapply(function(df) {
    result <- calc_lm(df, "bill_length_mm", "bill_depth_mm")
    result$species <- unique(df$species)
    result$sex     <- unique(df$sex)
    result
  }) |>
  bind_rows()

results_df
```
:::

------------------------------------------------------------------------

## Exercise 9 — Unit tests {#ex9}

::: {.callout-tip icon="false"}
## 🎯 Instructor cue — Slide: "Writing unit tests"

Paste calc_lm() and the prompt into a new chat. Run the test. It will
likely fail due to a perfect-collinearity warning. Paste the error back
into ChatGPT and ask it to fix the test — this demonstrates the
iterative back-and-forth workflow.
:::

**Prompt to paste into a NEW ChatGPT chat:**

```         
Please write a unit test for the following R function using the testthat package.
The function takes a dataframe and column names for y and x, fits a linear regression,
and returns slope, intercept, and p-value as a dataframe.

[paste calc_lm() code here]
```

```{r}
#| label: ex9-unit-test

# Paste ChatGPT's unit test here and run it
# If it fails, paste the error/warning back into ChatGPT and ask for a fix

```

::: {.callout-note collapse="true" icon="false"}
## ✅ What a working unit test looks like

``` r
library(testthat)

# Slightly noisy data avoids perfect-collinearity warning
set.seed(123)
test_data <- data.frame(
  x = 1:5,
  y = c(2, 4, 6, 8, 10) + rnorm(5, mean = 0, sd = 0.1)
)

test_that("calc_lm returns correct slope, intercept, and p-value", {
  result <- calc_lm(test_data, "y", "x")
  expect_equal(result$slope,     2, tolerance = 0.1)
  expect_equal(result$intercept, 0, tolerance = 0.1)
  expect_lt(result$p_value, 0.05)
})
```
:::

------------------------------------------------------------------------

## Exercise 10 — Translating to Python {#ex10}

::: {.callout-tip icon="false"}
## 🎯 Instructor cue — Slide: "Translating code from one language to another"

Paste the nested loop into a new chat with the prompt below. Point out:
the translated code is not stand-alone (it calls calc_lm() and
penguins_filtered which don't exist in Python yet). Ask students: what
would you need to do to make it runnable? As a bonus, ask ChatGPT to
also refactor the Python version.
:::

**Prompt to paste into a NEW ChatGPT chat:**

```         
Please translate the following R code into Python using pandas where appropriate.
Note that calc_lm() and penguins_filtered need to be defined separately in Python —
just translate the loop structure for now.

[paste the nested loop from Exercise 4 here]
```

```{python}
#| label: ex10-python-translation
#| eval: false

# Paste the Python translation here for reference

```

**Bonus — refactor in Python:**

```         
Now refactor this Python version for better readability and efficiency.
```

```{python}
#| label: ex10-python-refactored
#| eval: false

# Paste the refactored Python code here

```

------------------------------------------------------------------------

# Wrap-Up Notes {#wrapup}

::: {.callout-important icon="false"}
## ⚠️ Things to remember about AI coding tools

**Hallucinations:** LLMs can confidently generate wrong code. Always
test it. If you couldn't write the code yourself given the time, be
extra careful trusting the output.

**Privacy:** Don't paste sensitive or private data into a public LLM.
Prompts may be stored and reviewed by the provider.

**Attribution:** AI-generated code may reproduce patterns from
open-source repositories without proper attribution. Check licenses for
your use case.

**You are responsible:** Both tools make you faster at writing code —
they don't replace understanding it. Always verify that the output is
scientifically correct, not just syntactically valid.
:::

::: {.callout-tip icon="false"}
## 💡 Quick decision guide: Copilot vs. ChatGPT

| Situation                               | Better tool |
|-----------------------------------------|-------------|
| Writing code line-by-line in real time  | **Copilot** |
| Well-defined task with clear pseudocode | **Either**  |
| Complex logic that needs revision       | **ChatGPT** |
| Commenting unfamiliar code              | **ChatGPT** |
| Refactoring a block of code             | **ChatGPT** |
| Debugging with an error message         | **ChatGPT** |
| Translating between languages           | **ChatGPT** |
| Writing unit tests                      | **ChatGPT** |
:::
