---
title: "Advanced Quarto Features - Live Demo"
author: "Dr. Gretchen LeBuhn & Dr. Nicole Adelstein"
date: "Week 3 Day 2"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
    embed-resources: true
  pdf:
    toc: true
    geometry: margin=1in
---

## Introduction

This document demonstrates professional Quarto features including:

- Labeled figures with cross-references
- Formatted tables with captions
- Mathematical equations
- Callout blocks
- Multiple layout options

We'll analyze bee species richness data from three study sites.

## Load Libraries and Data

```{r}
#| label: setup
#| message: false
#| warning: false

# Load tidyverse package (includes ggplot2, dplyr, readr, and more)
# This is a collection of packages for data manipulation and visualization
library(tidyverse)

# Load knitr package for creating formatted tables with kable() function
library(knitr)

# Read bee data from CSV file
# read.csv() imports data from a comma-separated values file
# Returns a data frame with columns matching the CSV headers
# File must be in the same directory as this .qmd file
bee_data <- read.csv("bee_data.csv")

# head() displays the first 6 rows of the data frame
# Useful for quickly checking that data imported correctly
head(bee_data)
```

## Professional Figures

### Species Richness by Site

Let's visualize species richness across our three study sites.

```{r}
#| label: fig-richness-site
#| fig-cap: "Species richness varies significantly among sites, with the southern site showing highest diversity."
#| fig-width: 6
#| fig-height: 4

# ggplot() initializes a plot object
# First argument is the data frame
# aes() defines the aesthetic mappings (which variables map to x, y, colors, etc.)
ggplot(bee_data, aes(x = site, y = richness, fill = site)) +
  
  # geom_boxplot() adds a box-and-whisker plot layer
  # alpha = 0.7 sets transparency (0 = fully transparent, 1 = fully opaque)
  geom_boxplot(alpha = 0.7) +
  
  # geom_jitter() adds individual data points with random horizontal scatter
  # width = 0.2 controls how much horizontal jitter to apply
  # size = 2 sets the point size (larger number = bigger points)
  geom_jitter(width = 0.2, alpha = 0.5, size = 2) +
  
  # labs() sets axis labels and titles
  # x = sets the x-axis label, y = sets the y-axis label
  labs(x = "Study Site", 
       y = "Species Richness") +
  
  # theme_minimal() applies a clean, minimalist theme
  # base_size = 12 sets the base font size for all text in the plot
  theme_minimal(base_size = 12) +
  
  # theme() allows fine-tuning of specific plot elements
  # legend.position = "none" removes the legend entirely
  theme(legend.position = "none") +
  
  # scale_fill_brewer() uses ColorBrewer color palettes
  # palette = "Set2" selects a specific color scheme designed to be colorblind-friendly
  scale_fill_brewer(palette = "Set2")
```

As shown in @fig-richness-site, the southern site consistently shows higher species richness compared to northern and central sites. This pattern may be related to temperature gradients across sites.

### Temperature Patterns

```{r}
#| label: fig-temperature
#| fig-cap: "Mean temperature increases from north to south across study sites."
#| fig-width: 6
#| fig-height: 4

# Create a boxplot showing temperature distribution by site
# aes() maps: x = categorical site variable, y = numeric temperature, fill = colors by site
ggplot(bee_data, aes(x = site, y = temperature, fill = site)) +
  
  # Add boxplot layer with 70% opacity
  geom_boxplot(alpha = 0.7) +
  
  # Set axis labels
  labs(x = "Study Site",
       y = "Temperature (°C)") +
  
  # Apply minimal theme with 12pt base font size
  theme_minimal(base_size = 12) +
  
  # Remove legend (redundant since x-axis already shows sites)
  theme(legend.position = "none") +
  
  # Use ColorBrewer "Set1" palette (different from previous plot for variety)
  scale_fill_brewer(palette = "Set1")
```

:::{.callout-note}
## Sampling Period
All measurements were taken during peak flowering season (June-August 2024) to standardize environmental conditions.
:::

### Multiple Plots Side-by-Side

```{r}
#| label: fig-comparison
#| fig-cap: "Comparison of richness and temperature patterns across sites."
#| fig-subcap: 
#|   - "Species richness distribution"
#|   - "Temperature distribution"
#| layout-ncol: 2
#| fig-width: 4
#| fig-height: 3

# Plot 1: Richness histogram
# aes(x = richness) maps only x variable; y is automatically the count
ggplot(bee_data, aes(x = richness)) +
  
  # geom_histogram() creates a histogram (bars showing frequency distribution)
  # binwidth = 1 makes each bar represent a 1-unit interval
  # fill = "steelblue" sets the bar color
  geom_histogram(binwidth = 1, fill = "steelblue", alpha = 0.7) +
  
  # Set axis labels
  labs(x = "Species Richness", y = "Count") +
  
  # Apply minimal theme
  theme_minimal()

# Plot 2: Temperature histogram
# Same structure as Plot 1, but with temperature data and coral color
ggplot(bee_data, aes(x = temperature)) +
  
  # Create histogram with 1-degree bins
  geom_histogram(binwidth = 1, fill = "coral", alpha = 0.7) +
  
  # Set axis labels
  labs(x = "Temperature (°C)", y = "Count") +
  
  # Apply minimal theme
  theme_minimal()
```

## Professional Tables

### Summary Statistics

Now let's create a professional summary table of our data.

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary statistics of species richness and temperature by study site."

# Calculate summary statistics using tidyverse pipe operator %>%
# The pipe passes the result of each line to the next line as the first argument

# Start with bee_data
summary_stats <- bee_data %>%
  
  # group_by() splits data into groups (one for each site)
  # All subsequent operations are performed within each group
  group_by(site) %>%
  
  # summarise() creates new columns by calculating summaries
  # Each calculation is applied to each group separately
  summarise(
    # mean() calculates the average
    # round(x, 1) rounds to 1 decimal place
    # Backticks allow column names with spaces
    `Mean Richness` = round(mean(richness), 1),
    
    # sd() calculates the standard deviation (measure of variability)
    `SD Richness` = round(sd(richness), 1),
    
    # Same calculations for temperature
    `Mean Temp (°C)` = round(mean(temperature), 1),
    `SD Temp` = round(sd(temperature), 1),
    
    # n() counts the number of observations in each group
    N = n()
  )

# Create formatted table
# kable() converts data frame to formatted table
kable(summary_stats, 
      # digits = 1 controls decimal places for all numeric columns
      digits = 1,
      # align = controls text alignment: 'l' = left, 'c' = center, 'r' = right
      # One letter per column: left-align site names, center all numbers
      align = c('l', 'c', 'c', 'c', 'c', 'c'))
```

Statistical analysis (@tbl-summary) reveals that the southern site has the highest mean species richness (`r round(mean(bee_data$richness[bee_data$site == "South"]), 1)` species) and warmest temperatures.

### ANOVA Results

```{r}
#| label: tbl-anova
#| tbl-cap: "ANOVA results testing differences in species richness among sites."

# Run ANOVA (Analysis of Variance)
# aov() performs ANOVA test
# Formula syntax: response_variable ~ predictor_variable
# Tests if mean richness differs significantly among sites
anova_results <- aov(richness ~ site, data = bee_data)

# summary() extracts the ANOVA table from the model object
# Returns a list containing F-statistic, p-value, degrees of freedom, etc.
anova_summary <- summary(anova_results)

# Extract values from the summary
# [[1]] accesses the first element of the summary list (the ANOVA table)
# as.data.frame() converts it to a data frame for easier manipulation
anova_df <- as.data.frame(anova_summary[[1]])

# Format for display
# Create a cleaner version with row names as a column
anova_table <- data.frame(
  # row.names() extracts the row labels (Source names)
  Source = row.names(anova_df),
  
  # Access columns from the ANOVA results using $ notation
  Df = anova_df$Df,
  
  # round() to 2 decimal places for readability
  `Sum Sq` = round(anova_df$`Sum Sq`, 2),
  `Mean Sq` = round(anova_df$`Mean Sq`, 2),
  `F value` = round(anova_df$`F value`, 2),
  
  # ifelse() is a conditional: if p < 0.001, show "< 0.001", else round to 3 decimals
  # This handles the very small p-values more clearly
  `P value` = ifelse(anova_df$`Pr(>F)` < 0.001, "< 0.001", 
                     round(anova_df$`Pr(>F)`, 3)),
  
  # check.names = FALSE preserves column names with spaces
  check.names = FALSE
)

# kable() creates formatted table
# row.names = FALSE prevents duplicate row numbers
kable(anova_table, row.names = FALSE)
```

Results from @tbl-anova indicate highly significant differences in species richness among the three study sites.

:::{.callout-warning}
## Sample Size Consideration
Sample sizes were limited by weather conditions during the sampling period. Sites with rain events during scheduled sampling days have fewer replicates.
:::

## Mathematical Equations

### Shannon Diversity Index

We calculated Shannon diversity using the formula:

$$
H' = -\sum_{i=1}^{S} p_i \ln(p_i)
$$ {#eq-shannon}

where $S$ is the total number of species and $p_i$ is the proportion of individuals belonging to species $i$.

The Shannon diversity index (@eq-shannon) provides a measure that accounts for both species richness and evenness in the community.

### Simpson's Index

An alternative diversity metric is Simpson's index:

$$
D = 1 - \sum_{i=1}^{S} p_i^2
$$ {#eq-simpson}

Both @eq-shannon and @eq-simpson are commonly used in community ecology, with Simpson's index being less sensitive to rare species.

## Relationship Between Temperature and Richness

Let's explore whether temperature predicts species richness.

```{r}
#| label: fig-temp-richness
#| fig-cap: "Positive relationship between temperature and species richness across all sites."
#| fig-width: 6
#| fig-height: 5

# Create scatter plot with regression line
# aes() maps temperature to x-axis, richness to y-axis, color points by site
ggplot(bee_data, aes(x = temperature, y = richness, color = site)) +
  
  # geom_point() adds scatter plot points
  # size = 3 makes points larger for better visibility
  # alpha = 0.7 adds slight transparency
  geom_point(size = 3, alpha = 0.7) +
  
  # geom_smooth() adds a fitted line to show trend
  # method = "lm" fits a linear model (straight line)
  # se = TRUE includes the confidence interval (shaded area around line)
  # color = "black" makes the line black (overrides color aesthetic)
  # linetype = "dashed" creates a dashed line instead of solid
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
  
  # Set all labels including legend title
  # color = "Study Site" labels the legend (since color aesthetic is mapped to site)
  labs(x = "Temperature (°C)",
       y = "Species Richness",
       color = "Study Site") +
  
  # Apply minimal theme
  theme_minimal(base_size = 12) +
  
  # Move legend to bottom to save horizontal space
  # legend.position can be "top", "bottom", "left", "right", or "none"
  theme(legend.position = "bottom") +
  
  # Use ColorBrewer "Dark2" palette for better contrast with black regression line
  scale_color_brewer(palette = "Dark2")
```

:::{.callout-tip}
## Analysis Suggestion
The positive trend in @fig-temp-richness suggests temperature may be an important driver of bee diversity. Consider including additional environmental variables (humidity, floral resources) in future analyses to better understand this pattern.
:::

## Column Layout Example

:::: {.columns}

::: {.column width="50%"}
### Key Findings

- Southern site shows highest richness
- Temperature correlates with diversity
- Significant site differences
- Pattern consistent across sampling dates
:::

::: {.column width="50%"}
### Next Steps

- Expand sampling to additional sites
- Include rarefaction analysis
- Measure floral resource availability
- Conduct seasonal comparison
:::

::::

## Conclusions

This analysis demonstrates several important patterns:

1. Species richness varies significantly among study sites (@fig-richness-site, @tbl-anova)
2. The southern site shows consistently higher diversity (@tbl-summary)
3. Temperature appears to be positively correlated with richness (@fig-temp-richness)
4. These patterns warrant further investigation with expanded sampling

:::{.callout-important}
## Key Takeaway
Geographic gradients in bee diversity may be driven by temperature, but additional environmental factors should be considered before drawing causal conclusions.
:::

## Session Information

```{r}
#| label: session-info

# sessionInfo() prints information about the R session
# Includes: R version, operating system, loaded packages and their versions
# This is crucial for reproducibility - allows others to recreate your exact environment
sessionInfo()
```

---

**Document rendered:** `r format(Sys.Date(), "%B %d, %Y")`
