---
title: "Data Decision Log"
subtitle: "Your Project Name Here"
author: "Your Name"
date: today
date-format: "MMMM D, YYYY"
format: 
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    number-sections: true
    embed-resources: true
editor: 
  markdown: 
    wrap: 72
---

# Overview

This document records all major analytical decisions made during this
project. Each entry documents what was decided, why, what alternatives
were considered, and what trade-offs were involved.

::: callout-tip
## How to Use This Template

1.  Copy this file to your R Project folder
2.  Rename it `decision_log.qmd`
3.  Update the YAML header above (title, subtitle, author)
4.  Delete this callout box and the example entries below
5.  Start adding your own decisions using the template in the Appendix
:::

**Project start date:** \[Insert date\]\
**Last updated:** `r Sys.Date()`

------------------------------------------------------------------------

# Data Cleaning Decisions

## Entry 1: Example - Handling Missing Values {#entry-001}

**Date:** YYYY-MM-DD\
**Status:** ✅ Final

### Decision

\[State clearly what you decided to do\]

Example: Removed all rows where the primary key variable `site_id` was
NA.

### Reasoning

\[Explain WHY you made this decision. Reference documentation, theory,
or practical constraints.\]

Example: Cannot link observations to site characteristics without site
identifiers. Checked with field team and confirmed these 3 observations
were data entry errors where the site column was accidentally left blank
during transcription.

### Code Implementation

```{r}
#| eval: false
#| code-fold: false

# Show the code that implements your decision
# Set eval: false so it doesn't actually run in this template

# Example:
# original_data <- read_csv("data/raw_data.csv")
# nrow(original_data)  # 615 observations
# 
# cleaned_data <- original_data %>%
#   filter(!is.na(site_id))
# 
# nrow(cleaned_data)  # 612 observations
```

### Alternatives Considered

1.  **Alternative 1: \[Describe alternative approach\]**\
    Reason for rejection: \[Why you didn't choose this\]

2.  **Alternative 2: \[Describe another approach\]**\
    Reason for rejection: \[Why you didn't choose this\]

Example alternatives: - Create "Unknown Site" category → Rejected
because observations are unusable for site-level analysis - Try to
reconstruct site from GPS coordinates → Rejected because GPS data also
missing

### Trade-offs

-   **Gain:** \[What you gained from this decision\]
-   **Lose:** \[What you gave up or lost\]
-   **Risk:** \[Any risks or uncertainties\]
-   **Acceptable because:** \[Why the trade-off is reasonable\]

Example: - **Gain:** Data integrity for all site-level analyses -
**Lose:** 3 observations (0.5% of total dataset) - **Risk:** Minimal -
checked that excluded observations are randomly distributed across time
periods - **Acceptable because:** Research questions focus on site-level
patterns

### References (if applicable)

-   Citation 1
-   Citation 2
-   Discussion with \[person\] on \[date\]

------------------------------------------------------------------------

## Entry 2: \[Your Second Decision Title\] {#entry-002}

**Date:** YYYY-MM-DD\
**Status:** 🔄 Draft

\[Copy the structure from Entry 1 above\]

------------------------------------------------------------------------

# Variable Creation Decisions

## Entry 3: \[Decision About Creating/Recoding Variables\] {#entry-003}

**Date:** YYYY-MM-DD\
**Status:** ✅ Final

### Decision

\[What new variable did you create or how did you recode existing
variables?\]

### Reasoning

\[Why was this necessary? What does this variable represent?\]

### Code Implementation

```{r}
#| eval: false

# Example:
# data <- data %>%
#   mutate(
#     new_variable = case_when(
#       condition1 ~ "Category A",
#       condition2 ~ "Category B",
#       TRUE ~ "Other"
#     )
#   )
```

### Alternatives Considered

\[What other ways could you have created or defined this variable?\]

### Trade-offs

\[What information is gained or lost by this
categorization/transformation?\]

------------------------------------------------------------------------

# Analytical Decisions

## Entry 4: \[Statistical Method Choice\] {#entry-004}

**Date:** YYYY-MM-DD\
**Status:** ✅ Final

### Decision

\[What statistical approach or model did you choose?\]

### Context

\[What problem are you trying to solve? What are the characteristics of
your data?\]

### Reasoning

\[Why is this approach appropriate? What assumptions are you making?\]

### Code Implementation

```{r}
#| eval: false

# Show the key analytical code
# Example:
# model <- lm(response ~ predictor1 + predictor2, data = data)
# summary(model)
```

### Alternatives Considered

\[What other statistical approaches did you consider?\]

### Trade-offs

\[What are the strengths and limitations of this approach?\]

### Assumptions

\[List key assumptions and how you checked them\]

### References

\[Cite relevant statistical references or methodology papers\]

------------------------------------------------------------------------

# Visualization Decisions

## Entry 5: \[Figure Design Choice\] {#entry-005}

**Date:** YYYY-MM-DD\
**Status:** ✅ Final

### Decision

\[What type of visualization did you choose and why?\]

### Reasoning

\[What does this visualization show that others don't? What makes it
effective for your audience?\]

### Implementation

```{r}
#| eval: false
#| label: fig-example
#| fig-cap: "Your figure caption here"

# Example visualization code
# library(ggplot2)
# ggplot(data, aes(x = variable1, y = variable2)) +
#   geom_point() +
#   labs(title = "Your title",
#        x = "X axis label",
#        y = "Y axis label") +
#   theme_minimal()
```

### Alternatives Considered

\[What other visualization types did you consider? Why didn't you use
them?\]

### Trade-offs

\[What does this visualization show well? What doesn't it show?\]

------------------------------------------------------------------------

# Status Updates & Revisions

## Entry 6: \[Revisiting a Previous Decision\] {#entry-006}

**Date:** YYYY-MM-DD\
**Status:** ⚠️ Supersedes Entry #\[X\]

::: callout-warning
## Revision Notice

This entry supersedes [Entry X](#entry-00X). The original decision has
been revised based on new information or analysis results.
:::

### What Changed

\[What decision are you revising?\]

### Why the Revision

\[What did you learn that prompted this change?\]

### New Decision

\[What's your updated approach?\]

### Implications

\[How does this affect previous analyses? What needs to be redone?\]

### Lessons Learned

\[What did you learn from this experience? This is important scientific
thinking!\]

------------------------------------------------------------------------

# Summary of Key Decisions

::: callout-note
## Quick Reference

Use this section to summarize the most important decisions that affect
interpretation of your results:

1.  [Decision 1](#entry-001): Brief description
2.  [Decision 2](#entry-002): Brief description
3.  [Decision 3](#entry-003): Brief description
4.  [Decision 4](#entry-004): Brief description
5.  [Decision 5](#entry-005): Brief description

Update this list as you add significant decisions throughout your
project.
:::

------------------------------------------------------------------------

# Outstanding Questions / Future Decisions

Use this section to track decisions you still need to make or questions
you need to resolve:

## To Decide

1.  **\[Topic needing decision\]**
    -   What's the question?
    -   What information do you need?
    -   When will you decide?
    -   Who can you consult?
2.  **\[Another pending decision\]**
    -   Context
    -   Next steps

## Questions for Advisor/Collaborators

1.  \[Question 1\]
2.  \[Question 2\]

------------------------------------------------------------------------

# Appendix: Entry Template

Use this template when adding new entries to your decision log. Copy and
paste it, fill in the sections, and delete this appendix from your final
document.

## Entry X: \[Descriptive Title Goes Here\] {#entry-00X}

**Date:** YYYY-MM-DD\
**Status:** 🔄 Draft / ✅ Final / ⚠️ Revised

### Decision

\[One clear sentence: What did you decide?\]

### Reasoning

\[Why did you make this decision? What evidence or logic supports it?\]

### Code Implementation (if applicable)

```{r}
#| eval: false

# Show relevant code here
# Use eval: false so template code doesn't run
```

### Alternatives Considered

1.  **\[Alternative 1\]**\
    Reason for rejection:

2.  **\[Alternative 2\]**\
    Reason for rejection:

### Trade-offs

-   **Gain:**
-   **Lose:**
-   **Risk:**
-   **Acceptable because:**

### References (if applicable)

-   

-   

------------------------------------------------------------------------

# Notes on Using This Template

::: callout-tip
## Tips for Success

1.  **Write as you go** - Don't wait until the end to document decisions
2.  **Be honest** - Uncertainty is okay and should be documented
3.  **Use the test** - Could someone understand your reasoning without
    asking questions?
4.  **Update status badges** - 🔄 Draft → ✅ Final as decisions solidify
5.  **Link between entries** - Use cross-references when decisions
    relate
6.  **Render regularly** - Click "Render" to see your HTML output
7.  **Commit to Git** - Track evolution of your thinking over time
:::

## Status Badge Guide

-   🔄 **Draft** - Still thinking about this decision
-   ✅ **Final** - Decision is made and implemented\
-   ⚠️ **Revised** - Decision changed after seeing results/new
    information
-   📌 **Critical** - Particularly important decision that affects key
    results

## When to Write an Entry

Write a decision log entry when:

-   [x] You're choosing between two reasonable options
-   [x] You're combining or recoding categories
-   [x] You're excluding data (ALWAYS document exclusions)
-   [x] You're making an assumption that could be challenged
-   [x] Something didn't work and you want to remember why
-   [x] Someone might ask "why did you do it that way?"

Don't document:

-   ✗ Obvious or trivial decisions
-   ✗ Standard procedures followed exactly as documented
-   ✗ Exploratory poking around (unless you learn something important)

## The Quality Test

**Could a colleague/classmate understand your reasoning without asking
follow-up questions?**

-   If YES → You wrote enough
-   If NO → Add more explanation

------------------------------------------------------------------------

*This decision log is version controlled with Git. Use commit messages
like "Add decision about outlier handling" to track the evolution of
your analytical thinking.*

*Remember: The thinking is the work. The code is just execution.*
